### 2022-11-15 ms.Rmd
Init empty project and MS

---
title: TITLE
shorttitle: SHORT TITLE
leftheader: SHORT TITLE
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Oxford Internet Institute, University of Oxford
    email: matti.vuorre@oii.ox.ac.uk
    role:
      - role mcrole
  - name: Author McAuthor
    affiliation: 2
    address: Address McAddress
    email: email@mcemail.com
    role:
      - role mcrole

affiliation:
  - id: 1
    institution: Oxford Internet Institute, University of Oxford
  - id: 2
    institution: Institution McInstitution

authornote: |
  \noindent \textbf{This manuscript is not yet peer-reviewed.}

abstract: |
  Many things, few words.

keywords: words, key
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: "references.bib"
floatsintext: yes
linenumbers: yes
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: false
  papaja::apa6_docx: 
    number_sections: false
---

```{r setup, include = FALSE}
library(papaja)

# Parallel computation options
MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
```

R [@rcoreteamLanguageEnvironmentStatistical2022]

\clearpage

# References

::: {#refs custom-style="Bibliography"}
:::

<!-- Create an appendix with new page, table, and figure numbers. -->

\clearpage
\onecolumn

# Appendix A: Appendix McAppendix

<!-- Prepend table & figure numbers with S and reset counters -->
\setcounter{page}{1}
\setcounter{table}{0}
\setcounter{figure}{0}
\renewcommand{\thetable}{A\arabic{table}}
\renewcommand{\thefigure}{A\arabic{figure}}

```{r include-only-for-pdf}
#| include: !expr knitr::is_latex_output()
```

### 2022-11-21 Generation_Code.r
Annotated R-file for Latent-Mean centered data
Code generates latent-mean centered data for a 2 variable VAR model

@@ -1,0 +1,140 @@
+# Population Values ------------------------------------------------------------
+# Example values for VAR with two variables
+REPS <- 200
+nr_vars <- 2
+nr_indiv <- 100 
+nr_time <- 70 
+
+tau <- c(1,1) # Innovation Variance
+
+corr <- 0.3 # Corr between Y's for fixed inno
+alpha_location <- c(4,4.5)
+alpha_scale <- c(.5,.5)
+corr_alpha <- 0.3
+
+beta_location <- matrix(c(.4,.2,.1,.3), K)
+beta_scale <- matrix(c(.1,.1,.1,.1), K) 
+corr_beta <- 0.1
+
+
+
+# Code to generate VAR data ----------------------------------------------------
+
+# sd's of Y residuals
+tau <- tau_location
+
+# correlation between residuals
+omega <- diag(nr_vars)
+omega[1, 2] <- omega[2, 1] <- corr
+
+# turn tau and omega into cov_matrix
+sigma_y <- diag(tau) %*% omega %*% diag(tau)
+
+# Create individual means and allow them to covary (between person-networks)
+alpha_hat_location <- alpha_location # Mean of means
+alpha_hat_scale <- alpha_scale # SD of means
+
+omega_alpha <- diag(nr_vars) # correlation between means
+omega_alpha[lower.tri(omega_alpha)] <- omega_alpha[
+    upper.tri(omega_alpha)] <- corr_alpha
+
+sigma_alpha <- diag(alpha_hat_scale) %*% omega_alpha %*% diag(alpha_hat_scale)
+
+# Draw individual means
+alphas <- rmvnorm(nr_indiv, alpha_hat_location, sigma_alpha)
+
+# Create individual (cross)lagged effects
+# Storage vector
+betas <- array(NA, c(nr_indiv, nr_vars, nr_vars))
+
+# Means of (cross)lagged effects      
+beta_hat_location <- beta_location
+# SDs of (cross)lagged effects 
+beta_hat_scale <- beta_scale
+
+# correlation between AR's and cross-lagged
+omega_beta <- diag(nr_vars * nr_vars)
+omega_beta[lower.tri(omega_beta)] <- omega_beta[
+    upper.tri(omega_beta)] <- corr_beta
+
+sigma_beta <- diag(c(beta_hat_scale)) %*% omega_beta %*% diag(c(beta_hat_scale))
+
+# Generate initial individual (cross)lagged effects
[…]

### 2022-11-23 ms.Rmd
Acknowledgements

@@
 ---
-title: TITLE
-shorttitle: SHORT TITLE
-leftheader: SHORT TITLE
+title: Why, when and how to latent mean center variables in multilevel models
+shorttitle: Latent mean centering
+leftheader: Latent mean centering
 author: 
@@
 
+# Acknowledgements
+
+We would like to thank Mauricio Garnier-Villarre, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
+Special thanks to Mauricio Garnier-Villarre for being the first to post the syntax that the code here was based on.
+We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.
+
 # References

### 2022-11-23 Generation_Code.r
Make script a function

@@ -1,140 +1,158 @@
-# Population Values ------------------------------------------------------------
-# Example values for VAR with two variables
-REPS <- 200
-nr_vars <- 2
-nr_indiv <- 100 
-nr_time <- 70 
 
-tau <- c(1,1) # Innovation Variance
-
-corr <- 0.3 # Corr between Y's for fixed inno
-alpha_location <- c(4,4.5)
-alpha_scale <- c(.5,.5)
-corr_alpha <- 0.3
-
-beta_location <- matrix(c(.4,.2,.1,.3), K)
-beta_scale <- matrix(c(.1,.1,.1,.1), K) 
-corr_beta <- 0.1
-
-
-
-# Code to generate VAR data ----------------------------------------------------
-
-# sd's of Y residuals
-tau <- tau_location
-
-# correlation between residuals
-omega <- diag(nr_vars)
-omega[1, 2] <- omega[2, 1] <- corr
-
-# turn tau and omega into cov_matrix
-sigma_y <- diag(tau) %*% omega %*% diag(tau)
-
-# Create individual means and allow them to covary (between person-networks)
-alpha_hat_location <- alpha_location # Mean of means
-alpha_hat_scale <- alpha_scale # SD of means
-
-omega_alpha <- diag(nr_vars) # correlation between means
-omega_alpha[lower.tri(omega_alpha)] <- omega_alpha[
+#' Simulate data from a multilevel model
+#'
+#' @param nr_vars 
+#' @param nr_indiv 
+#' @param nr_time 
+#' @param tau 
+#' @param corr 
+#' @param alpha_location 
+#' @param alpha_scale 
+#' @param corr_alpha 
+#' @param beta_location 
+#' @param beta_scale 
+#' @param corr_beta 
+#'
+#' @return
+#' @export
+#'
+#' @examples
+simulate_data <- function(
+    # Example values for VAR with two variables
+    nr_vars = 2,
+    nr_indiv = 100, 
+    nr_time = 70,
+    
+    tau = c(1,1), # Innovation Variance
+    
+    corr = 0.3, # Corr between Y's for fixed inno
+    alpha_location = c(4,4.5),
+    alpha_scale = c(.5,.5),
+    corr_alpha = 0.3,
+    
+    beta_location = matrix(c(.4,.2,.1,.3), K),
+    beta_scale = matrix(c(.1,.1,.1,.1), K), 
+    corr_beta = 0.1
+) {
+  
+  # sd's of Y residuals
+  tau <- tau_location
+  
[…]

### 2022-11-23 Generation_Code.r
Document simulation function parameters

@@ -1,31 +1,30 @@
 
 #' Simulate data from a multilevel model
 #'
-#' @param nr_vars 
-#' @param nr_indiv 
-#' @param nr_time 
-#' @param tau 
-#' @param corr 
-#' @param alpha_location 
-#' @param alpha_scale 
-#' @param corr_alpha 
-#' @param beta_location 
-#' @param beta_scale 
-#' @param corr_beta 
+#' @param nr_vars Number of outcome variables
+#' @param nr_indiv Number of individuals
+#' @param nr_time Number of timepoints
+#' @param tau Innovation variance (vector of length `nr_vars`)
+#' @param corr Correlation between outcomes 
+#' @param alpha_location Outcome intercept means
+#' @param alpha_scale Outcome intercept variances (standard deviations?)
+#' @param corr_alpha Correlation between intercepts
+#' @param beta_location Predictor means
+#' @param beta_scale Predictor variances (standard deviations?)
+#' @param corr_beta Correlation between predictors
 #'
 #' @return
 #' @export
 #'
 #' @examples
 simulate_data <- function(
-    # Example values for VAR with two variables
     nr_vars = 2,
     nr_indiv = 100, 
     nr_time = 70,
     
-    tau = c(1,1), # Innovation Variance
+    tau = c(1,1),
     
-    corr = 0.3, # Corr between Y's for fixed inno
+    corr = 0.3, 
     alpha_location = c(4,4.5),
     alpha_scale = c(.5,.5),
     corr_alpha = 0.3,

### 2022-11-23 ms.Rmd
Manuscript structure?
Name typo

@@
 
+# Introduction
+
+## Centering
+
+## Latent mean centering
+
+# Example analysis
+
+# Simulation study
+
+# Discussion
+
 R [@rcoreteamLanguageEnvironmentStatistical2022]
@@
 Special thanks to Mauricio Garnier-Villarre for being the first to post the syntax that the code here was based on.
+We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.
+We would like to thank Mauricio Garnier-Villarreal, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
+Special thanks to Mauricio Garnier-Villarreal for being the first to post the syntax that the code here was based on.
 We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.

### 2022-12-07 ms.Rmd

@@
 ---
-title: Why, when and how to latent mean center variables in multilevel models
+title: 'Bias no more! A practical guide to latent mean centering in multilevel models'
 shorttitle: Latent mean centering
@@
 abstract: |
-  Many things, few words.
+  Here's a plan.
+
+  - Longitudinal and repeated measures are becoming more common because SCIENCE
+  - Centering is crucial, but...
+  - Latent mean centering! Like a whole section of words!
+  - tbd
 
@@
 library(papaja)
+library(posterior)
+library(ggdist)
+library(brms)
+library(tidyverse)
 
-# Parallel computation options
+# HMC and parallel computation options
 MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
 if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
+THREADS <- 1
+CORES <- MAX_CORES
+ITER <- 10000
+if (require(cmdstanr)) {
+  options(brms.backend = "cmdstanr")
+  THREADS <- MAX_CORES %/% 4
+}
+
+# Save models in a directory
+dir.create("models", FALSE)
 ```
@@
 # Example analysis
+
+<!-- Just a toy example for now to write some code -->
+
+```{r}
+# Data from McNeish
+d <- read_csv("data/Two-Level Data.csv", col_names = FALSE)[,c(5:6,1:2)]
+names(d) <- c("person","time","urge", "dep")
+d <- d %>% 
+  # Create lagged variable
+  group_by(person) %>% 
+  mutate(urge1 = lag(urge)) %>% 
+  ungroup()
+
+# Model definition from https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424/22
+# Estimate depBI (latent means) using nlf() inside bf()
+model <- bf(
+  urge ~ alpha + phi*(urge1 - alpha) + beta*(dep - depB),
+  alpha ~ 1 + (1 | person),
+  phi ~ 1 + (1 | person),
+  beta ~ 1 + (1 | person),
+  nlf(depB ~ depBI),
+  depBI ~ 1 + (1 | person),
+  nl = TRUE
+) +
+  gaussian()
+
+p <- get_prior(model, data = d) %>%
+  mutate(
+    prior = case_when(
+      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
+      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
+      TRUE ~ prior
+    )
+  )
+
+fit1 <- brm(
+  model,
+  data = d,
+  prior = p,
+  cores = MAX_CORES,
+  threads = THREADS,
+  iter = ITER,
+  control = list(adapt_delta = 0.99),
+  file = "models/m_mh91"
+)
+
+# Code example from Ethan McCormick
+# Matti's code edited to add `depb` as 'main effect'
+# Estimate model without nlf() with depb added to the first formula
+# This seems to be slightly different as it adds depb as predictor of urge?
+# But under typical conditions these should give similar key estimates (?)
+model <- bf(
+  urge ~ alpha + phi*(urge1 - alpha) + depb + beta*(dep - depb),
+  alpha ~ 1 + (1 | person),  # Latent intercept of urge
+  depb ~ 1 + (1 | person),  # Latent mean of depression
+  phi ~ 1 + (1 | person),  # Slope of latent mean-person centered lagged urge on urge
+  beta ~ 1 + (1 | person),  # Slope of latent mean-person centered depression on urge
+  nl = TRUE,
+  center = FALSE. # This seems key
+) +
+  gaussian()
+
+#Set priors
+p <- get_prior(model, data = d) %>% 
+  mutate(
+    prior = case_when(
+      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
+      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
+      TRUE ~ prior
+    )
+  )
+
+fit2 <- brm(
+  model,
+  data = d,
+  prior = p, 
+  cores = MAX_CORES,
+  threads = THREADS,
+  iter = ITER,
+  control = list(adapt_delta = 0.99),
+  file = "models/m_mh5"
+)
+
+bind_rows(
+  as_draws_df(fit1, variable = c("b_", "sd_", "sigma"), regex = TRUE),
+  as_draws_df(fit2, variable = c("b_", "sd_", "sigma"), regex = TRUE) %>% 
+    rename_with(~str_replace(., "depb", "depBI")), 
+  .id = "Model"
+) %>% 
+  mutate(
+    Model = factor(Model, labels = c("nlf()", "bf()")),
+    ) %>% 
+  select(-c(.chain, .iteration, .draw)) %>% 
+  pivot_longer(-Model) %>% 
+  ggplot(aes(value, name, col = Model, fill = Model)) +
+  stat_halfeye(
+    data = . %>% filter(Model == "nlf()"),
+    side = "top", normalize = "xy", height = .5
+  ) +
+  stat_halfeye(
+    data = . %>% filter(Model == "bf()"),
+    side = "bottom", normalize = "xy", height = .5
+  )
+```
 
@@
 
-We would like to thank Mauricio Garnier-Villarre, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
-Special thanks to Mauricio Garnier-Villarre for being the first to post the syntax that the code here was based on.
-We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.
 We would like to thank Mauricio Garnier-Villarreal, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).

### 2022-12-07 ms.Rmd
Figure options

@@
 library(posterior)
+library(scales)
 library(ggdist)
@@
 dir.create("models", FALSE)
+
+# Plotting options
+theme_set(
+  theme_linedraw(base_size = 9.6) +
+    theme(
+      strip.background = element_rect(fill = "gray90", color = "gray90"),
+      strip.text = element_text(color = "black", hjust = 0),
+      panel.grid = element_blank()
+    )
+)
 ```

### 2022-12-07 ms.Rmd
Fix typo

@@
   nl = TRUE,
-  center = FALSE. # This seems key
+  center = FALSE # This seems key
 ) +

### 2022-12-07 ms.Rmd
A suggested figure.

@@
 
+```{r}
+#| label: fig-me-in-means
+#| eval: false
+#| include: false
+#| fig.cap: Illustration of the risks associated with using observed means for group-mean centering, and how those are avoided by using latent means to group-mean center instead.
+
+# A suggestion for an illustrative figure, showing risks (bias) of observed mean centering, but also how latent mean centering avoids them.
+
+tibble(
+  n = seq(1, 100, by = .1),
+  s = 1/sqrt(n)
+  ) %>% 
+  ggplot(aes(n, s)) +
+  scale_y_continuous(
+    "Measurement error in observed mean",
+    limits = 0:1, 
+    breaks = extended_breaks(),
+    expand = expansion(c(.02, .05))
+  ) +
+  scale_x_continuous(
+    "Number of measurements",
+    expand = expansion(.025, .025)
+  ) +
+  # Theoretical measurement error in mean
+  geom_line()
+  # More lines or panels
+  # Observed dynamic panel bias from a simulation
+  # geom_line(something something)
+  # Observed bias in other covariate from a simulation
+  # geom_line(also something something)
+  # Latent mean centering results--lines that are not biased
+  # geom_line(...)
+```
+
 # Discussion

### 2022-12-07 ms.Rmd
Revise abstract

@@
 ---
-title: 'Bias no more! A practical guide to latent mean centering in multilevel models'
-shorttitle: Latent mean centering
-leftheader: Latent mean centering
+title: 'Bias no more: A practical guide to latent group-mean centering in longitudinal and other hierarchical bayesian models'
+shorttitle: Latent group-mean centering
+leftheader: Latent group-mean centering
 author: 
@@
 authornote: |
-  \noindent \textbf{This manuscript is not yet peer-reviewed.}
+  \noindent \textit{This manuscript is not yet peer-reviewed.}
 
 abstract: |
-  Here's a plan.
+  Researchers studying the contributions of observed variables on other variables in longitudinal and other repeated measures data sets have long recognized the importance of group-mean centering variables in order to isolate within- and between-person causal processes.
+However, a critically important issue in centering---treating the person means as unknown parameters rather than observed quantities---has received less attention among applied practitioners.
+One reason for this has been the general unavailability and perceived complexity of latent group-mean centering versus observed mean centering.
+  In this practical tutorial, we introduce code for latent group-mean centering in longitudinal and hierarchical models using the free and open source software R and specifically the brms R package.
+We show via analyses of data on well-being and technology use that latent mean centering is not only important, easy, but also fun.
+We hope that these examples, and the general FOSS availability of the discussed methods encourage more researchers to adopt latent mean centering instead of oserved mean centering to avoid bias in parameter estimates.
+We also discuss relations between longitudinal models / SEM / multilevel models / bayes.
+Bayes!
 
-  - Longitudinal and repeated measures are becoming more common because SCIENCE
-  - Centering is crucial, but...
-  - Latent mean centering! Like a whole section of words!
-  - tbd
-
-keywords: words, key
+keywords: longitudinal data, repeated measures, centering, well-being, technology effects
 wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"

### 2022-12-07 ms.Rmd
Set document options

@@
 library(posterior)
+library(knitr)
 library(scales)
@@
 library(tidyverse)
+
+# Document options
+opts_chunk$set(
+  cache = TRUE,
+  echo = FALSE
+)
 

### 2022-12-07 ms.Rmd
Acknowledgements -> authors

@@
 
-We would like to thank Mauricio Garnier-Villarreal, Simon Brauer, Ethan McCormick, and everyone else who contributed to the discussion and development of the correct brms syntax on the Stan discussion board (https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
-Special thanks to Mauricio Garnier-Villarreal for being the first to post the syntax that the code here was based on.
-We believe developing methods and theories collaboratively, and in the open, to be a fruitful path forward in better understanding human behavior and how to best model it.
-
 # References

### 2022-12-07 ms.Rmd
Safer parallelization

@@
 MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
-if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
+if (is.na(MAX_CORES)) MAX_CORES <- parallelly::availableCores(logical = FALSE)
 THREADS <- 1
-CORES <- MAX_CORES
 ITER <- 10000
+CHAINS <- 4
 if (require(cmdstanr)) {
   options(brms.backend = "cmdstanr")
-  THREADS <- MAX_CORES %/% 4
+  THREADS <- MAX_CORES %/% CHAINS
 }

### 2022-12-07 ms.Rmd
Draft a motivational paragraph.

@@
 theme_set(
-  theme_linedraw(base_size = 9.6) +
+  theme_linedraw(base_size = 10) +
     theme(
@@
 
+However, previously latent mean centering has only been available through commercial software packages (MPlus), and was suggested to only apply to latent variables models (M & H).
+We show a general approach to latent group mean centering using the freely available and flexible R package Bayesian Regression Modelling using Stan (brms; Bürkner), and that latent mean centering applies to *all* analyses of repeated measures (or, generally, clustered) data.
+
 ## Centering
@@
 # Example analysis
+
+To illustrate latent group-mean centering in practice, we introduce an example dataset of 100 Dutch adolescents, from a study by [@cite].
 

### 2023-02-14 ms.Rmd
Updates

--- 2022-12-07 ms.Rmd
@@
 ---
-title: 'Bias no more: A practical guide to latent group-mean centering in longitudinal and other hierarchical bayesian models'
+title: 'Bias no more: Introduction to latent group-mean centering in hierarchical bayesian models'
 shorttitle: Latent group-mean centering
@@
     corresponding: yes
-    address: Oxford Internet Institute, University of Oxford
-    email: matti.vuorre@oii.ox.ac.uk
-    role:
-      - role mcrole
-  - name: Author McAuthor
+    address: Tilburg University
+    email: m.j.vuorre@tilburguniversity.edu
+  - name: Author 2
     affiliation: 2
-    address: Address McAddress
-    email: email@mcemail.com
-    role:
-      - role mcrole
+    address: 
+    email:
 
@@
   - id: 1
-    institution: Oxford Internet Institute, University of Oxford
+    institution: Tilburg University
   - id: 2
-    institution: Institution McInstitution
+    institution: Institution 2
 
@@
 abstract: |
-  Researchers studying the contributions of observed variables on other variables in longitudinal and other repeated measures data sets have long recognized the importance of group-mean centering variables in order to isolate within- and between-person causal processes.
-However, a critically important issue in centering---treating the person means as unknown parameters rather than observed quantities---has received less attention among applied practitioners.
-One reason for this has been the general unavailability and perceived complexity of latent group-mean centering versus observed mean centering.
-  In this practical tutorial, we introduce code for latent group-mean centering in longitudinal and hierarchical models using the free and open source software R and specifically the brms R package.
-We show via analyses of data on well-being and technology use that latent mean centering is not only important, easy, but also fun.
-We hope that these examples, and the general FOSS availability of the discussed methods encourage more researchers to adopt latent mean centering instead of oserved mean centering to avoid bias in parameter estimates.
-We also discuss relations between longitudinal models / SEM / multilevel models / bayes.
-Bayes!
+  Researchers studying links between observed variables in longitudinal and other repeated measures data sets have long recognized the importance of person-mean centering in order to isolate within- and between-person associations.
+However, a critical issue in centering---treating the person means as unknown parameters rather than observed values---has received less attention among applied practitioners.
+One reason for this has been the general unavailability and perceived complexity of latent person-mean centering in contrast to observed mean centering.
+Here, we introduce latent person-mean centering and accessible syntax for implementing it in a wide variety of flexible bayesian models with the R package brms.
+We demonstrate the syntax, and benefits of latent centering, through an example analysis of social media effects on adolescent well-being.
+In appendices, we further contrast the performance of latent mean centering to observed mean centering via simulations, and show how it extends to more than two levels of analysis.
+Latent mean centering provides unbiased estimates of within- and between-person associations, and we hope that the availability and flexibility of our proposed approach motivates researchers to apply this method widely in longitudinal and other multilevel modelling scenarios.
 
@@
 floatsintext: yes
-linenumbers: yes
+linenumbers: no
 draft: no
@@
 footnotelist: no
-csl: "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
 documentclass: apa7
@@
 library(papaja)
+library(citr)
 library(posterior)
@@
 MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
-if (is.na(MAX_CORES)) MAX_CORES <- parallel::detectCores(logical = FALSE)
+if (is.na(MAX_CORES)) MAX_CORES <- parallelly::availableCores(logical = FALSE)
 THREADS <- 1
-CORES <- MAX_CORES
-ITER <- 10000
+ITER <- 5000
+CHAINS <- 4
 if (require(cmdstanr)) {
   options(brms.backend = "cmdstanr")
-  THREADS <- MAX_CORES %/% 4
+  THREADS <- MAX_CORES %/% CHAINS
 }
@@
 
+```{r data-download}
+# Download data if not downloaded
+path <- "data/beyens-et-al-2021.csv"
+dir.create(dirname(path), FALSE)
+if (!file.exists(path)) 
+  download.file("https://uvaauas.figshare.com/ndownloader/files/28057809", path)
+
+# Load and clean
+d <- read_csv(path) %>%
+  # Subset data to a handful of participants' evening observations
+  filter(
+    BeepTE1 == 6,
+    ID %in% sample(.$ID, 30)
+  ) %>% 
+  mutate(
+    person = factor(ID),
+    time = BeepNE1,
+    # Sum of all passive use in the past hour
+    sm = rowSums(select(., INP01E1, INP02E1, WAP01E1, SNP01E1, SNP02E1), na.rm = TRUE),
+    # It can exceed 60 if p reported multiple overlapping uses
+    sm = if_else(sm > 60, 60, sm),
+    happy = AWBE1,
+    .keep = "none"
+  ) %>% 
+  # Drop missing data
+  drop_na(sm, happy)
+
+# Lagged and centered variables
+d <- d %>% 
+  # Lag happiness by person
+  mutate(
+    happy1 = lag(happy), 
+    .by = person
+  ) %>% 
+  # Grand mean center so between-person component is centered on zero
+  mutate(
+    smg = sm - mean(sm, na.rm = TRUE)
+  ) %>% 
+  # Calculate person means and person-specific deviations therein
+  mutate(
+    smb = mean(smg, na.rm = TRUE), 
+    smw = smg - smb,
+    happy1w = happy1 - mean(happy1, na.rm = TRUE),
+    .by = person
+  )
+```
+
 # Introduction
 
-## Centering
+In longitudinal and other repeated measures data where predictors are observed associations can emerge at two (or more) levels: Between individuals and within individuals [@bolgerIntensiveLongitudinalMethods2013].
+Because these two levels of analysis can embody different theoretical constructs, researchers have widely recognized the importance of centering the data appropriately, such that within-person associations can be estimated independently from between-person associations [@endersCenteringPredictorVariables2007].
+If this distinction between sources of variance is not made, the resulting estimates are likely to be confusing mixtures of associations that occur between individuals and within individuals [@cronbachResearchClassroomsSchools1976 p.18; @raudenbushHierarchicalLinearModels2002 p.138].
+Specifically, when a predictor can vary at two levels---observations varying around a person's average, and persons' averages varying around the global average---it is common to manipulate the data such that the analysis distinguishes those two levels.
 
-## Latent mean centering
+In typical applications, a predictor is separated into two components: First, each individual's mean is calculated.
+Including the person mean as predictor in a regression model will describe the extent to which individuals with different average levels of the predictor differ on the outcome variable.
+Second, the individuals' means are subtracted from each observation, leading to a variable that indicates deviations around each person's average.
+This latter variable represents pure within-person variability.
+Including it as a model predictor leads to estimating how the outcome differs at different levels of the predictor, for the average person.
+Typically, it is this latter within-person association that is of key theoretical interest in psychology: It can be used to describe a counterfactual causal scenario where increasing the predictor for an individual leads to changes in that person's outcome [@rohrerTheseAreNot2021].
 
-# Example analysis
+While simple to implement, this data manipulation trick has one potentially severe shortcoming: It can lead to biases in the estimated regression coefficients, especially when the number of observations per person is small.
+These biases result from ignoring the uncertainty inherent in the person means: They can only be estimated based on a sample of limited data, but simple calculations cannot carry this uncertainty forward into the resulting estimates.
+While this issue is recognized in the methodological literature [e.g.
+@asparouhovLatentVariableCentering2019; @mcneishPrimerTwolevelDynamic2020], practical implementations to overcome it have so far been limited to commercial solutions, most prominently the SEM software MPlus.
+In this manuscript, we provide a practical tutorial on latent mean centering in the context of longitudinal and repeated measures data in psychology.
+We implement latent mean centering in a series of multilevel models using the brms R package.
+Our aim is to provide this illustrative code so that latent mean centering can be more widely adopted within the psychological sciences.
 
-<!-- Just a toy example for now to write some code -->
+In appendix A(?), we also conduct a simulation study to illustrate the conditions under which latent mean centering is most useful: As the number of observations per participant decreases, the beneficial debiasing effect of latent mean centering becomes more noticeable and important.
+In addition, while previous work has focused on latent mean centering specifically in the context of longitudinal datasets and timeseries analyses, we point out that it is a more general issue and applicable to datasets with potentially multiple levels of analysis?
+In appendix B, we extend the syntax to a three level model where latent means are estimated for adolescents nested within classes?
+
+# Latent mean centering tutorial
+
+In this example analysis, we examine a longitudinal dataset of Dutch adolescents' social media use and well-being.
+The authors of the original research article have made their valuable data openly available, which we appreciate.
+The sample consists of 387 Dutch adolescents (mean age: 14, 54% girls), who answered 34,930 experience sampling questions about their well-being ("How happy do you feel right now?"; 1: "Not at all", 4: "A little", 7: "Completely") and social media use.
+For this analysis, we used the sum of passive Instagram, WhatsApp, and Snapchat reported in the past hour.
+We show a snippet of this data in Table \@ref(tab:tab-data).
 
 ```{r}
-# Data from McNeish
-d <- read_csv("data/Two-Level Data.csv", col_names = FALSE)[,c(5:6,1:2)]
-names(d) <- c("person","time","urge", "dep")
-d <- d %>% 
-  # Create lagged variable
-  group_by(person) %>% 
-  mutate(urge1 = lag(urge)) %>% 
-  ungroup()
+#| label: tab-data
+d %>% 
+  slice(1:3, .by = person) %>% 
+  head() %>% 
+  apa_table(
+    caption = "Six rows of example data", 
+    note = "sm = passive social media use; smg = grand-mean centered sm; smb = between-person sm; smw = within-person sm; happyw1 = within-person lagged happiness rating.",
+    col_spanners = list(`Original data` = c(1, 5), `Centered values` = c(6, 9)),
+    digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1), 
+    font_size = "small", 
+    span_text_columns = FALSE
+  )
+```
 
-# Model definition from https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424/22
-# Estimate depBI (latent means) using nlf() inside bf()
+Table \@ref(tab:tab-data) shows the three first observations of two randomly selected participants from the sample.
+The first four columns contain an anomyous person identifier, the observation time point (an increasing integer), the number of minutes of passive social media use in the past hour, and the happiness rating.
+Then, for illustration, we have created three additional columns containing the centered social media use variables.
+The first, "smg", is the simple grand-mean centered "sm" variable.
+This variable is not required in all analyses.
+However, it can be used to zero-center the between-person variable if it is of interest.
+"smb" is the between-person component, which is simply that person's sample mean.
+"smw" is the deviation of the social media use value on that row from the person's mean, and thus the within-person component.
+
+We loosely follow the authors' original analysis, which focused on the associations between social media use and happiness.
+To ground our tutorial on common practices, we first ignore latent mean centering, and describe the analysis in terms of the observed values.
+In this analysis, we are interested in the within- and between-person associations between passive social media use in the past hour and current well-being.
+To do so, we define a multilevel model of happiness ratings on both "smw" and "smb", and allow for random variability in person-specific averages in happiness ratings, and the association between happiness ratings and "smw".
+
+**Throw in a line here about why the autoregressive term is included.**
+
+In typical R formula syntax, generalized for multilevel models in the lme4 package, this model is
+
+which translates to the mixed effects model in Box 1.
+This model returns three key population-level estimates---parameter values assumed to hold for the average person in the population from which the sample was drawn: An intercept, slope of "smb", and slope of "smw".
+The first describes the happiness level of an average individual who passively used an average amount of social media in the past hour, and
+
+```{r}
+#| eval: false
 model <- bf(
-  urge ~ alpha + phi*(urge1 - alpha) + beta*(dep - depB),
-  alpha ~ 1 + (1 | person),
-  phi ~ 1 + (1 | person),
-  beta ~ 1 + (1 | person),
-  nlf(depB ~ depBI),
-  depBI ~ 1 + (1 | person),
-  nl = TRUE
+  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
+  b0happy ~ 1 + (1 | person),
+  b1happy ~ 1 + (1 | person),
+  b2happy ~ 1 + (1 | person),
+  b0sm ~ 1 + (1 | person),
+  nl = TRUE,
+  center = FALSE
 ) +
@@
 
-p <- get_prior(model, data = d) %>%
-  mutate(
-    prior = case_when(
-      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
-      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
-      TRUE ~ prior
-    )
-  )
+# Set priors
+p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)
 
-fit1 <- brm(
+fit <- brm(
   model,
@@
   control = list(adapt_delta = 0.99),
-  file = "models/m_mh91"
+  file = "models/brm-fit-1"
 )
-
-# Code example from Ethan McCormick
-# Matti's code edited to add `depb` as 'main effect'
-# Estimate model without nlf() with depb added to the first formula
-# This seems to be slightly different as it adds depb as predictor of urge?
-# But under typical conditions these should give similar key estimates (?)
-model <- bf(
-  urge ~ alpha + phi*(urge1 - alpha) + depb + beta*(dep - depb),
-  alpha ~ 1 + (1 | person),  # Latent intercept of urge
-  depb ~ 1 + (1 | person),  # Latent mean of depression
-  phi ~ 1 + (1 | person),  # Slope of latent mean-person centered lagged urge on urge
-  beta ~ 1 + (1 | person),  # Slope of latent mean-person centered depression on urge
-  nl = TRUE,
-  center = FALSE # This seems key
-) +
-  gaussian()
-
-#Set priors
-p <- get_prior(model, data = d) %>% 
-  mutate(
-    prior = case_when(
-      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
-      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
-      TRUE ~ prior
-    )
-  )
-
-fit2 <- brm(
-  model,
-  data = d,
-  prior = p, 
-  cores = MAX_CORES,
-  threads = THREADS,
-  iter = ITER,
-  control = list(adapt_delta = 0.99),
-  file = "models/m_mh5"
-)
-
-bind_rows(
-  as_draws_df(fit1, variable = c("b_", "sd_", "sigma"), regex = TRUE),
-  as_draws_df(fit2, variable = c("b_", "sd_", "sigma"), regex = TRUE) %>% 
-    rename_with(~str_replace(., "depb", "depBI")), 
-  .id = "Model"
-) %>% 
-  mutate(
-    Model = factor(Model, labels = c("nlf()", "bf()")),
-    ) %>% 
-  select(-c(.chain, .iteration, .draw)) %>% 
-  pivot_longer(-Model) %>% 
-  ggplot(aes(value, name, col = Model, fill = Model)) +
-  stat_halfeye(
-    data = . %>% filter(Model == "nlf()"),
-    side = "top", normalize = "xy", height = .5
-  ) +
-  stat_halfeye(
-    data = . %>% filter(Model == "bf()"),
-    side = "bottom", normalize = "xy", height = .5
-  )
+summary(fit, prior = TRUE)
 ```
@@
 
-\clearpage
+# Author Contributions
+
+Conceptualization: MV
+Methodology: MV
+Software: MV
+Validation: MV
+Formal Analysis: MV
+Data Curation: MV
+Writing -- Original Draft: MV
+Writing -- Review & Editing: MV
+Visualization: MV
+Supervision: MV
+Project Administration: MV
+
+# Conflicts of Interest
+
+The author(s) declare that there were no conflicts of interest with respect to the authorship or the publication of this article.
 
 # Acknowledgements
+
+We thank everyone who contributed to this discussion on the Stan discussion board (https://web.archive.org/web/20230207111825/https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
 
@@
 
-# Appendix A: Appendix McAppendix
+# Appendix Z: Simulation study
 
@@
 
-```{r include-only-for-pdf}
-#| include: !expr knitr::is_latex_output()
+Hereafter be monsters and dragons
+
+```{r data-create}
+# Simple simulation scheme
+# happiness predicted from within-person social media use, and happiness(t-1).
+library(MASS)
+library(faux)
+library(lme4)
+
+# Numbers of people and observations
+n_person <- 50
+n_time <- 5
+# Population level parameters
+beta <- c(
+  4, # Happiness intercept
+  .2, # happy <- social media within
+  .25, # happy <- happy(t-1)
+  30 # Social media intercept
+)
+names(beta) <- c("b0happy", "b1happy", "b2happy", "b0sm")
+# Standard deviations across people (in same order)
+# Note b3happy is "fixed" (0)
+tau <- c(1, .2, .075, 5)
+# Random effect standard deviations (same for all pairs)
+rho <- .3
+# Standard deviations of observations
+sigma <- c(sigmahappy = 0.5, sigmasm = 5)
+
+# Generate person-level parameters
+x <- rnorm_multi(
+  n = n_person,
+  mu = beta,
+  sd = tau,
+  r = rho,
+  empirical = TRUE
+) %>% 
+  mutate(
+    sigmahappy = sigma[1],
+    sigmasm = sigma[2]
+  )
+
+# Repeate rows and create outcomes
+x <- x %>% 
+  rownames_to_column("person") %>% 
+  crossing(time = 1:n_time) %>% 
+  mutate(
+    sm = rnorm(nrow(.), b0sm, sigmasm),
+    happy = rnorm(nrow(.), b0happy + b1happy*(sm - b0sm), sigmahappy)
+  ) %>% 
+  mutate(
+    happy1 = lag(happy),
+    happy = happy + happy1*b2happy,
+    .by = person
+  )
+
+x <- x %>% 
+  mutate(
+    smb = mean(sm),
+    smw = sm - smb,
+    .by = person
+  )
+y <- lmer(
+  happy ~ smw + happy1 + (smw + happy1 | person),
+  data = x
+)
 ```
+
+```{r}
+model <- bf(
+  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
+  b0happy ~ 1 + (1 | person),
+  b1happy ~ 1 + (1 | person),
+  b2happy ~ 1 + (1 | person),
+  b0sm ~ 1 + (1 | person),
+  nl = TRUE,
+  center = FALSE
+) +
+  gaussian()
+
+# Set priors
+p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)
+
+fit <- brm(
+  model,
+  data = d,
+  prior = p,
+  cores = MAX_CORES,
+  threads = THREADS,
+  iter = ITER,
+  control = list(adapt_delta = 0.99),
+  file = "models/brm-fit-1"
+)
+summary(fit, prior = TRUE)
+```

### 2023-03-17 Generation_Code.r
Wrap simulation code into a function

@@ -1,140 +1,191 @@
-# Population Values ------------------------------------------------------------
-# Example values for VAR with two variables
-REPS <- 200
-nr_vars <- 2
-nr_indiv <- 100 
-nr_time <- 70 
 
-tau <- c(1,1) # Innovation Variance
+library(tidyverse)
+library(mvtnorm)
 
-corr <- 0.3 # Corr between Y's for fixed inno
-alpha_location <- c(4,4.5)
-alpha_scale <- c(.5,.5)
-corr_alpha <- 0.3
-
-beta_location <- matrix(c(.4,.2,.1,.3), K)
-beta_scale <- matrix(c(.1,.1,.1,.1), K) 
-corr_beta <- 0.1
-
-
-
-# Code to generate VAR data ----------------------------------------------------
-
-# sd's of Y residuals
-tau <- tau_location
-
-# correlation between residuals
-omega <- diag(nr_vars)
-omega[1, 2] <- omega[2, 1] <- corr
-
-# turn tau and omega into cov_matrix
-sigma_y <- diag(tau) %*% omega %*% diag(tau)
-
-# Create individual means and allow them to covary (between person-networks)
-alpha_hat_location <- alpha_location # Mean of means
-alpha_hat_scale <- alpha_scale # SD of means
-
-omega_alpha <- diag(nr_vars) # correlation between means
-omega_alpha[lower.tri(omega_alpha)] <- omega_alpha[
-    upper.tri(omega_alpha)] <- corr_alpha
-
-sigma_alpha <- diag(alpha_hat_scale) %*% omega_alpha %*% diag(alpha_hat_scale)
-
-# Draw individual means
-alphas <- rmvnorm(nr_indiv, alpha_hat_location, sigma_alpha)
-
-# Create individual (cross)lagged effects
-# Storage vector
-betas <- array(NA, c(nr_indiv, nr_vars, nr_vars))
-
-# Means of (cross)lagged effects      
-beta_hat_location <- beta_location
-# SDs of (cross)lagged effects 
-beta_hat_scale <- beta_scale
-
-# correlation between AR's and cross-lagged
-omega_beta <- diag(nr_vars * nr_vars)
-omega_beta[lower.tri(omega_beta)] <- omega_beta[
-    upper.tri(omega_beta)] <- corr_beta
-
-sigma_beta <- diag(c(beta_hat_scale)) %*% omega_beta %*% diag(c(beta_hat_scale))
-
[…]

### 2023-03-17 ms.Rmd
Run simulation

@@
 library(citr)
+library(parameters)
 library(posterior)
+library(lme4)
 library(knitr)
@@
 if (is.na(MAX_CORES)) MAX_CORES <- parallelly::availableCores(logical = FALSE)
-THREADS <- 1
-ITER <- 5000
-CHAINS <- 4
 if (require(cmdstanr)) {
-  options(brms.backend = "cmdstanr")
-  THREADS <- MAX_CORES %/% CHAINS
+  options(
+    brms.backend = "cmdstanr", 
+    brms.threads = MAX_CORES %/% 4
+  )
 }
@@
 
-# Simulation study
-
-```{r}
-#| label: fig-me-in-means
-#| eval: false
-#| include: false
-#| fig.cap: Illustration of the risks associated with using observed means for group-mean centering, and how those are avoided by using latent means to group-mean center instead.
-
-# A suggestion for an illustrative figure, showing risks (bias) of observed mean centering, but also how latent mean centering avoids them.
-
-tibble(
-  n = seq(1, 100, by = .1),
-  s = 1/sqrt(n)
-  ) %>% 
-  ggplot(aes(n, s)) +
-  scale_y_continuous(
-    "Measurement error in observed mean",
-    limits = 0:1, 
-    breaks = extended_breaks(),
-    expand = expansion(c(.02, .05))
-  ) +
-  scale_x_continuous(
-    "Number of measurements",
-    expand = expansion(.025, .025)
-  ) +
-  # Theoretical measurement error in mean
-  geom_line()
-  # More lines or panels
-  # Observed dynamic panel bias from a simulation
-  # geom_line(something something)
-  # Observed bias in other covariate from a simulation
-  # geom_line(also something something)
-  # Latent mean centering results--lines that are not biased
-  # geom_line(...)
-```
-
 # Discussion
@@
 
-Hereafter be monsters and dragons
+Hereafter be monsters and dragons.
 
-```{r data-create}
-# Simple simulation scheme
-# happiness predicted from within-person social media use, and happiness(t-1).
-library(MASS)
-library(faux)
-library(lme4)
+Here, we conduct a small simulation study to verify that the proposed estimation method recovers specified parameter values.
 
-# Numbers of people and observations
-n_person <- 50
-n_time <- 5
-# Population level parameters
-beta <- c(
-  4, # Happiness intercept
-  .2, # happy <- social media within
-  .25, # happy <- happy(t-1)
-  30 # Social media intercept
-)
-names(beta) <- c("b0happy", "b1happy", "b2happy", "b0sm")
-# Standard deviations across people (in same order)
-# Note b3happy is "fixed" (0)
-tau <- c(1, .2, .075, 5)
-# Random effect standard deviations (same for all pairs)
-rho <- .3
-# Standard deviations of observations
-sigma <- c(sigmahappy = 0.5, sigmasm = 5)
+We simulate data from a bivariate model of $x$ and $y$ measured over $i$ timepoints from 1 to $I$ and for $j$ subjects in 1 to $J$.
+In this simulation, we set $J = 30$ and vary $I$ from 6 to 60.
+Both variables are gaussian, with $x$ a simple random variable with a person-varying mean $\alpha_x$, and $y$ is affected by its previous value and the previous value of $x$.
 
-# Generate person-level parameters
-x <- rnorm_multi(
-  n = n_person,
-  mu = beta,
-  sd = tau,
-  r = rho,
-  empirical = TRUE
-) %>% 
-  mutate(
-    sigmahappy = sigma[1],
-    sigmasm = sigma[2]
+\begin{align*}
+x_{ij} &\sim \operatorname{Normal}(\alpha_{xj}, \sigma^2_x) \\
+y_{ij} &\sim \operatorname{Normal}(
+  \alpha_{yj} +
+  \beta_{1j}(y_{[i-1]j} - \alpha_{yj}) +
+  \beta_{2j}(x_{[i-1]j} - \alpha_{xj}), 
+  \sigma^2_y
+) \\ \\
+\begin{bmatrix} 
+  \alpha_x \\ \alpha_y \\ \beta_1 \\ \beta_2
+\end{bmatrix} &\sim \operatorname{MVN}(
+\begin{bmatrix} 
+  \bar{\alpha}_x \\ \bar{\alpha}_y \\ \bar{\beta}_1 \\ \bar{\beta}_2
+\end{bmatrix}, \mathbf{\Sigma}) \\ \\
+\Sigma &= \mathbf{SRS} \\ \\
+\mathbf{S} &= \begin{bmatrix} 
+  \tau_0 &0 &0 &0 \\ 
+  0 &\tau_1 &0 &0\\ 
+  0 &0 &\tau_2 &0 \\ 
+  0 &0 &0 &\tau_3
+\end{bmatrix} \\ \\
+\mathbf{R} &= 
+\begin{bmatrix} 
+  1 &\rho_{\alpha_x\alpha_y} &\rho_{\alpha_x\beta_1} &\rho_{\alpha_x\beta_2} \\ 
+  \rho_{\alpha_x\alpha_y} &1 &\rho_{\alpha_y\beta_1} &\rho_{\alpha_y\beta_2} \\ 
+  \rho_{\alpha_x\beta_1} &\rho_{\alpha_y\beta_1} &1 &\rho_{\beta_1\beta_2} \\ 
+  \rho_{\alpha_x\beta_2} &\rho_{\alpha_y\beta_2} &\rho_{\beta_1\beta_2} &1 \\
+\end{bmatrix}.
+\end{align*}
+
+```{r}
+# Run simulations
+# Save simulation results to file for every nr_individuals * nr_time
+dir.create("sims", FALSE)
+
+# Load function to simulate data
+source("Generation_Code.r")
+
+# Specify population level parameters
+pop <- tibble(
+  sigma_x = 1.0,
+  sigma_y = 1.0,
+  cor_xy = 0.0,
+  alpha_x = 0.0,
+  alpha_y = 0.0,
+  tau_alpha_x = 0.5,
+  tau_alpha_y = 0.5,
+  cor_alpha = 0.2,
+  beta_yy1 = 0.3,
+  beta_yx = 0.25,
+  beta_xy = 0.0,
+  beta_xx1 = 0.0,
+  tau_beta_yy1 = 0.1,
+  tau_beta_yx = 0.2,
+  tau_beta_xy = 0.0,
+  tau_beta_xx1 = 0.0,
+  cor_beta = 0.0
 )
 
-# Repeate rows and create outcomes
-x <- x %>% 
-  rownames_to_column("person") %>% 
-  crossing(time = 1:n_time) %>% 
-  mutate(
-    sm = rnorm(nrow(.), b0sm, sigmasm),
-    happy = rnorm(nrow(.), b0happy + b1happy*(sm - b0sm), sigmahappy)
-  ) %>% 
-  mutate(
-    happy1 = lag(happy),
-    happy = happy + happy1*b2happy,
-    .by = person
+# Save empty brms model
+model <- bf(
+  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
+  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
+  nl = TRUE
+) +
+  gaussian()
+p <- prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphax) +
+  prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphay) +
+  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayy) +
+  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayx) +
+  prior(lkj(2), class = cor, group = individual) +
+  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphax) +
+  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphay) +
+  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayy) +
+  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayx)
+fit_brms_empty <- brm(
+  model,
+  data = simulate_twolevel_data()$data[[1]],
+  prior = p,
+  control = list(adapt_delta = 0.95),
+  chains = 0,
+  file = "models/brm-fit-empty"
 )
 
-x <- x %>% 
+# Function to estimate parameters
+simulate_estimate <- function(nr_indiv = 30, nr_time = 10, iter = 2) {
+  path <- str_glue("sims/nr_indiv-{nr_indiv}__nr_time-{nr_time}.rds")
+  if (!file.exists(path)) {
+
+    library(furrr)
+    plan(multisession, workers = 8)
+
+    sims <- pop %>% 
+      # Cross design with sample sizes and simulation iterations
+      crossing(
+        nr_indiv = nr_indiv,
+        nr_time = nr_time,
+        i = 1:iter
+      ) %>% 
+      # Generate sets of person level parameters and data
+      mutate(out = pmap(., simulate_twolevel_data)) %>% 
+      unnest(out) %>% 
+      nest(pop_pars = -c(i, nr_indiv, nr_time, indiv_pars, data))
+
+    sims <- sims %>% 
       mutate(
-    smb = mean(sm),
-    smw = sm - smb,
-    .by = person
+        lmer = future_map(
+          data,
+          ~lmer(y ~ ylag_c + xlag_c + (ylag_c + xlag_c | individual), data = .x) %>% 
+            parameters(effects = "all", group_level = TRUE) %>% 
+            tibble() %>% 
+            suppressMessages() %>% 
+            suppressWarnings()
         )
-y <- lmer(
-  happy ~ smw + happy1 + (smw + happy1 | person),
-  data = x
       )
+
+    sims <- sims %>% 
+      mutate(
+        brms = map(
+          data,
+          ~update(
+            fit_brms_empty, 
+            newdata = .x,
+            cores = MAX_CORES,
+            recompile = FALSE,
+            threads = 2,
+            iter = 2000
+          ) %>% 
+            parameters(centrality = "mean", effects = "all", group_level = TRUE, test = NULL) %>% 
+            tibble()
+        )
+      )
+
+    saveRDS(sims, path)
+  } else {sims <- readRDS(path)}
+  sims
+}
+
+# Run for specified sample sizes
+sims <- tibble(people = 30, timepoints = c(6, 12, 18, 24, 30, 60)) %>% 
+  mutate(
+    out = map2(people, timepoints, ~simulate_estimate(.x, .y, iter = 100))
+    ) %>% 
+  unnest(out)
 ```
@@
 ```{r}
-model <- bf(
-  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
-  b0happy ~ 1 + (1 | person),
-  b1happy ~ 1 + (1 | person),
-  b2happy ~ 1 + (1 | person),
-  b0sm ~ 1 + (1 | person),
-  nl = TRUE,
-  center = FALSE
+#| label: fig-sim-res
+#| fig-cap: Simulation study results.
+Each panel displays the mean posterior mean calculated over 100 iterations of model fitting.
+Vertical bars indicate standard errors over those 100 simulation runs.
+Each panel is labelled with a parameter name corresponding to the above equations.
+#| fig-width: 7
+
+tmp <- sims %>% 
+  mutate(
+    result = pmap(
+      list(lmer, brms, data),
+      ~bind_rows(
+        # Wrangle lmer results
+        ..1 %>% 
+          filter(Effects == "fixed") %>% 
+          mutate(
+            Parameter = fct_recode(
+              Parameter, 
+              `alpha[y]` = "(Intercept)",
+              `beta[1]` = "ylag_c",
+              `beta[2]` = "xlag_c"
+            ),
+            Centering = "Observed (lmer)"
+          ),
+        # Wrangle brms results
+        ..2 %>% 
+          filter(str_starts(Parameter, "b_|sd_|cor_|sigma")) %>% 
+          mutate(
+            Parameter = fct_recode(
+              Parameter, 
+              `alpha[x]` = "b_alphax_Intercept",
+              `alpha[y]` = "b_alphay_Intercept",
+              `beta[1]` = "b_betayy_Intercept",
+              `beta[2]` = "b_betayx_Intercept",
+              `tau[alpha[x]]` = "sd_individual__alphax_Intercept",
+              `tau[alpha[y]]` = "sd_individual__alphay_Intercept",
+              `tau[beta[1]]` = "sd_individual__betayy_Intercept",
+              `tau[beta[2]]` = "sd_individual__betayx_Intercept",
+              `rho[alpha[x]~alpha[y]]` = "cor_individual__alphax_Intercept__alphay_Intercept",
+              `sigma[y]` = "sigma"
+            )
+          ) %>% 
+          filter(!str_starts(Parameter, "cor_i")) %>% 
+          mutate(Centering = "Latent (brms)") %>% 
+          rename(Coefficient = Mean),
+      ) %>% 
+        bind_rows(
+          # Observed mean of X
+          ..3 %>% 
+            summarise(Parameter = "alpha[x]", Coefficient = mean(x_b), Centering = "Observed (lmer)")
+        )
+    )
+  ) %>% 
+  select(1:4, result) %>% 
+  unnest(result)
+tmp %>%   
+  ggplot(aes(nr_time, Coefficient, col = Centering, fill = Centering)) +
+  scale_x_continuous(
+    "Number of timepoints per person",
+    breaks = extended_breaks()
   ) +
-  gaussian()
-
-# Set priors
-p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)
-
-fit <- brm(
-  model,
-  data = d,
-  prior = p,
-  cores = MAX_CORES,
-  threads = THREADS,
-  iter = ITER,
-  control = list(adapt_delta = 0.99),
-  file = "models/brm-fit-1"
+  scale_y_continuous(
+    "Posterior mean",
+    breaks = extended_breaks(),
+    expand = expansion(.5)
+  ) +
+  geom_hline(
+    data = pop %>% 
+      select(
+        `alpha[x]` = alpha_x, 
+        `alpha[y]` = alpha_y, 
+        `beta[1]` = beta_yy1, 
+        `beta[2]` = beta_yx, 
+        `tau[alpha[x]]` = tau_alpha_x, 
+        `tau[alpha[y]]` = tau_alpha_y,
+        `tau[beta[1]]` = tau_beta_yy1, 
+        `tau[beta[2]]` = tau_beta_yx,
+        `rho[alpha[x]~alpha[y]]` = cor_alpha, 
+        `sigma[y]` = sigma_y
+      ) %>% 
+      pivot_longer(everything(), names_to = "Parameter", values_to = "Coefficient"),
+    aes(yintercept = Coefficient),
+    linewidth = .15
+  ) +
+  stat_summary(
+    geom = "line", position = position_dodge(width = 2),
+    fun = "mean", linewidth = .33
+  ) +
+  stat_summary(
+    geom = "pointrange", position = position_dodge(width = 2),
+    fun.data = "mean_se", fatten = 1.5, linewidth = .25
+  ) +
+  facet_wrap(
+    "Parameter",
+    nrow = 2,
+    scales = "free_y",
+    labeller = label_parsed
+  ) +
+  theme(
+    legend.position = "bottom",
+    strip.text = element_text(size = rel(1.1))
     )
-summary(fit, prior = TRUE)
 ```

### 2023-03-17 ms.Rmd
Some notation fixes

@@
   \sigma^2_y
-) \\ \\
+)
+\end{align*}
+
+The second model equation above specifies that the $\beta$s model effects of the within-person deviated variables of lagged y and x, by virtue of subtracting from each their respective person-specific (latent) mean.
+I hope.
+Then we specify the rest of the multilevel model as usual, butchering the notation with bars to indicate average or population-level regression coefficients.
+We also specify matrices of standard deviations S and correlations R to pop into the covariance matrix (as those will be estimated, rather than covariances, by brms/Stan).
+
+\begin{align*}
 \begin{bmatrix} 
@@
 \mathbf{S} &= \begin{bmatrix} 
-  \tau_0 &0 &0 &0 \\ 
-  0 &\tau_1 &0 &0\\ 
-  0 &0 &\tau_2 &0 \\ 
-  0 &0 &0 &\tau_3
+  \tau_{\alpha_x} &0 &0 &0 \\ 
+  0 &\tau_{\alpha_y} &0 &0\\ 
+  0 &0 &\tau_{\beta_1} &0 \\ 
+  0 &0 &0 &\tau_{\beta_2}
 \end{bmatrix} \\ \\
@@
 \end{align*}
+
+We then estimate the resulting datasets with two models.
+The first is an lmer model that substitutes the observed means $\bar{x}_{.j}$ and $\bar{y}_{.j}$ for the latent intercepts $\alpha_x$ and $\alpha_y$, respectively.
+The second is our main brms model, specified as follows:
+
+```{r}
+#| include: true
+#| echo: true
+model <- bf(
+  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
+  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
+  nl = TRUE
+) +
+  gaussian()
+```
+
+Note that this model doesn't estimate an outcome model for $x$, which may or may not have consequences.
+We then went ahead and ran the simulations with 100 repetitions at each row of the simulation design matrix.
+This only took a few hours on a M1 laptop.
+We show results in Figure \@ref(fig:fig-sim-res).
+First, we see that observed mean centering recovers the specified values of $\alpha_x$ well--those are not in the model output but just means of $\bar{x}_{.j}$.
+The latent means estimates here are more variable across simulation runs, but also recover the mean well when the number of timepoints is larger than about 10.
+$\alpha_y$ remains biased upward even with a large number of timepoints per person, but not with the latent means model.
+
+The key coefficients $\beta_1$ and $_2$ show that latent mean centering effectively debiases the autocorrelated parameter, even at low numbers of timepoints.
+Estimates of $\beta_2$ are noisier across simulation runs but both models seem to recover the values well (observed mean model is consistently greater?)
+
+Other notable findings:
+- The correlation between intercepts (between-person correlation) $\rho$ is massively downward biased?
+- The between-person standard deviations are all over the place.
 
@@
 # Save empty brms model
-model <- bf(
-  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
-  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
-  nl = TRUE
-) +
-  gaussian()
 p <- prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphax) +

### 2023-03-29 Generation_Code.r
Simulation supplementary

@@ -182,10 +182,12 @@ simulate_twolevel_data <- function(
       .by = individual
     )
   
+  # Also save person-level parameters in a tibble
   coefs <- cbind(alphas, betas[,,1], betas[,,2]) %>% 
     as_tibble() %>% 
     set_names(c("alpha_y", "alpha_x", "beta_yy1", "beta_xy", "beta_yx", "beta_xx1"))
   
+  # Return both person-parameters and generated data in a tibble with list-columns
   tibble(indiv_pars = list(coefs), data = list(data_tot))
   
 }

### 2023-05-21 mlm-lmc.Rmd
Create mlm-lmc.Rmd

---
title: "latent-mean-center-mlm"
author: "Ethan M. McCormick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| warning: false
#| message: false
#| include: false

packages <- c("dplyr", "lmerTest", "brms")
invisible(lapply(packages, library, character.only = TRUE))
if (require(cmdstanr)) {
  options(brms.backend = "cmdstanr")
}
```

```{r}
btDat <- read.csv("~/Dropbox/Professional/projects/longitudinal-primer/codebook/data/feedback-learning.csv") %>%
  mutate(wave = wave - 2,
         age = (age - mean(age, na.rm = TRUE)) / sd(age, na.rm = TRUE),
         age2 = age^2) %>%
  group_by(id) %>%
  mutate(modul_id_mcent = modularity - mean(modularity, na.rm = TRUE),
         modul_id_mean = mean(modularity, na.rm = TRUE))
```

```{r}
lmer.fit <- lmer(learning.rate ~ age + age2 + 
                   modul_id_mcent + modul_id_mean + (1 | id),
                 data = btDat)

summary(lmer.fit)
```

```{r}
model <- bf(
  learning.rate ~ 1 + age + age2 + modul_id_mcent + modul_id_mean + (1 | id),
  #age + age2 + beta + beta2 ~ 1,
  nl = F
) +
  gaussian()

p <- get_prior(model, data = btDat) %>%
  mutate(
    prior = case_when(
      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
      TRUE ~ prior
    )
  )

job::job({fit1 <- brm(
  model,
  data = btDat,
  prior = p,
  cores = round(parallel::detectCores(logical = FALSE) %/% 2),
  iter = 10000,
  control = list(adapt_delta = 0.99),
  file = "models/braintime-fixed"
)})
summary(fit1)
```

```{r}
model <- bf(
  learning.rate ~ alpha + b1*age + b2*age2 + b3*(modularity - modC),
  alpha ~ 1 + (1 |id),
  b1 + b2 + b3 ~ 1,
  nlf(modC ~ modCi),
  modCi ~ 1 + (1 | id),
  nl = TRUE
) +
  gaussian()

p <- get_prior(model, data = btDat) %>%
  mutate(
    prior = case_when(
      class == "b" & coef == "Intercept" ~ "normal(0, 1.5)",
      class == "sd" & coef == "Intercept" ~ "student_t(7, 0, 1)",
      TRUE ~ prior
    )
  )

job::job({fit2 <- brm(
  model,
  data = btDat,
  prior = p,
  cores = round(parallel::detectCores(logical = FALSE) %/% 2),
  iter = 10000,
  control = list(adapt_delta = 0.99),
  file = "models/braintime-lmc"
)})
summary(fit2)
```

### 2023-10-31 simulation.qmd
Init new simple simulation scheme

---
title: "Simulation"
execute: 
  cache: true
format: 
  html:
    toc: true
    code-fold: true
---

```{r}
#| label: options

library(posterior)
library(mvtnorm)
library(ggdist)
library(brms)
library(tidyverse)

N_CORES <- min(parallel::detectCores(), 8)
options(
  mc.cores = N_CORES,
  brms.backend = if(require(cmdstanr)) "cmdstanr",
  brms.threads = N_CORES %/% 4
)
```

In this document we specify, simulate, and validate a univariate latent means brms model. Code based on Joran's blog post at <https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/>.

# Data

We first set up the data simulation. I used [this code](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/#testing) but put it inside a function.

```{r}
#' Simulate ar(1) data
#' 
#' Adapted from <https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/#testing>
#'
#' @param I number of individuals
#' @param t number of time points per individual
#' @param sigma residual standard deviation of observations
#' @param alpha_hat intercept mean
#' @param alpha_scale intercept standard deviation
#' @param beta_hat ar(1) slope mean
#' @param beta_scale ar(1) slope standard deviation
#'
#' @return a list of variables to send to Stan
#'
#' @examples simulate()
simulate <- function(
    I = 30, 
    t = 10, 
    sigma = 1, 
    alpha_hat = 4, 
    alpha_scale = 1, 
    beta_hat = 0.4, 
    beta_scale = 0.1
) {

  # Define data
  N <- I*t # total sample size
  individual <- rep(1:I, each = t)
  time <- rep(1:t, I)
  alphas <- rnorm(I, alpha_hat, alpha_scale) # person-specific intercepts
  betas <- rnorm(I, beta_hat, beta_scale) # person-specific slopes
  for (i in 1:I) {
    # The while loop avoids non-stationary AR processes
    # See Hamilton  pg. 259
    while(betas[i] <= -1 | betas[i] >= 1) {
      betas[i] <- rnorm(1, beta_hat, beta_scale)
    }
  }

  # Determine first observations for everyone. The variance for this first 
  # observation is different than for the subsequent ones and so it needs to be 
  # samples separately
  IndT1 <- match(unique(individual), individual)

  # Determine variance at first measurement for everyone 
  # (depends on their AR-parameter)
  sigmaT1 <- rep(NA, I)

  for (k in 1:I) {
    sigmaT1[k] <- sigma/(1-((betas[k])^2))
  }

  # First create storage matrices for non-centered and centered y-scores.
  # We need centered values, because of we use person-centered values as predictors, 
  # alpha will be equal to individual means instead of individual intercepts 
  # which are less informative.
  Y <- rep(NA, N)
  Yc <- rep(NA, N)

  # Draw first observation for each individual first
  for (l in 1:I) {
    Y[IndT1[l]] <- rnorm(1, alphas[l], sigmaT1[l])
    Yc[IndT1[l]] <-  Y[IndT1[l]] - alphas[l]
  }

  # Draw subsequent observations
  for (m in 1:N) {
    # This if statement makes sure I don't try to predict a persons first 
    # observation which is impossible there is no measurement before the first 
    # observation and so no predictor values for that observation
    if (time[m]>1) {
      Y[m] <- rnorm(1, (alphas[individual[m]] + betas[individual[m]]*Yc[m-1]), sigma)
      Yc[m] <-  Y[m] - alphas[individual[m]]
    }
  }

  return(
    list(
      I = I,
      T = t,
      N = N,
      sigma = sigma, 
      alpha_hat = alpha_hat, 
      alpha_scale = alpha_scale, 
      beta_hat = beta_hat, 
      beta_scale = beta_scale,
      individual = individual,
      time = time,
      y = Y
    )
  )

}
```

The function returns a list that we can pass to Stan. It also contains the data generation parameters.

```{r}
set.seed(31121)
dat <- simulate()
glimpse(simulate())
```

### 2023-10-31 simulation.qmd
Canonicalize Stan code
I don't think that's a word.

@@
 library(mvtnorm)
+library(cmdstanr)
 library(ggdist)
@@
 ```
+
+Let us then compile and sample from the cmdstanr model
+
+```{r}
+fit <- cmdstanr::cmdstan_model("model-ar1.stan", compile = FALSE)
+fit$format(
+  canonicalize = list("deprecations"), 
+  overwrite_file = TRUE, 
+  backup = FALSE
+)
+```

### 2023-10-31 simulation.qmd
Sample from cmdstanr model

@@
 simulate <- function(
-    I = 30, 
-    t = 10, 
+    I = 100, 
+    t = 50, 
     sigma = 1, 
@@
 ```{r}
-fit <- cmdstanr::cmdstan_model("model-ar1.stan", compile = FALSE)
-fit$format(
+model <- cmdstanr::cmdstan_model("model-ar1.stan", compile = FALSE)
+model$format(
   canonicalize = list("deprecations"), 
@@
 )
+model$compile()
+fit <- model$sample(
+  data = dat, 
+  adapt_delta = 0.99
+)
 ```
+
+```{r}
+as_draws_df(fit)
+```

### 2023-10-31 simulation.qmd
Wrangle data for brms

--- 2023-10-31 simulation.qmd
@@
 simulate <- function(
-    I = 30, 
-    t = 10, 
+    I = 50, 
+    t = 25, 
     sigma = 1, 
@@
 
+## Stan model
+
 Let us then compile and sample from the cmdstanr model
@@
 ```{r}
-fit <- cmdstanr::cmdstan_model("model-ar1.stan", compile = FALSE)
-fit$format(
-  canonicalize = list("deprecations"), 
-  overwrite_file = TRUE, 
-  backup = FALSE
+# model <- cmdstan_model("model-ar1.stan", compile = FALSE)
+# model$format(
+#   canonicalize = list("deprecations"), 
+#   overwrite_file = TRUE, 
+#   backup = FALSE
+# )
+# model$compile()
+model <- cmdstan_model(exe_file = "model-ar1")
+path <- "models/cmdstanr-example.rds"
+if (!file.exists(path)) {
+post <- model$sample(
+  data = dat, 
+  adapt_delta = 0.99
+) |> 
+  as_draws_df()
+saveRDS(post, path)
+} else {post <- readRDS(path)}
+```
+
+Visualize estimated and defined parameters
+
+```{r}
+post <- post |> 
+  select(starts_with("beta_"), ends_with("_raw")) |> 
+  rename_with(~str_remove(., "_raw"))
+post |> 
+  pivot_longer(everything()) |> 
+  ggplot(aes(value, name)) +
+  scale_x_continuous(
+    "Parameter value"
+  ) +
+  stat_histinterval(
+    normalize = "xy",
+    scale = 0.25, 
+    size = 1
+  ) +
+  geom_point(
+    data = as_tibble(dat) |> 
+      select(sigma:beta_scale) |> 
+      distinct() |> 
+      pivot_longer(everything()),
+    position = position_nudge(y = -0.1),
+    color = "red"
+  ) +
+  theme(
+    axis.title.y = element_blank()
   )
 ```
+
+## brms
+
+```{r}
+dat <- as_tibble(dat)
+dat <- dat |> 
+  mutate(yt1 = lag(y), .by = individual)
+dat
+```

### 2023-10-31 simulation.qmd
Wrap up univariate brms

@@
 saveRDS(post, path)
-} else {post <- readRDS(path)}
+} else {post_cmdstanr <- readRDS(path)}
 ```
@@
 ```{r}
-post <- post |> 
+post_cmdstanr <- post_cmdstanr |> 
   select(starts_with("beta_"), ends_with("_raw")) |> 
   rename_with(~str_remove(., "_raw"))
-post |> 
+post_cmdstanr |> 
   pivot_longer(everything()) |> 
@@
   mutate(yt1 = lag(y), .by = individual)
-dat
+head(dat)
 ```
+
+We specify the model as a missing data model as follows
+
+```{r}
+bform <- bf(y | mi() ~ -1 + mi(mean) + b2*(yt1 - mi(mean))) + 
+  bf(mean | mi() ~ 1 + (1 | individual))
+fit <- brm(
+  bform,
+  data = dat
+)
+```
+
+However this needs some brms syntax edits. We also change the parameter names to fit the blog post
+
+```{r}
+bform <- bf(
+  y | mi() ~ 0 + mi(alpha) + beta * (yt1 - mi(alpha)),
+  alpha | mi() ~ 1 + (1 | individual),
+  beta ~ 1,
+  nl = TRUE
+)
+fit <- brm(
+  bform,
+  data = dat
+)
+```
+
+```{r}
+bform <- bf(
+  y ~ 0 + alpha + beta * (yt1 - alpha),
+  alpha ~ 1 + (1 | individual),
+  beta ~ 1 + (1 | individual),
+  nl = TRUE
+)
+fit <- brm(
+  bform,
+  data = dat,
+  file = "models/brm-example.rds"
+)
+```
+
+Visualize estimated and defined parameters
+
+```{r}
+post_brms <- as_draws_df(fit) |> 
+  select(
+    alpha_hat = b_alpha_Intercept,
+    alpha_scale = sd_individual__alpha_Intercept,
+    beta_hat = b_beta_Intercept,
+    beta_scale = sd_individual__beta_Intercept,
+    sigma
+  )
+
+bind_rows(
+  "Stan" = post_cmdstanr |> 
+  pivot_longer(everything()),
+  "brms" = post_brms |> 
+  pivot_longer(everything()),
+  .id = "Model"
+) |> 
+  mutate(side = if_else(Model == "Stan", "bottom", "top")) |> 
+  ggplot(aes(value, name)) +
+  scale_x_continuous(
+    "Parameter value"
+  ) +
+  scale_color_brewer(
+    palette = "Dark2",
+    aesthetics = c("color", "fill")
+  ) +
+  stat_histinterval(
+    aes(side = side, fill = Model, color = Model),
+    normalize = "xy",
+    scale = 0.5, 
+    size = 1,
+    breaks = 100,
+    slab_alpha = 0.5
+  ) +
+  geom_point(
+    data = as_tibble(dat) |> 
+      select(sigma:beta_scale) |> 
+      distinct() |> 
+      pivot_longer(everything()),
+    aes(color = "Real value", fill = "Real value"),
+    size = 2
+  ) +
+  theme_linedraw() +
+  theme(
+    axis.title.y = element_blank()
+  ) +
+  labs(
+    title = str_glue(
+      "bf(y ~ 0 + alpha + beta * (yt1 - alpha),
+          alpha ~ 1 + (1 | individual),
+          beta ~ 1 + (1 | individual),
+          nl = TRUE)"
+    )
+  ) +
+  facet_wrap("name", scales = "free", ncol = 2)
+```
+
+# Next
+
+Next problem is to generalize this to the multivariate situation where we have some (e.g.) X[t-1] predicting Y[t]. to be continued...

### 2023-11-01 simulation.qmd
Add metadata

@@
 ---
-title: "Simulation"
+title: "Simulations"
+subtitle: "Specification and simulation of brms latent means model"
+authors:
+  - name: Matti Vuorre
+    email: mjvuorre@uvt.nl
+    url: https://vuorre.com
+    orcid: 0000-0001-5052-066X
+    affiliation:
+      - ref: 1
+  - name: Joran Jongerling
+    orcid: 0000-0001-5697-1381
+    affiliation:
+      - ref: 1
+affiliations:
+  - id: 1
+    name: Tilburg University
 execute: 
   cache: true
+  error: true
 format: 
   html:
+    theme: materia
     toc: true
@@
 ---
+
+Here we simulate from, specify, and validate latent means models with brms.
 
@@
 
-In this document we specify, simulate, and validate a univariate latent means brms model. Code based on Joran's blog post at <https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/>.
+# Univariate model
 
-# Data
+The data simulation and Stan code here is based on [Joran's blog post](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/).
 
-We first set up the data simulation. I used [this code](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/#testing) but put it inside a function.
+## Data
+
+We first set up the data simulation.
 
@@
 
-## Stan model
+### Stan model
 
@@
 
-## brms
+### brms
 
@@
 
-# Next
+# Multivariate model
 

### 2023-12-04 examples.qmd
Init Quarto

---
title: Example analyses
---

In this document we illustrate the brms latent means syntax through various examples.

```{r}
#| label: setup
#| cache: false

library(scales)
library(brms)
library(gt)
library(posterior)
library(tidyverse)

N_CORES <- min(parallel::detectCores(), 8)
ITER <- 1500
options(
  mc.cores = N_CORES,
  brms.backend = if(require(cmdstanr)) "cmdstanr",
  brms.threads = N_CORES %/% 4
)

gt2 <- function(x) {
  x |> 
    gt() |> 
    fmt_number(decimals = 2, drop_trailing_zeros = TRUE) |> 
    sub_missing(missing_text = "")
}
```

## Example data

To begin with we load the example data from @mcneishPrimerTwolevelDynamic2020 that we cleaned before. These example data are 100 individuals' reports of urge to smoke (`urge`) and depression (`dep`) at 50 time points. We also have their (non-time-varying) job stress (`js`) and home stress (`hs`). We also create lagged versions of urge to smoke (`u_lag`) and depression (`u_dep`) for use in the analyses. We show these data in @tbl-data-1.

```{r}
#| label: tbl-data-1
#| tbl-cap: First six rows of the example dataset on urge to smoke and depression.

dat <- read_rds("data/mcneish-hamaker.rds") |> 
  mutate(
    u_lag = lag(urge), 
    d_lag = lag(dep), 
    .by = person
  )
dat |> 
  head() |> 
  gt2()
```

## Univariate latent means model

We begin with a univariate model of the urge to smoke. This model examines the degree of autocorrelation in the urge to smoke and how it varies between people. For individual *i* in 1...I=100 and time point *t* in 1...T=50, we model `urge` (U) as normally distributed. We model the mean on person-specific intercepts $\alpha_i$ and slopes of that person's within-person centered `urge` at a previous time point $\phi_i$. Person-specific parameters are modelled as multivariate normal on the population-level parameters (with hats). We do not model correlations between the intercepts and slopes.

First, let us pay some attention to the issue of within-person centering. Instead of decomposing urge to smoke into its within- and between-person components before fitting the model, we use "latent mean centering". What this means is that we estimate the person means along with other model parameters, and subtract those means from the observed values in one step. We refer to the latent person mean variables with an upperscript c (e.g. $U^c_{it}$).

$$
\begin{align}
U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1}, \sigma^2), \\
\begin{bmatrix}
  \alpha_i \\ \phi_i
\end{bmatrix} &\sim MVN\left(
  \begin{bmatrix}
    \bar{\alpha} \\ \bar{\phi}
  \end{bmatrix},
  \begin{pmatrix}
  \tau_\alpha \ &0 \\ 0 \ &\tau_\phi
  \end{pmatrix}
\right).
\end{align}
$$ {#eq-1}

This model is sometimes written with separate "level 1" and "level 2" equations:

$$
\begin{align}
U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1}, \sigma^2), \\
\alpha_i &= \gamma_{00} + u_{0i}, \\
\phi_i &= \gamma_{10} + u_{1i}, \\
\beta_i &= \gamma_{20} + u_{2i}, \\
\mathbf{u} &\sim MVN(\dots)
\end{align}
$$ {#eq-1.2}

We will use the R package brms [@bürkner2017] to estimate this model. The following code chunk shows how to specify this model inside brms' `bf()` ("brmsformula") function. In the first line, we specify a regression equation for `urge`. Everything on the right-hand side of this formula (to the right of the tilde) is treated as a regression coefficient to be estimated from data *unless* it is the exact name of a variable in the data. Thus we will be estimating an `alpha` (intercept) and a `phi` (the autoregressive coefficient).

The only "special" part in this syntax is `(u_lag - alpha)` that does the "latent mean centering" of the lagged urge variable. That is, it just subtracts `alpha` from each lagged value, and then uses the resulting demeaned lagged variable to predict the current urge. The first line can be considered the "level 1" equation or rather the *nonlinear* part of the model.

```{r}
#| code-fold: show

model <- bf(
  urge ~ alpha + phi * (u_lag - alpha),
  alpha ~ 1 + (1 | person),
  phi ~ 1 + (1 | person),
  nl = TRUE
)
```

The second and third lines then specify the "level 2" equation, or rather linear equations to predict the parameters in the above (potentially) nonlinear model. Both regression parameters are modelled on a population level intercept (the gamma in @eq-1.2) and person-specific deviations from it.

The fourth line specifying `nl = TRUE` is critical, because it allows us to specifically name parameters inside `bf()`, and thereby to e.g. construct the latent mean centered variable on the first row.

Note that the linear models of `alpha` and `phi` are identical. We can therefore use the following shorthand to assign them the same linear model:

```{r}
#| code-fold: show

model <- bf(
  urge ~ alpha + phi * (u_lag - alpha),
  alpha + phi ~ 1 + (1 | person),
  nl = TRUE
)
```

We will use that shorthand going forward. We could also indicate the distribution that we assume for the data. But in this work we model everything as gaussian, which is the software default and thus doesn't need to be separately indicated.

We then sample from the model. Everything from here on is standard operating procedure.

```{r}
fit <- brm(
  model,
  data = dat,
  iter = ITER,
  file = "cache/brm-example-univariate"
)
```

The object `fit` now contains the estimated model (the data, posterior samples, and lots of brms-specific information). We can call `summary(fit)` to see a default summary of the model.

```{r}
summary(fit)
```

The first few rows above print information about the model (the formulas, data, and number of posterior samples). Then, "Group-Level Effects" are standard deviations (and correlations, if estimated) of the parameters that we allowed to vary across individuals (as indicated by `~person`). For each of those parameters, one row indicates its posterior summary statistics; "Estimate" is the posterior mean, "Est.Error" is the posterior standard deviation, "l-" and "u-95% CI" are the lower and upper bounds of the 95% credibility interval (so the 2.5 and 97.5 percentiles of the posterior samples). Then, Rhat is the convergence metric which should be smaller than 1.05 (optimally 1.00) to indicate that the estimation algorithm has converged. "Bulk_" and "Tail_ESS" indicate the effective sample sizes of the posterior draws, and should be pretty large.

The "Population-Level Effects" indicate the same information but for the means of the person-specific parameters' distributions; or the "fixed effects". For the average person, there is a positive autocorrelation in these data. Finally, the "Family Specific Parameters" indicate parameters that are specific to the outcome distribution. We used the default gaussian distribution, and thus get an estimated residual standard deviation.

Going forward we will create a small function to print out model summaries. It will take samples of the population level, group-level, and family-specific parameters, and return their 50th (median), 2.5th, and 97.5th quantiles.

```{r}
sm <- function(fitted_model) {
  fitted_model |> 
    as_draws_df(variable = c("b_", "sd_", "sigma"), regex = TRUE) |> 
    summarise_draws(
      ~quantile2(.x, c(.5, .025, .975))
    )
}
```

We show the results in @tbl-model-1.

```{r}
#| label: tbl-model-1
#| tbl-cap: Summaries of main parameters from the example univariate model.
fit |> 
  sm() |> 
  gt2()
```

## Multilevel AR(1) Model

We then turn to models with more than one variable of interest. In what follows we replicate "Two-Level Dynamic Structural Equation Models" (equations 4 to 7) from @mcneishPrimerTwolevelDynamic2020.

We start with the two-level AR(1) model in @mcneishPrimerTwolevelDynamic2020 (equations 4a-c). This model is similar to our example model from above, but we also now include a regression coefficient for within-person centered depression, where this centering is on the latent mean of depression. The model is

$$
\begin{align}
U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1} + \beta_i D^c_{it}, \sigma^2), \\
U^{c}_{it} &= U^{\text{raw}}_{it} - \alpha^U_i, \\
D^{c}_{it} &= D^{\text{raw}}_{it} - \alpha^D_i, \\
\begin{bmatrix}
  \alpha^U_i \\ \alpha^D_i \\ \phi_i \\ \beta_i
\end{bmatrix} &\sim MVN\left(
  \begin{bmatrix}
    \bar{\alpha^U} \\ \bar{\alpha^D} \\ \bar{\phi} \\ \bar{\beta_i}
  \end{bmatrix},
  \begin{pmatrix}
    \tau_\alpha^U \ &0 \ &0 &0 \\ 
    0 \ &\tau_\alpha^D \ &0 \ &0 \\ 
    0 \ &0 \ &\tau_\beta \ &0 \\
    0 \ &0 \ &0 \ &\tau_\phi
  \end{pmatrix}
\right)
\end{align}
$$ {#eq-2}

We then see from @eq-2 that we need to refer to different outcomes' parameters across model formulas. That is, when predicting the urge to smoke, we need a way to refer to the (latent) mean of depression so that we can appropriately center the depression predictor. Currently brms does not support sharing parameters across formulas for different outcomes, but we can overcome this limitation with a small data wrangling trick

That is, we "stack" our data into the long format with respect to the two different outcomes, urge to smoke and depression. Then, on each row we have all variables from that measurement occasion, in addition to new ones that indicate the value of the outcome, and which outcome it refers to (@tbl-data-2).

```{r}
#| label: tbl-data-2
#| tbl-cap: Rearranged data for multivariate models.

dat <- dat |> 
  select(-js, -hs) |> 
  pivot_longer(c(urge, dep), names_to = "outcome") |> 
  mutate(
    i_urge = if_else(outcome == "urge", 1, 0),
    i_dep = if_else(outcome == "dep", 1, 0)
  ) |> 
  # Include predictors from each row
  left_join(dat)

dat |> 
  head() |> 
  gt2()
```

Now we can model `value` and use `i_urge` and `i_dep` to control whether a parameter should apply to the outcome on a particular row. So here we want alpha, phi, and beta to apply to urge, but depb to depression. Then we can use depb to create the predictor for urge on the fly.

```{r}
#| code-fold: show

bform <- bf(
  value ~ 
    i_urge * (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2)) + 
    i_dep * alpha2,
  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
  alpha1 + phi + beta + alpha2 ~ 1 + (1 | person),
  sigma1 + sigma2 ~ 1,
  nl = TRUE
)
```

Sample from model

```{r}
fit <- brm(
  bform,
  data = dat,
  iter = ITER,
  control = list(adapt_delta = 0.95),
  file = "cache/brm-example-4"
)
```

Compare model summary to @mcneishPrimerTwolevelDynamic2020. We can see the estimates match to within differences in priors and MCSE. Note in the code I transform standard deviations by first exponentiating draws of residual standard deviations, and then square to put them on the variance scale as in @mcneishPrimerTwolevelDynamic2020.

```{r}
source("mplus-parameters.R")
as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
  mutate(
    across(starts_with("sd_"), ~.^2),
    across(starts_with("b_sigma"), ~exp(.)^2)
  ) |> 
  summarise_draws(
    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
      number(.01) |> 
      str_glue_data("{q50} [{q2.5}, {q97.5}]")
  ) |> 
  mutate(
    variable = str_replace(variable, "sd_person__", "var_"),
    mplus = mplus$equation.4$mplus
  ) |> 
  gt2()
```

## Time-Invariant Covariates

This is as above but includes time-invariant covariates js and hs, and the between-person effect of depression on urge. I label the latter as "gamma03" as in Equation 5 in @mcneishPrimerTwolevelDynamic2020. Note also that I must include it in the regression equation for the outcome, and not in the "level 2" equation because the latter must be "linear formulas" in brms. There's probably other ways of doing it too.

```{r}
#| code-fold: show

bform <- bf(
  value ~ 
    i_urge * 
    (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2) + gamma03 * alpha2) + 
    i_dep * alpha2,
  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
  alpha1 + phi + beta ~ 1 + js + hs + (1 | person),
  alpha2 ~ 1 + (1 | person),
  gamma03 + sigma1 + sigma2 ~ 1,
  nl = TRUE
)
```

With flat priors the gamma03 parameter is wonky so I set a normal(0, 3) on it.

```{r}
p <- get_prior(bform, dat)
p <- p |> 
  mutate(
    prior = case_when(
      class == "b" & nlpar != "gamma03" ~ "normal(0, 1)",
      class == "b" & nlpar == "gamma03" ~ "normal(0, 3)",
      class == "sd" ~ "student_t(7, 0, 1)",
      TRUE ~ prior
    )
  )
```

Sample from model

```{r}
fit <- brm(
  bform,
  data = dat,
  prior = p,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "cache/brm-example-5",
)
```

Model summary.

```{r}
as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
  mutate(
    across(starts_with("sd_"), ~.^2),
    across(starts_with("b_sigma"), ~exp(.)^2)
  ) |> 
  summarise_draws(
    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
      number(.01) |> 
      str_glue_data("{q50} [{q2.5}, {q97.5}]")
  ) |> 
  mutate(
    variable = str_replace(variable, "sd_person__", "var_"),
    mplus = mplus$equation.5$mplus
  ) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

## Multilevel Location-Scale model

Equations 5 and 6 in @mcneishPrimerTwolevelDynamic2020.

```{r}
#| code-fold: show

bform <- bf(
  value ~ 
    i_urge * 
    (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2) + gamma03 * alpha2) + 
    i_dep * alpha2,
  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
  alpha1 + phi + beta ~ 1 + js + hs + (1 | person),
  alpha2 + sigma1 + sigma2 ~ 1 + (1 | person),
  gamma03 ~ 1,
  nl = TRUE
)
```

Sample from model

```{r}
fit <- brm(
  bform,
  data = dat,
  prior = p,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "cache/brm-example-6"
)
```

Model summary. These match.

```{r}
as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
  mutate(
    across(starts_with("sd_"), ~.^2),
    across(starts_with("b_sigma"), ~exp(.)^2)
  ) |> 
  summarise_draws(
    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
      number(.01) |> 
      str_glue_data("{q50} [{q2.5}, {q97.5}]")
  ) |> 
  mutate(
    variable = str_replace(variable, "sd_person__", "var_")
  ) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

## Multilevel VAR(1) Model

Equations 7 and 8 in @mcneishPrimerTwolevelDynamic2020.

```{r}
#| code-fold: show

bform <- bf(
  value ~ 
    i_urge * (alpha1 + phi1 * (u_lag - alpha1) + phi4 * (d_lag - alpha2)) + 
    i_dep * (alpha2 + phi2 * (d_lag - alpha2) + phi3 * (u_lag - alpha1)),
  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
  alpha1 + alpha2 + phi1 + phi2 + phi3 + phi4 + sigma1 + sigma2 ~ 1 + (1 | person),
  nl = TRUE
)
```

Sample

```{r}
fit <- brm(
  bform,
  data = dat,
  iter = ITER,
  control = list(adapt_delta = 0.99),
  file = "cache/brm-example-7"
)
```

Summarise. I wasn't sure how to transform the variance components here to match the variances on the link function scale so left them as is.

```{r}
as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
  mutate(
    across(starts_with("sd_"), ~.^2)
  ) |> 
  summarise_draws(
    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
      number(.01) |> 
      str_glue_data("{q50} [{q2.5}, {q97.5}]")
  ) |> 
  mutate(
    variable = str_replace(variable, "sd_person__", "var_"),
    mplus = mplus$equation.7$mplus
  ) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

## Conclusion

We estimated models 4-7 in @mcneishPrimerTwolevelDynamic2020 with brms and got very similar if not identical results.

### 2023-12-04 simulation.qmd
Init Quarto

--- 2023-10-31 simulation.qmd
@@
 ---
-title: "Simulation"
+title: Simulation
+subtitle: Specification and parameter recovery simulation of brms latent means model
 execute: 
-  cache: true
-format: 
-  html:
-    toc: true
-    code-fold: true
+  error: true
+  message: false
+  warning: false
 ---
 
-```{r}
+Here we simulate from, specify, and validate latent means models with brms.
+
+```{r echo = FALSE}
 #| label: options
+#| output: false
 
@@
 library(mvtnorm)
+library(scales)
+library(gt)
 library(cmdstanr)
 library(ggdist)
+library(future)
 library(brms)
@@
 
-In this document we specify, simulate, and validate a univariate latent means brms model. Code based on Joran's blog post at <https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/>.
+## Univariate model
 
-# Data
+The data simulation and Stan code here is based on [Joran's blog post](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/).
 
-We first set up the data simulation. I used [this code](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/#testing) but put it inside a function.
+The model is for each individual i and timepoint t,
+
+y_it ~ normal(eta_it, sigma)
+eta_it = (alpha_hat + alpha_i) + (beta_hat + beta_i) * (y_i[t-1])
+[alpha_i, beta_i] ~ mvn([0, 0], Sigma)
+
+That is we model y on an intercept and an autoregressive coefficient with varying intercepts and slopes over people.
+
+### Data
+
+We first set up the data simulation.
 
@@
 
-## Stan model
+### Stan
 
-Let us then compile and sample from the cmdstanr model
+Let us then compile and sample from the Stan model (`model-ar1.stan`) with cmdstanr
 
 ```{r}
-# model <- cmdstan_model("model-ar1.stan", compile = FALSE)
-# model$format(
-#   canonicalize = list("deprecations"), 
-#   overwrite_file = TRUE, 
-#   backup = FALSE
-# )
-# model$compile()
-model <- cmdstan_model(exe_file = "model-ar1")
-path <- "models/cmdstanr-example.rds"
+#| output: false
+
+path <- "cache/cmdstanr-example.rds"
 if (!file.exists(path)) {
-post <- model$sample(
+  model <- cmdstan_model("model-ar1.stan", compile = FALSE)
+  model$format(
+    canonicalize = list("deprecations"),
+    overwrite_file = TRUE,
+    backup = FALSE
+  )
+  model$compile()
+  post_cmdstanr <- model$sample(
     data = dat, 
@@
     as_draws_df()
-saveRDS(post, path)
-} else {post <- readRDS(path)}
+  saveRDS(post_cmdstanr, path)
+} else {
+  post_cmdstanr <- readRDS(path)
+}
 ```
 
-Visualize estimated and defined parameters
+Posterior summary:
 
 ```{r}
-post <- post |> 
-  select(starts_with("beta_"), ends_with("_raw")) |> 
+post_cmdstanr <- post_cmdstanr |> 
+  select(
+    alpha_hat_raw, alpha_scale_raw, 
+    beta_hat, beta_scale, 
+    sigma_raw
+  ) |> 
   rename_with(~str_remove(., "_raw"))
-post |> 
-  pivot_longer(everything()) |> 
-  ggplot(aes(value, name)) +
-  scale_x_continuous(
-    "Parameter value"
-  ) +
-  stat_histinterval(
-    normalize = "xy",
-    scale = 0.25, 
-    size = 1
-  ) +
-  geom_point(
-    data = as_tibble(dat) |> 
-      select(sigma:beta_scale) |> 
-      distinct() |> 
-      pivot_longer(everything()),
-    position = position_nudge(y = -0.1),
-    color = "red"
-  ) +
-  theme(
-    axis.title.y = element_blank()
-  )
+
+summarize_draws(post_cmdstanr) |> 
+  gt() |> 
+  fmt_number(decimals = 2)
 ```
 
-## brms
+### brms
+
+brms takes a data frame, so we convert the list here
 
@@
   mutate(yt1 = lag(y), .by = individual)
-dat
+head(dat) |> 
+  gt() |> 
+  fmt_number(decimals = 2)
 ```
+
+#### Model specification
+
+We can specify this model using brms' nonlinear formulas. Note that we created the `yt1` variable (i.e. y[t-1]) above.
+
+```{r}
+#| output: false
+
+bform <- bf(
+  y ~ 0 + alpha + beta * (yt1 - alpha),
+  alpha ~ 1 + (1 | individual),
+  beta ~ 1 + (1 | individual),
+  nl = TRUE
+)
+fit_brms <- brm(
+  bform,
+  data = dat,
+  file = "cache/brm-example.rds"
+)
+```
+
+We see that the estimated parameters are very close to coefficients from the Stan model
+
+```{r}
+post_brms <- as_draws_df(fit_brms) |> 
+  select(
+    alpha_hat = b_alpha_Intercept,
+    alpha_scale = sd_individual__alpha_Intercept,
+    beta_hat = b_beta_Intercept,
+    beta_scale = sd_individual__beta_Intercept,
+    sigma
+  )
+
+bind_rows(
+  "Stan" = summarize_draws(post_cmdstanr),
+  "brms" = summarize_draws(post_brms),
+  .id = "Model"
+) |> 
+  arrange(variable, Model) |> 
+  gt() |> 
+  fmt_number(decimals = 2)
+```
+
+### Simulation
+
+Above, we estimated the Stan and brms models once. Here, we evaluate whether they recover the specified parameters with a small simulation.
+
+```{r}
+path <- "cache/sim-univariate.rds"
+if (!file.exists(path)) {
+  # Ensure that models are compiled
+  model_stan <- cmdstan_model("model-ar1.stan", compile = TRUE)
+  fit_brms <- brm(
+    bform,
+    data = dat,
+    chains = 0,
+    threads = 1
+  )
+
+  # Define simulation outputs
+  K <- 30
+  out <- vector("list", length = K)
+
+  # Parallelization scheme:
+  # Each of two workers will launch a session that runs 4 chains in parallel
+  # resulting in total 8 cores used
+  plan(multisession, workers = 2)
+
+  for (k in 1:K) {
+    this_sample <- simulate()
+
+    this_stan <- future({
+      model_stan$sample(
+        data = this_sample, 
+        adapt_delta = 0.99, 
+        parallel_chains = 4,
+        iter_sampling = 2000,
+        refresh = 0, 
+        show_messages = FALSE
+      )
+    }, seed = TRUE)
+
+    this_brms <- future({
+      update(
+        fit_brms, 
+        newdata = as_tibble(this_sample) |> 
+          mutate(yt1 = lag(y), .by = individual),
+        control = list(adapt_delta = 0.99),
+        cores = 4,
+        iter = 2000,
+        recompile = FALSE,
+        refresh = 0,
+        silent = 2
+      )
+    }, seed = TRUE)
+
+    this_stan <- value(this_stan) 
+    this_stan <- this_stan |> 
+      as_draws_df() |> 
+      select(
+        alpha_hat_raw, alpha_scale_raw, 
+        beta_hat, beta_scale, 
+        sigma_raw
+      ) |> 
+      rename_with(~str_remove(., "_raw")) |> 
+      summarize_draws()
+
+    this_brms <- value(this_brms) 
+    this_brms <- this_brms |> 
+      as_draws_df() |> 
+      select(
+        alpha_hat = b_alpha_Intercept,
+        alpha_scale = sd_individual__alpha_Intercept,
+        beta_hat = b_beta_Intercept,
+        beta_scale = sd_individual__beta_Intercept,
+        sigma
+      ) |> 
+      summarize_draws()
+
+    out[[k]] <- bind_rows(
+      "Stan" = this_stan,
+      "brms" = this_brms,
+      .id = "model"
+    )
+  }
+
+  out <- bind_rows(out, .id = "k")
+  saveRDS(out, path)
+} else {out <- readRDS(path)}
+```
+
+#### Results
+
+```{r}
+#| label: tbl-simulation
+
+real_values <- tribble(
+  ~variable, ~true,
+  "sigma", 1, 
+  "alpha_hat", 4, 
+  "alpha_scale", 1, 
+  "beta_hat", 0.4, 
+  "beta_scale", 0.1
+)
+out |> 
+  summarise(
+    mean = str_glue("{number(mean(mean), .01)} ({number(sd(mean), .01)})"),
+    .by = c(model, variable)
+  ) |> 
+  pivot_wider(
+    names_from = model, values_from = mean
+  ) |> 
+  left_join(
+    real_values
+  ) |> 
+  gt() |> 
+  fmt_number(decimals = 1) |> 
+  fmt_missing(missing_text = " ")
+```
+
+## Conclusion
+
+This small simulation shows that brms perfectly recovers parameters of a simple univariate autoregressive latent variable model.
+The mean estimates over 30 simulation iterations were identical for brms and Stan to the second significant digit, and within MCSE of the true values.

### 2023-12-04 mplus-parameters.R

@@ -1,0 +1,21 @@
+library(MplusAutomation)
+
+mplus <- readModels("data/mh/Output Files")
+mplus <- map(mplus, ~pluck(., "parameters") |> pluck("unstandardized"))
+names(mplus) <- str_extract(names(mplus), "equation\\.[0-9]")
+
+mplus$equation.4 <- mplus$equation.4[c(3, 5, 6, 4, 2, 1, 7, 9, 10, 8), ]
+mplus$equation.5 <- mplus$equation.5[c(11, 7, 8, 12, 3, 4, 13, 5, 6, 10, 9, 2, 1, 15, 16, 17, 14), ]
+mplus$equation.6 # This is missing parameters
+mplus$equation.7
+
+mplus <- map(mplus, ~rename(., q50 = est, q2.5 = lower_2.5ci, q97.5 = upper_2.5ci))
+
+mplus <- map(
+  mplus, 
+  ~mutate(
+    .,
+    across(where(is.numeric), ~number(., .01)),
+    mplus = str_glue("{q50} [{q2.5}, {q97.5}]")
+  )
+)

### 2023-12-04 data.qmd

---
title: Example datasets
---

Here we download and wrangle the example datasets used in our analyses.

```{r}
#| label: setup

library(tidyverse)
```

Throughout, we use an example dataset from "A primer on two-level dynamic structural equation models for intensive longitudinal data in Mplus" [@mcneishPrimerTwolevelDynamic2020]. Those data are available on the OSF (<https://osf.io/wuprx/>). Here we download, clean, and save the data for use in our analyses.

```{r}
#| label: mh-download

path <- "data/mh.zip"
url <- "https://files.osf.io/v1/resources/wuprx/providers/osfstorage/5bfc839601593f0016774697/?zip="
if (!file.exists(path)) {
  dir.create(dirname(path), showWarnings = FALSE)
  download.file(
    url,
    destfile = path
  )
  unzip(path, exdir = "data/mh/")
}
```

That extracted all the data and Mplus files to `data/` in the project root. We then wrangle the main data file and save for our analyses:

```{r}
#| label: mh-wrangle

d <- read_csv(
  "data/mh/Data/Two-Level Data.csv", 
  col_names = c("urge", "dep", "js", "hs", "person", "time")
) %>% 
  select(
    person, time, js, hs, urge, dep
  ) |> 
  mutate(
    person = factor(person),
    time = as.integer(time-1)
  )
write_rds(d, "data/mcneish-hamaker.rds")
```

### 2023-12-04 index.qmd

---
author:
  - name: Matti Vuorre
    url: https://vuorre.com
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    affiliation:
      - ref: 1
  - name: Ethan McCormick
    affiliation:
      - ref: 3
  - name: Mauricio Garnier-Villarreal
    affiliation:
      - ref: 4
  - name: Joran Jongerling
    orcid: 0000-0001-5697-1381
    affiliation:
      - ref: 2
affiliations:
  - id: 1
    name: Tilburg University
  - id: 2
    name: Tilburg University
  - id: 3
    name: todo
  - id: 4
    name: todo
---

{{< include README.qmd >}}

### 2023-12-04 README.qmd

---
title: Introduction to latent means models in R
format: gfm
---

Repository for materials supporting our manuscript discussing practical implementations of latent mean centering in multilevel models of longitudinal psychological data.

- Repo on GitHub: <https://github.com/mvuorre/latent-mean-centering-ms>
- Archived repo: tbd
- Online: tbd

## Reproduce

Everything is coded in [R](https://cran.r-project.org/) and bunched together as a [Quarto](https://quarto.org/) website. To reproduce, clone the repo and render the Quarto documents:

```bash
# Download materials
git clone https://github.com/mvuorre/latent-mean-centering-ms.git
cd latent-mean-centering-ms

# Set up R environment
Rscript -e 'renv::restore(prompt = FALSE)'

# Run code and render outputs
quarto render
```

## Contribute

Contributions preferably via pull requests.

### 2023-12-04 manuscript.qmd

--- 2023-02-14 ms.Rmd
@@
 ---
-title: 'Bias no more: Introduction to latent group-mean centering in hierarchical bayesian models'
-shorttitle: Latent group-mean centering
-leftheader: Latent group-mean centering
+title: Estimating latent means model with regression models in R
 author:
   - name: Matti Vuorre
-    affiliation: 1
-    corresponding: yes
-    address: Tilburg University
-    email: m.j.vuorre@tilburguniversity.edu
-  - name: Author 2
-    affiliation: 2
-    address: 
-    email:
-
+    url: https://vuorre.com
+    orcid: 0000-0001-5052-066X
+    email: mjvuorre@uvt.nl
     affiliation:
+      - ref: 1
+  - name: Ethan McCormick
+    affiliation:
+      - ref: 3
+  - name: Mauricio Garnier-Villarreal
+    affiliation:
+      - ref: 4
+  - name: Joran Jongerling
+    orcid: 0000-0001-5697-1381
+    affiliation:
+      - ref: 1
+affiliations:
   - id: 1
-    institution: Tilburg University
-  - id: 2
-    institution: Institution 2
-
+    name: Tilburg University
+  - id: 3
+    name: todo
+  - id: 4
+    name: todo
 authornote: |
-  \noindent \textit{This manuscript is not yet peer-reviewed.}
-
+  Work in progress.
 abstract: |
@@
 Latent mean centering provides unbiased estimates of within- and between-person associations, and we hope that the availability and flexibility of our proposed approach motivates researchers to apply this method widely in longitudinal and other multilevel modelling scenarios.
+keywords: "longitudinal data, repeated measures, centering, R, brms"
+bibliography: references.bib
 
-keywords: longitudinal data, repeated measures, centering, well-being, technology effects
-wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
-bibliography: "references.bib"
-floatsintext: yes
-linenumbers: no
-draft: no
-mask: no
-figurelist: no
-tablelist: no
-footnotelist: no
-documentclass: apa7
-classoption: jou
-output: 
-  papaja::apa6_pdf:
-    number_sections: false
-    keep_tex: false
-  papaja::apa6_docx: 
-    number_sections: false
+format: 
+  html: default
+  pdf: default
+  docx: default
 ---
@@
 #| label: tab-data
+
 d %>% 
@@
 
-```{r}
-#| eval: false
-model <- bf(
-  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
-  b0happy ~ 1 + (1 | person),
-  b1happy ~ 1 + (1 | person),
-  b2happy ~ 1 + (1 | person),
-  b0sm ~ 1 + (1 | person),
-  nl = TRUE,
-  center = FALSE
-) +
-  gaussian()
-
-# Set priors
-p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)
-
-fit <- brm(
-  model,
-  data = d,
-  prior = p,
-  cores = MAX_CORES,
-  threads = THREADS,
-  iter = ITER,
-  control = list(adapt_delta = 0.99),
-  file = "models/brm-fit-1"
-)
-summary(fit, prior = TRUE)
-```
-
-# Simulation study
-
-```{r}
-#| label: fig-me-in-means
-#| eval: false
-#| include: false
-#| fig.cap: Illustration of the risks associated with using observed means for group-mean centering, and how those are avoided by using latent means to group-mean center instead.
-
-# A suggestion for an illustrative figure, showing risks (bias) of observed mean centering, but also how latent mean centering avoids them.
-
-tibble(
-  n = seq(1, 100, by = .1),
-  s = 1/sqrt(n)
-  ) %>% 
-  ggplot(aes(n, s)) +
-  scale_y_continuous(
-    "Measurement error in observed mean",
-    limits = 0:1, 
-    breaks = extended_breaks(),
-    expand = expansion(c(.02, .05))
-  ) +
-  scale_x_continuous(
-    "Number of measurements",
-    expand = expansion(.025, .025)
-  ) +
-  # Theoretical measurement error in mean
-  geom_line()
-  # More lines or panels
-  # Observed dynamic panel bias from a simulation
-  # geom_line(something something)
-  # Observed bias in other covariate from a simulation
-  # geom_line(also something something)
-  # Latent mean centering results--lines that are not biased
-  # geom_line(...)
-```
-
 # Discussion
@@
 Data Curation: MV  
-Writing -- Original Draft: MV
-Writing -- Review & Editing: MV
+Writing---Original Draft: MV  
+Writing---Review & Editing: MV  
 Visualization: MV  
-Supervision: MV
 Project Administration: MV
@@
 The author(s) declare that there were no conflicts of interest with respect to the authorship or the publication of this article.
-
-# Acknowledgements
-
-We thank everyone who contributed to this discussion on the Stan discussion board (https://web.archive.org/web/20230207111825/https://discourse.mc-stan.org/t/latent-mean-centering-latent-covariate-models-in-brms/29424).
-
-# References
-
-::: {#refs custom-style="Bibliography"}
-:::
-
-<!-- Create an appendix with new page, table, and figure numbers. -->
-
-\clearpage
-\onecolumn
-
-# Appendix Z: Simulation study
-
-<!-- Prepend table & figure numbers with S and reset counters -->
-\setcounter{page}{1}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-\renewcommand{\thefigure}{A\arabic{figure}}
-
-Hereafter be monsters and dragons
-
-```{r data-create}
-# Simple simulation scheme
-# happiness predicted from within-person social media use, and happiness(t-1).
-library(MASS)
-library(faux)
-library(lme4)
-
-# Numbers of people and observations
-n_person <- 50
-n_time <- 5
-# Population level parameters
-beta <- c(
-  4, # Happiness intercept
-  .2, # happy <- social media within
-  .25, # happy <- happy(t-1)
-  30 # Social media intercept
-)
-names(beta) <- c("b0happy", "b1happy", "b2happy", "b0sm")
-# Standard deviations across people (in same order)
-# Note b3happy is "fixed" (0)
-tau <- c(1, .2, .075, 5)
-# Random effect standard deviations (same for all pairs)
-rho <- .3
-# Standard deviations of observations
-sigma <- c(sigmahappy = 0.5, sigmasm = 5)
-
-# Generate person-level parameters
-x <- rnorm_multi(
-  n = n_person,
-  mu = beta,
-  sd = tau,
-  r = rho,
-  empirical = TRUE
-) %>% 
-  mutate(
-    sigmahappy = sigma[1],
-    sigmasm = sigma[2]
-  )
-
-# Repeate rows and create outcomes
-x <- x %>% 
-  rownames_to_column("person") %>% 
-  crossing(time = 1:n_time) %>% 
-  mutate(
-    sm = rnorm(nrow(.), b0sm, sigmasm),
-    happy = rnorm(nrow(.), b0happy + b1happy*(sm - b0sm), sigmahappy)
-  ) %>% 
-  mutate(
-    happy1 = lag(happy),
-    happy = happy + happy1*b2happy,
-    .by = person
-  )
-
-x <- x %>% 
-  mutate(
-    smb = mean(sm),
-    smw = sm - smb,
-    .by = person
-  )
-y <- lmer(
-  happy ~ smw + happy1 + (smw + happy1 | person),
-  data = x
-)
-```
-
-```{r}
-model <- bf(
-  happy ~ b0happy + b1happy*(happy1 - b0happy) + b2happy*(sm - b0sm),
-  b0happy ~ 1 + (1 | person),
-  b1happy ~ 1 + (1 | person),
-  b2happy ~ 1 + (1 | person),
-  b0sm ~ 1 + (1 | person),
-  nl = TRUE,
-  center = FALSE
-) +
-  gaussian()
-
-# Set priors
-p <- prior(normal(30, 15), class = b, coef = Intercept, nlpar = b0sm)
-
-fit <- brm(
-  model,
-  data = d,
-  prior = p,
-  cores = MAX_CORES,
-  threads = THREADS,
-  iter = ITER,
-  control = list(adapt_delta = 0.99),
-  file = "models/brm-fit-1"
-)
-summary(fit, prior = TRUE)
-```

### 2024-01-25 data.qmd
Organize materials better (?)

@@
 ```
+
+```{r data-process}
+# The data is cited in the paper as "The dataset generated and analysed during the current study is available in Figshare." (https://uvaauas.figshare.com/articles/dataset/Dataset_belonging_to_Beyens_et_al_2020_The_effect_of_social_media_on_well-being_differs_from_adolescent_to_adolescent/12497990)
+
+# Download data if not yet downloaded
+path <- "data/beyens-et-al-2020.sav"
+if (!file.exists(path)) {
+  dir.create("data", FALSE)
+  # Data from figshare
+  download.file(
+    "https://uvaauas.figshare.com/ndownloader/files/24028271", 
+    path
+  )
+  # Code from OSF
+  download.file(
+    "https://files.de-1.osf.io/v1/resources/bvfuw/providers/osfstorage/5df9effa64e19d000d0f51da/?zip=",
+    "data/Analyses.zip"
+  )
+  unzip(
+    "data/Analyses.zip",
+    exdir = "data/Analyses"
+  )
+}
+
+# Active, Passive, and Total were sums of the platform (e.g. whatsapp) times
+d <- read_spss(path) %>% 
+  mutate(
+    class = factor(ClassNr),
+    person = factor(EthicaID),
+    # I think they divided this by ten
+    time = (OccNrR - 1),
+    # active = rowSums(select(., contains("Act")), na.rm = TRUE),
+    sm = rowSums(select(., contains("Pas")), na.rm = TRUE) / 60,
+    wb = as.numeric(AffWB) / 7, 
+    .keep = "none"
+  ) %>% 
+  arrange(person, class, time)
+
+# Remove times in excess of 60min (1 hour)
+d <- d %>% 
+  mutate(
+    across(c(sm), ~if_else(. > 1, 1, .))
+  )
+
+# Centered and lagged variables
+d <- d %>% 
+  mutate(
+    sm_b = mean(sm, na.rm = TRUE),
+    sm_c = sm - mean(sm, na.rm = TRUE),
+    across(
+      c(wb),
+      .fns = list(lag = ~lag(.))
+    ),
+    .by = person
+  )
+
+d
+
+write_rds(d, "data/beyens-etal.rds")
+```

### 2024-01-25 examples.qmd
Organize materials better (?)

@@
 
-N_CORES <- min(parallel::detectCores(), 8)
-ITER <- 1500
+# HMC and parallel computation options
+MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"), 4)
 options(
-  mc.cores = N_CORES,
-  brms.backend = if(require(cmdstanr)) "cmdstanr",
-  brms.threads = N_CORES %/% 4
+  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
+  brms.threads = MAX_CORES %/% 4,
+  mc.cores = MAX_CORES
 )
@@
   data = dat,
-  iter = ITER,
   file = "cache/brm-example-univariate"
@@
   data = dat,
-  iter = ITER,
   control = list(adapt_delta = 0.95),
@@
   prior = p,
-  iter = ITER,
   control = list(adapt_delta = 0.99),
@@
   data = dat,
-  iter = ITER,
   control = list(adapt_delta = 0.99),
@@
   ) |> 
-  gt() |> 
-  fmt_number(decimals = 2)
+  gt2()
+```
+
+## Social media use
+
+```{r}
+d <- read_rds("data/beyens-etal.rds")
+
+# Rescale everything to the unit interval
+d <- d |> 
+  mutate(
+    across(c(sm), ~./60),
+    across(starts_with("wb"), ~./7)
+    )
+
+d <- d |> 
+  pivot_longer(c(sm, wb), names_to = "outcome") |> 
+  mutate(
+    i_wb = if_else(outcome == "wb", 1, 0),
+    i_sm = if_else(outcome == "sm", 1, 0)
+  ) |> 
+  # Include predictors from each row
+  left_join(d)
+```
+
+We regress wellbeing on an autocorrelative term and within person effect of passive social media use.
+Include correlations between person parameters; one of these will be the between-person relationship.
+
+```{r}
+bform <- bf(
+  value ~ 
+    i_wb * (alpha1 + phi * (wb_lag - alpha1) + beta * (sm - alpha2)) + 
+    i_sm * alpha2,
+  nlf(sigma ~ i_wb * sigma1 + i_sm * sigma2),
+  alpha1 + alpha2 + phi + beta ~ 1 + (1 |p| person),
+  sigma1 + sigma2 ~ 1,
+  nl = TRUE
+)
+```
+
+```{r}
+fit <- brm(
+  bform,
+  data = d,
+  control = list(adapt_delta = 0.95),
+  file = "cache/brm-example-sm-1"
+)
+```
+
+```{r}
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2),
+    across(starts_with("b_sigma"), ~exp(.)^2)
+  ) |> 
+  summarise_draws(
+    mean, sd, ~quantile2(.x, c(.025, .975))
+  )
+  gt2()
+```
+
+## Social media use 2
+
+Same as above but a three-level model class > person > observation.
+
+```{r}
+bform <- bf(
+  value ~ 
+    i_wb * (alpha1 + phi * (wb_lag - alpha1) + beta * (sm - alpha2)) + 
+    i_sm * alpha2,
+  nlf(sigma ~ i_wb * sigma1 + i_sm * sigma2),
+  alpha1 + alpha2 + phi + beta ~ 1 + (1 |p| person) + (1 |c| class),
+  sigma1 + sigma2 ~ 1,
+  nl = TRUE
+)
+```
+
+```{r}
+fit <- brm(
+  bform,
+  data = d,
+  control = list(adapt_delta = 0.95),
+  file = "cache/brm-example-sm-threelevel"
+)
+```
+
+```{r}
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2),
+    across(starts_with("b_sigma"), ~exp(.)^2)
+  ) |> 
+  summarise_draws(
+    mean, sd, ~quantile2(.x, c(.025, .975))
+  )
+  gt2()
 ```

### 2024-01-25 manuscript.qmd
Organize materials better (?)

@@
 ---
-title: Estimating latent means model with regression models in R
+title: Latent variable and dynamic structural equation models with R and brms
 author:
@@
 Latent mean centering provides unbiased estimates of within- and between-person associations, and we hope that the availability and flexibility of our proposed approach motivates researchers to apply this method widely in longitudinal and other multilevel modelling scenarios.
-keywords: "longitudinal data, repeated measures, centering, R, brms"
+keywords: "longitudinal data, repeated measures, centering, latent variables, R"
 bibliography: references.bib
-
+csl: apa.csl
 format: 
+  docx: default
   html: default
-  pdf: default
-  docx: default
 ---
@@
 ```{r setup, include = FALSE}
-library(papaja)
-library(citr)
-library(parameters)
+library(scales)
+library(gt)
+library(knitr)
 library(posterior)
-library(lme4)
-library(knitr)
-library(scales)
 library(ggdist)
@@
 
-# Document options
-opts_chunk$set(
-  cache = TRUE,
-  echo = FALSE
+# HMC and parallel computation options
+MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"), 4)
+options(
+  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
+  brms.threads = MAX_CORES %/% 4,
+  mc.cores = MAX_CORES
 )
-
-# HMC and parallel computation options
-MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"))
-if (is.na(MAX_CORES)) MAX_CORES <- parallelly::availableCores(logical = FALSE)
-if (require(cmdstanr)) {
-  options(
-    brms.backend = "cmdstanr", 
-    brms.threads = MAX_CORES %/% 4
-  )
-}
 
@@
 )
-```
 
-```{r data-download}
-# Download data if not downloaded
-path <- "data/beyens-et-al-2021.csv"
-dir.create(dirname(path), FALSE)
-if (!file.exists(path)) 
-  download.file("https://uvaauas.figshare.com/ndownloader/files/28057809", path)
-
-# Load and clean
-d <- read_csv(path) %>%
-  # Subset data to a handful of participants' evening observations
-  filter(
-    BeepTE1 == 6,
-    ID %in% sample(.$ID, 30)
-  ) %>% 
-  mutate(
-    person = factor(ID),
-    time = BeepNE1,
-    # Sum of all passive use in the past hour
-    sm = rowSums(select(., INP01E1, INP02E1, WAP01E1, SNP01E1, SNP02E1), na.rm = TRUE),
-    # It can exceed 60 if p reported multiple overlapping uses
-    sm = if_else(sm > 60, 60, sm),
-    happy = AWBE1,
-    .keep = "none"
-  ) %>% 
-  # Drop missing data
-  drop_na(sm, happy)
-
-# Lagged and centered variables
-d <- d %>% 
-  # Lag happiness by person
-  mutate(
-    happy1 = lag(happy), 
-    .by = person
-  ) %>% 
-  # Grand mean center so between-person component is centered on zero
-  mutate(
-    smg = sm - mean(sm, na.rm = TRUE)
-  ) %>% 
-  # Calculate person means and person-specific deviations therein
-  mutate(
-    smb = mean(smg, na.rm = TRUE), 
-    smw = smg - smb,
-    happy1w = happy1 - mean(happy1, na.rm = TRUE),
-    .by = person
+# Document options
+knitr::opts_chunk$set(
+  echo = FALSE
 )
@@
 
-In longitudinal and other repeated measures data where predictors are observed associations can emerge at two (or more) levels: Between individuals and within individuals [@bolgerIntensiveLongitudinalMethods2013].
-Because these two levels of analysis can embody different theoretical constructs, researchers have widely recognized the importance of centering the data appropriately, such that within-person associations can be estimated independently from between-person associations [@endersCenteringPredictorVariables2007].
+Longitudinal and other repeated measures data allow examining associations at two (or more) levels: Between and within individuals [@bolgerIntensiveLongitudinalMethods2013].
+Because these separate levels of analysis can indicate different theoretical constructs, researchers have widely recognized the importance of centering the data appropriately, such that within-person associations can be estimated independently from between-person associations [@endersCenteringPredictorVariables2007].
 If this distinction between sources of variance is not made, the resulting estimates are likely to be confusing mixtures of associations that occur between individuals and within individuals [@cronbachResearchClassroomsSchools1976 p.18; @raudenbushHierarchicalLinearModels2002 p.138].
@@
 In typical applications, a predictor is separated into two components: First, each individual's mean is calculated.
-Including the person mean as predictor in a regression model will describe the extent to which individuals with different average levels of the predictor differ on the outcome variable.
 Second, the individuals' means are subtracted from each observation, leading to a variable that indicates deviations around each person's average.
 This latter variable represents pure within-person variability.
-Including it as a model predictor leads to estimating how the outcome differs at different levels of the predictor, for the average person.
-Typically, it is this latter within-person association that is of key theoretical interest in psychology: It can be used to describe a counterfactual causal scenario where increasing the predictor for an individual leads to changes in that person's outcome [@rohrerTheseAreNot2021].
+Including it as a predictor leads to estimating how the outcome differs at different levels of the predictor, for the average person.
+Typically, it is this latter within-person association that is of key theoretical interest in psychology: It can be used to describe a counterfactual causal scenario where increasing the predictor for an individual leads to changes in that person's outcome [@rohrer2023].
+Including the person mean as predictor in a regression model will describe the extent to which individuals with different average levels of the predictor differ on the outcome variable.
 
-While simple to implement, this data manipulation trick has one potentially severe shortcoming: It can lead to biases in the estimated regression coefficients, especially when the number of observations per person is small.
+While simple to implement, this data manipulation trick has one shortcoming: It can lead to biases in the estimated regression coefficients, especially when the number of observations per person is small.
 These biases result from ignoring the uncertainty inherent in the person means: They can only be estimated based on a sample of limited data, but simple calculations cannot carry this uncertainty forward into the resulting estimates.
 While this issue is recognized in the methodological literature [e.g.
-@asparouhovLatentVariableCentering2019; @mcneishPrimerTwolevelDynamic2020], practical implementations to overcome it have so far been limited to commercial solutions, most prominently the SEM software MPlus.
-In this manuscript, we provide a practical tutorial on latent mean centering in the context of longitudinal and repeated measures data in psychology.
-We implement latent mean centering in a series of multilevel models using the brms R package.
-Our aim is to provide this illustrative code so that latent mean centering can be more widely adopted within the psychological sciences.
+@asparouhovLatentVariableCentering2019; @mcneishPrimerTwolevelDynamic2020], the practical solution to overcome it---latent mean centering---has so far been limited to commercial solutions, most prominently the structural equation modelling software Mplus.
 
-In appendix A(?), we also conduct a simulation study to illustrate the conditions under which latent mean centering is most useful: As the number of observations per participant decreases, the beneficial debiasing effect of latent mean centering becomes more noticeable and important.
-In addition, while previous work has focused on latent mean centering specifically in the context of longitudinal datasets and timeseries analyses, we point out that it is a more general issue and applicable to datasets with potentially multiple levels of analysis?
-In appendix B, we extend the syntax to a three level model where latent means are estimated for adolescents nested within classes?
+Our aim in this manuscript is to illustrate how latent mean centering can be implemented in the free and open source R language using the brms package [@rcoreteamLanguageEnvironmentStatistical2023; @bürkner2017].
 
-# Latent mean centering tutorial
+# A two-level model with latent mean centering
 
-In this example analysis, we examine a longitudinal dataset of Dutch adolescents' social media use and well-being.
-The authors of the original research article have made their valuable data openly available, which we appreciate.
-The sample consists of 387 Dutch adolescents (mean age: 14, 54% girls), who answered 34,930 experience sampling questions about their well-being ("How happy do you feel right now?"; 1: "Not at all", 4: "A little", 7: "Completely") and social media use.
-For this analysis, we used the sum of passive Instagram, WhatsApp, and Snapchat reported in the past hour.
-We show a snippet of this data in Table \@ref(tab:tab-data).
+In this example analysis, we examine a longitudinal dataset of Dutch adolescents' social media use and well-being, using the authors' openly available data.
+In the 7-day experience sampling study, 63 adolescents (54% girls) were surveyed six times per day, on their mobile phones, about their well-being and social media use.
+During each approximately 2-minute report, participants reported, for variour social media apps, how many minutes in the past hour they used each.
+They answered this question separately for active and passive use.
+They also reported their well-being on a seven-point scale ("How happy do you feel right now?"; 1: "Not at all", 4: "A little", 7: "Completely").
+
+With these data, the authors found little relation between active or passive social media use and well-being, but that the associations varied strongly between adolescents.
+These data are currently among some of the most valuable for addressing associations between adolescents technology use and well-being, and here we use them to illustrate a latent means model in R examining that issue.
 
 ```{r}
-#| label: tab-data
+#| label: tbl-data-1
+#| tbl-cap: Example data.
 
-d %>% 
-  slice(1:3, .by = person) %>% 
-  head() %>% 
-  apa_table(
-    caption = "Six rows of example data", 
-    note = "sm = passive social media use; smg = grand-mean centered sm; smb = between-person sm; smw = within-person sm; happyw1 = within-person lagged happiness rating.",
-    col_spanners = list(`Original data` = c(1, 5), `Centered values` = c(6, 9)),
-    digits = c(0, 0, 0, 0, 0, 0, 1, 1, 1), 
-    font_size = "small", 
-    span_text_columns = FALSE
+dat <- read_rds("data/beyens-etal.rds")
+
+dat |> 
+  slice(1:3, .by = person) |> 
+  head() |> 
+  kable(digits = 2)
+```
+
+Table @tbl-data-1 shows the three first observations of the first two participants in the sample.
+The first three columns contain a classroom identifier, an anomyous person identifier, and the observation time point (an increasing integer).
+`sm` is the number of hours of passive social media use in the past hour, and `wb` (for well-being) the happiness rating.
+(We rescaled these two variables from the original minutes and 1-7 scale to the unit scale.) `wb_lag` is the time-1 lagged version of `wb`.
+
+Our research question with these data is: To what extent does an individual's passive social media use in the past hour predict that individual's current well-being when their immediately preceding well-being is adjusted for?
+That is, we are specifically interested in the within-person correspondence between social media use and well-being.
+
+## Centering
+
+The issue of isolating between- and within-cluster associations in repeated measures data examined with multilevel models is widely appreciated.
+The common way to address this issue is to center variables appropriately, such that two new variables are created: One reflecting only between-person variance, and one within-person variance.
+The variable `sm_c` in @tbl-data-1 is the within-person centered version of social media use; or `sm` from which each person's mean, `sm_b` is subtracted.
+
+In typical multilevel analyses, including `sm_c` as a level-1 predictor, and `sm_b` as a level-2 predictor leads to two population level parameters that index the extent to which social media use predicts well-being within- and between-people, respectively.
+
+However, using these calculated variables can come with the cost of biased parameter estimates, because this method relies on using `sm_b` as known data when, in reality, it is an unknown quantity estimate from data with its own estimation uncertainty.
+
+Latent mean centering is a method by which the predictor variables are centered within the model such that the estimation uncertainty is incorporated in the calculations.
+
+## Model
+
+We loosely follow the authors' original analysis, which focused on the associations between social media use and well-being.
+In this analysis, we are interested in the within- and between-person associations between passive social media use in the past hour and current well-being, while adjusting for well-being at the previous timepoint.
+Because the data consist of repeated measures of multiple individuals, we specify a multilevel model where the regression coefficients are allowed to vary randomly between persons.
+These models for intensive longitudinal data (ILD) where autocorrelative terms are modelled as random are also sometimes called dynamic structural equation models.
+
+More formally, for each individual i in 1 to 63, and timepoint t in 1 to 42, we model the happiness rating as normally distributed with a common residual standard deviation.
+We model the mean on an intercept, a regression coefficient of the previous happiness rating, and a coefficient of the current passive social media use variable.
+We have superscripted the latter two predictor variables with a c, indicating that they are within-person centered.
+The next two lines define this within-person centering, where the centered value is the raw value minus the estimated person-mean of that variable.
+
+$$
+\begin{align}
+H_{it} &\sim N(\alpha_i + \phi_i H^c_{it-1} + \beta_i SM^c_{it}, \sigma^2_1), \\
+H^{c}_{it} &= H_{it} - \alpha_{1i}, \\
+SM_{it} &\sim N(\alpha_{2i}, \sigma^2_2), \\
+SM^{c}_{it} &= SM_{it} - \alpha_{2i}.
+\end{align}
+$$ {#eq-1.1}
+
+Above, we have subscripted all person-specific parameters with i; the means of H and SM (\alpha_1 and \alpha_2, respectively), the autocorrelation in H (\phi_i), and the coefficient of within-person centered SM on H (\beta_i).
+To complete this model, we specify that all person-specific parameters are multivariate normal distributed.
+
+$$
+\begin{align}
+\begin{bmatrix}
+  \alpha_{1i} \\ \alpha_{2i} \\ \phi_i \\ \beta_i
+\end{bmatrix} &\sim MVN\left(
+  \begin{bmatrix}
+    \bar{\alpha_1} \\ \bar{\alpha_2} \\ \bar{\phi} \\ \bar{\beta}
+  \end{bmatrix},
+  \begin{pmatrix}
+    \tau_{\alpha_1} \  \\ 
+    \rho_{\alpha_1\alpha_2} \ &\tau_{\alpha_2} \\ 
+    \rho_{\alpha_1\beta} \ &\rho_{\alpha_2\beta} \ &\tau_\beta \\
+    \rho_{\alpha_1\phi} \ &\rho_{\alpha_2\phi} \ &\rho_{\beta\phi} \ &\tau_\phi
+  \end{pmatrix}
+\right)
+\end{align}
+$$ {#eq-1.2}
+
+This model is difficult to specify in common multilevel modelling software, because it shares parameters across regression models of different outcomes.
+Specifically, in @eq-1.1 we create the within-person centered variable $SM^c_{it}$ "dynamically" within the model by subtracting the estimated mean of SM for person i from the current observation.
+In other words, @eq-1.1 is a multivariate model where some parameters are shared across the outcomes.
+However, there is a somewhat simple solution to this issue: We can "stack" the multivariate data two a longer univariate data table.
+@tbl-data-2 shows the result of this data stacking: Instead of separate columns for SM and H, there are now two indicator columns `i_wb` and `i_sm` that are 1 when that outcome is in the `value` column and 0 otherwise.
+
+```{r}
+#| label: tbl-data-2
+#| tbl-cap: Example data 2.
+
+dat <- dat |> 
+  pivot_longer(sm:wb, names_to = "outcome") |> 
+  mutate(
+    i_wb = if_else(outcome == "wb", 1, 0),
+    i_sm = if_else(outcome == "sm", 1, 0), 
+    .before = value
+  ) |> 
+  left_join(dat)
+
+dat |> 
+  slice(1:3, .by = person) |> 
+  head() |> 
+  kable(digits = 2)
+```
+
+In practice, we can then use this stacked data and reparameterize our model as a univariate model:
+
+$$
+\begin{align}
+Y_{it} &\sim N(\eta_{it}, \sigma_{it}^2), \\
+\eta_{it} &= 
+  I_{WBit}(\alpha_{1i} + 
+    \phi_i(H_{it-1} - \alpha_{1i}) + 
+    \beta_i(SM_{it} - \alpha_{2i})) + 
+  I_{SMit}(\alpha_{2i}), \\
+\sigma_{it} &= \text{exp}(I_{WBit}\omega_1 + I_{SMit}\omega_2).
+\end{align}
+$$
+
+This model then estimates the latent mean of H (\alpha_1) and SM (\alpha_2), the autoregressive coefficient of H (\phi) and coefficient of SM on H (\beta).
+Importantly, the last line also models separate residual standard deviations for the two outcomes by specifying a linear model of the two outcome indicator variables, and then passing the linear predictor through a log-link function.
+
+Using the data in @tbl-data-2, we can now specify this model using brms' model syntax:
+
+```{r}
+model <- bf(
+  value ~ 
+    i_wb * (alpha1 + phi * (wb_lag - alpha1) + beta * (sm - alpha2)) + 
+    i_sm * alpha2,
+  nlf(sigma ~ i_wb * omega1 + i_sm * omega2),
+  alpha1 + alpha2 + phi + beta ~ 1 + (1 |p| person),
+  sigma1 + sigma2 ~ 1,
+  nl = TRUE
 )
@@
 
-Table \@ref(tab:tab-data) shows the three first observations of two randomly selected participants from the sample.
-The first four columns contain an anomyous person identifier, the observation time point (an increasing integer), the number of minutes of passive social media use in the past hour, and the happiness rating.
-Then, for illustration, we have created three additional columns containing the centered social media use variables.
-The first, "smg", is the simple grand-mean centered "sm" variable.
-This variable is not required in all analyses.
-However, it can be used to zero-center the between-person variable if it is of interest.
-"smb" is the between-person component, which is simply that person's sample mean.
-"smw" is the deviation of the social media use value on that row from the person's mean, and thus the within-person component.
+```{r}
+fit <- brm(
+  bform,
+  data = d,
+  iter = ITER,
+  control = list(adapt_delta = 0.95),
+  file = "cache/brm-example-sm-1"
+)
+```
 
-We loosely follow the authors' original analysis, which focused on the associations between social media use and happiness.
-To ground our tutorial on common practices, we first ignore latent mean centering, and describe the analysis in terms of the observed values.
-In this analysis, we are interested in the within- and between-person associations between passive social media use in the past hour and current well-being.
-To do so, we define a multilevel model of happiness ratings on both "smw" and "smb", and allow for random variability in person-specific averages in happiness ratings, and the association between happiness ratings and "smw".
+With these data, model estimation took about one minute on a modern laptop computer. We show the estimated population-level parameters in @tbl-model-1.
 
-**Throw in a line here about why the autoregressive term is included.**
+```{r}
+#| label: tbl-model-1
+#| tbl-cap: Model results.
 
-In typical R formula syntax, generalized for multilevel models in the lme4 package, this model is
+as_draws_df(fit, variable = c("b_", "sd_", "cor_"), regex = TRUE) %>% 
+  rename_with(~str_remove_all(., "_Intercept")) |> 
+  mutate(
+    across(starts_with("b_sigma"), ~exp(.))
+  ) |> 
+  summarise_draws(
+    mean, sd, ~quantile2(.x, c(.025, .975))
+  ) |> 
+  kable(digits = 2)
+```
 
-which translates to the mixed effects model in Box 1.
-This model returns three key population-level estimates---parameter values assumed to hold for the average person in the population from which the sample was drawn: An intercept, slope of "smb", and slope of "smw".
-The first describes the happiness level of an average individual who passively used an average amount of social media in the past hour, and
+# A three-level model with latent mean centering
 
@@
 
-R [@rcoreteamLanguageEnvironmentStatistical2022]
+## Advantages
+
+Using brms for latent means models enables the use of a wide variety of methods, including nonlinear model terms, smooth terms (generalized additive mixed models), mixture models, and more.
+
+## Limitations
+
+Here, we outlined the use of the general-purpose regression modelling R package brms for estimating latent means models.
+This approach has two main limitations.
+First, by requiring us to stack the data we cannot model different variables using different distributions.
+In our examples, we assumed that all outcomes were normally distributed.
+This practical limitation can be problematic in cases, for example, when one of the outcomes is binary and the other one is a continuous variable.
+Second, although a wide variety of e.g.
+cross-lagged location-scale within-between latent means models can be fitted with brms, the commercial alternative Mplus is currently more flexible.
+For example, users of Mplus can specify residual DSEMs which cannot be fitted with brms.
+
+Moreover, data stacking increases the practical size of the dataset and thus can be slower to estimate than a true multivariate model, such as ones that can be fitted with Mplus.
+
+Missing data.
 
@@
 
-Conceptualization: MV  
-Methodology: MV  
-Software: MV  
-Validation: MV  
-Formal Analysis: MV  
-Data Curation: MV  
-Writing---Original Draft: MV  
-Writing---Review & Editing: MV  
-Visualization: MV  
+Conceptualization: MV\
+Methodology: MV\
+Software: MV\
+Validation: MV\
+Formal Analysis: MV\
+Data Curation: MV\
+Writing---Original Draft: MV\
+Writing---Review & Editing: MV\
+Visualization: MV\
 Project Administration: MV

### 2024-01-25 simulation.qmd
Organize materials better (?)

@@
 
-N_CORES <- min(parallel::detectCores(), 8)
+# HMC and parallel computation options
+MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"), 4)
 options(
-  mc.cores = N_CORES,
-  brms.backend = if(require(cmdstanr)) "cmdstanr",
-  brms.threads = N_CORES %/% 4
+  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
+  brms.threads = MAX_CORES %/% 4,
+  mc.cores = MAX_CORES
 )

### 2024-01-25 README.qmd
Organize materials better (?)
Update README smore

@@
 
+# Modify environment variables in .Renviron.example & rename it to .Renviron
+
 # Run code and render outputs
@@
 
+Once the computations are finished the rendered documents are in docs/.
+
 ## Contribute
@@
 Contributions preferably via pull requests.
+You can also grab the rendered manuscript (Word, PDF) from the latest [release](https://github.com/mvuorre/latent-mean-centering-ms/releases) for comments & edits.

### 2024-02-19 README.qmd
Update readme

@@
 
-Repository for materials supporting our manuscript discussing practical implementations of latent mean centering in multilevel models of longitudinal psychological data.
+Repository for our manuscript discussing an [R](https://cran.r-project.org/) and [brms](https://cran.r-project.org/web/packages/brms/) implementation of latent mean centering in multilevel models of longitudinal psychological data.
 
@@
 - Online: tbd
+- preprint: tbd
 
@@
 
-Once the computations are finished the rendered documents are in docs/.
+Once the computations are finished (this will take a long time depending on your computer) the rendered documents are in `docs/`.
 

### 2024-02-19 data.qmd
Cite beyens

@@
 
+library(haven)
 library(tidyverse)
@@
 
-Throughout, we use an example dataset from "A primer on two-level dynamic structural equation models for intensive longitudinal data in Mplus" [@mcneishPrimerTwolevelDynamic2020].
+We will use two example datasets.
+The first is from "A primer on two-level dynamic structural equation models for intensive longitudinal data in Mplus" [@mcneishPrimerTwolevelDynamic2020].
 Those data are available on the OSF (<https://osf.io/wuprx/>).
@@
 
+Our second dataset, analyzed in the manuscript, is from @beyens2020. Below, we download & wrangle then save it for later analyses.
+
 ```{r data-process}
@@
     person = factor(EthicaID),
-    # I think they divided this by ten
     time = (OccNrR - 1),
-    # active = rowSums(select(., contains("Act")), na.rm = TRUE),
-    sm = rowSums(select(., contains("Pas")), na.rm = TRUE) / 60,
-    wb = as.numeric(AffWB) / 7, 
+    sm = rowSums(select(., contains("Pas"), contains("Pas")), na.rm = TRUE) / 60,
+    happy = as.numeric(AffWB) / 7, 
     .keep = "none"
@@
 
-# Remove times in excess of 60min (1 hour)
+# Sum scores exceeding 60minutes (i.e., less than 2.5% of all occasions per subscale) were recoded to 60 minutes. (Beyens et al. p.9)
+d$sm <- if_else(d$sm > 1.0, 1.0, d$sm)
+
+# Lag well-being
 d <- d %>% 
   mutate(
-    across(c(sm), ~if_else(. > 1, 1, .))
-  )
-
-# Centered and lagged variables
-d <- d %>% 
-  mutate(
-    sm_b = mean(sm, na.rm = TRUE),
-    sm_c = sm - mean(sm, na.rm = TRUE),
-    across(
-      c(wb),
-      .fns = list(lag = ~lag(.))
-    ),
+    happy_lag = lag(happy),
     .by = person
@@
 
-d
-
 write_rds(d, "data/beyens-etal.rds")

### 2024-02-19 examples.qmd
Take out analysis in MS

@@
 
-## Social media use
-
-```{r}
-d <- read_rds("data/beyens-etal.rds")
-
-# Rescale everything to the unit interval
-d <- d |> 
-  mutate(
-    across(c(sm), ~./60),
-    across(starts_with("wb"), ~./7)
-    )
-
-d <- d |> 
-  pivot_longer(c(sm, wb), names_to = "outcome") |> 
-  mutate(
-    i_wb = if_else(outcome == "wb", 1, 0),
-    i_sm = if_else(outcome == "sm", 1, 0)
-  ) |> 
-  # Include predictors from each row
-  left_join(d)
-```
-
-We regress wellbeing on an autocorrelative term and within person effect of passive social media use.
-Include correlations between person parameters; one of these will be the between-person relationship.
-
-```{r}
-bform <- bf(
-  value ~ 
-    i_wb * (alpha1 + phi * (wb_lag - alpha1) + beta * (sm - alpha2)) + 
-    i_sm * alpha2,
-  nlf(sigma ~ i_wb * sigma1 + i_sm * sigma2),
-  alpha1 + alpha2 + phi + beta ~ 1 + (1 |p| person),
-  sigma1 + sigma2 ~ 1,
-  nl = TRUE
-)
-```
-
-```{r}
-fit <- brm(
-  bform,
-  data = d,
-  control = list(adapt_delta = 0.95),
-  file = "cache/brm-example-sm-1"
-)
-```
-
-```{r}
-as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
-  mutate(
-    across(starts_with("sd_"), ~.^2),
-    across(starts_with("b_sigma"), ~exp(.)^2)
-  ) |> 
-  summarise_draws(
-    mean, sd, ~quantile2(.x, c(.025, .975))
-  )
-  gt2()
-```
-
-## Social media use 2
-
-Same as above but a three-level model class > person > observation.
-
-```{r}
-bform <- bf(
-  value ~ 
-    i_wb * (alpha1 + phi * (wb_lag - alpha1) + beta * (sm - alpha2)) + 
-    i_sm * alpha2,
-  nlf(sigma ~ i_wb * sigma1 + i_sm * sigma2),
-  alpha1 + alpha2 + phi + beta ~ 1 + (1 |p| person) + (1 |c| class),
-  sigma1 + sigma2 ~ 1,
-  nl = TRUE
-)
-```
-
-```{r}
-fit <- brm(
-  bform,
-  data = d,
-  control = list(adapt_delta = 0.95),
-  file = "cache/brm-example-sm-threelevel"
-)
-```
-
-```{r}
-as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
-  mutate(
-    across(starts_with("sd_"), ~.^2),
-    across(starts_with("b_sigma"), ~exp(.)^2)
-  ) |> 
-  summarise_draws(
-    mean, sd, ~quantile2(.x, c(.025, .975))
-  )
-  gt2()
-```
-
 ## Conclusion

### 2024-02-19 mplus-parameters.R
Explain what script does

@@ -1,3 +1,5 @@
+# Read and wrangle MH's Mplus model parameters for comparison
+
 library(MplusAutomation)
 
 mplus <- readModels("data/mh/Output Files")

### 2024-06-26 index.qmd
Infra update

@@
 ---
+title: Estimating dynamic structural equation models in R with the brms package
+running-head: DSEM with R & brms
 author: 
   - name: Matti Vuorre
-    url: https://vuorre.com
-    orcid: 0000-0001-5052-066X
-    email: mjvuorre@uvt.nl
     affiliation: 
       - ref: 1
+    corresponding: true
+    email: mjvuorre@uvt.nl
+    orcid: 0000-0001-5052-066X
+  - name: Mauricio Garnier-Villarreal
+    affiliation: 
+      - ref: 2
   - name: Ethan McCormick
@@
       - ref: 3
-  - name: Mauricio Garnier-Villarreal
+  - name: Joran Jongerling
     affiliation: 
       - ref: 4
-  - name: Joran Jongerling
-    orcid: 0000-0001-5697-1381
-    affiliation:
-      - ref: 2
 affiliations:
@@
     name: Tilburg University
+    department: Department of Social Psychology
   - id: 2
+    name: Vrije Universiteit Amsterdam
+  - id: 3
+    name: Leiden University
+  - id: 4
     name: Tilburg University
-  - id: 3
-    name: todo
-  - id: 4
-    name: todo
+    department: Methods and Statistics
+abstract: |
+  Dynamic Structural Equation Models (DSEM) combine latent variables, time-series (lagged) relations, and hierarchical structures for flexible analyses of intensive longitudinal data, such as collected in experience sampling and other longitudinal designs [@asparouhov2018].
+However, the general availability of DSEM is currently limited to commercial and closed source software packages.
+Here, we show how to specify flexible DSEMs with the popular free and open source brms R package.
+Using example data on adolescent well-being from the published literature [@beyens2020], we demonstrate two- and three-level location-scale models with lagged relations, time varying and time-invariant covariates, and how to post-process the models for effective communication of results.
+In an online supplement, we confirm the validity of our approach with a parameter recovery simulation, and compare our results using example analyses from @mcneish2020.
+DSEMs provide an exciting and robust framework for analyses of intensive longitudinal data and we hope our illustration of estimating them with free and open source software packages spurs advances in the area.
+keywords: [R, brms, statistics, latent variable, dynamic structural equation model]
+authornote: |
+  This manuscript is not yet peer reviewed.
+bibliography: bibliography.bib
+citeproc: true
+csl: apa.csl
 ---
 
-{{< include README.qmd >}}
+```{r}
+#| label: setup
+#| include: false
+
+# Packages
+library(latex2exp)
+library(tidybayes)
+library(modelr)
+library(ellipse)
+library(scales)
+library(knitr)
+library(posterior)
+library(patchwork)
+library(ggdist)
+library(cmdstanr)
+library(brms)
+library(tidyverse)
+
+# Output options
+knitr::opts_chunk$set(
+  eval = TRUE,
+  cache = TRUE,
+  warning = FALSE,
+  error = FALSE,
+  message = FALSE,
+  echo = FALSE
+)
+options(digits = 2)
+
+# Plotting options
+theme_set(
+  theme_classic(base_size = 9) +
+    theme(
+      strip.text = element_text(color = "black", hjust = 0),
+      strip.background = element_rect(color = NA, fill = NA),
+      strip.text.x = element_text(size = rel(0.9)),
+      line = element_line(linewidth = .25),
+      plot.tag = element_text(size = rel(1))
+    )
+)
+
+# Model estimation options, customize in .Renviron
+dir.create("cache", FALSE)
+options(
+  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
+  brms.threads = as.numeric(Sys.getenv("MAX_CORES"), 1),
+  mc.cores = as.numeric(Sys.getenv("MAX_CORES"), 4)
+)
+```
+
+```{r}
+#| label: functions
+
+# Neat posterior summaries
+sm <- function(
+    x, 
+    variables = c("b_", "sd_", "cor_", "sigma")
+) {
+  x |> 
+    as_draws_df(variable = variables, regex = TRUE) |> 
+    mutate(
+      across(any_of(c("k0_Intercept", "k1_Intercept")), ~exp(.x))
+    ) |> 
+    summarise_draws(
+      mean, sd,
+      ~quantile2(.x, c(.025, .975))
+    ) |> 
+    mutate(
+      variable = str_remove_all(variable, "_Intercept")
+    )
+}
+```
+
+```{r}
+#| label: data-beyens-etal
+
+# The data is cited in the paper as "The dataset generated and analysed during the current study is available in Figshare." (https://uvaauas.figshare.com/articles/dataset/Dataset_belonging_to_Beyens_et_al_2020_The_effect_of_social_media_on_well-being_differs_from_adolescent_to_adolescent/12497990)
+
+# Download data if not yet downloaded
+path_in <- "data/beyens-et-al-2020.sav"
+if (!file.exists(path_in)) {
+  dir.create("data", FALSE)
+  # Data from figshare
+  download.file(
+    "https://uvaauas.figshare.com/ndownloader/files/24028271", 
+    path_in
+  )
+  # Code from OSF
+  download.file(
+    "https://files.de-1.osf.io/v1/resources/bvfuw/providers/osfstorage/5df9effa64e19d000d0f51da/?zip=",
+    "data/Analyses.zip"
+  )
+  unzip(
+    "data/Analyses.zip",
+    exdir = "data/Analyses"
+  )
+}
+
+path_out <- "data/beyens-etal.rds"
+if (!file.exists(path_out)) {
+
+  library(haven)
+
+  # Active, Passive, and Total were sums of the platform (e.g. whatsapp) times
+  d <- read_spss(path_in) %>% 
+    mutate(
+      class = factor(ClassNr),
+      person = factor(EthicaID),
+      time = as.integer(OccNrR - 1),
+      # Rescale to 0-1 interval
+      sm = rowSums(
+        select(., contains("Pas"), contains("Pas")), 
+        na.rm = TRUE
+      ) / 60,
+      happy = (as.numeric(AffWB)-1) / 6, 
+      .keep = "none"
+    ) %>% 
+    arrange(person, class, time)
+
+  # Sum scores exceeding 60minutes (i.e., less than 2.5% of all occasions per subscale) were recoded to 60 minutes. (Beyens et al. p.9)
+  d$sm <- if_else(d$sm > 1.0, 1.0, d$sm)
+
+  # Lag well-being
+  d <- d %>% 
+    mutate(
+      hlag = lag(happy),
+      .by = person
+    )
+
+  write_rds(d, path_out)
+}
+
+dat <- read_rds("data/beyens-etal.rds")
+```
+
+# Introduction
+
+The dynamics of psychological attributes over time are central to psychological research.
+To examine, longitudinal datasets where individuals are repeatedly assessed over time have typically been analysed under one of two broad statistical frameworks: Multilevel and structural equation models (SEM) [@bolger2013].
+Dynamic structural equation models (DSEM) [@asparouhov2018] bring features of those two frameworks together and allow analysts to examine longitudinal data for dynamic relationships (relationships over time), latent variables, and more.
+However, accessible software implementations of DSEM models in psychology are currently limited to Mplus [@muthén2017] and Stan [@standevelopmentteam2023]—the former being a commercial closed-source package, and the latter perceived to have a prohibitively steep learning curve for many applied psychologists.
+
+To encourge and enable a broader application and examination of DSEM, we show how to implement the main features of those models, latent variables and dynamic relations, in the free and open source R programming environment [@rcoreteam2024] using the brms package [@bürkner2017; @bürkner2018].
+The code for reproducing this manuscript is at <https://github.com/mvuorre/funky-firefly> and this document is available at [todo].
+
+# Example 1: univariate two-level AR(1) model
+
+## Dataset
+
+Throughout our example analyses, we investigate a longitudinal dataset of 63 Dutch adolescents' social media use and well-being described in and shared by @beyens2020.
+The adolescents (54% girls; 8th and 9th graders) took part in a 7-day experience sampling study in which they reported their well-being and social media use six times per day on their mobile phones.
+In each \~2-minute survey, participants reported how many minutes in the past hour they used each of several popular social media platforms (e.g.
+Instagram, WhatsApp).
+They then answered "How happy do you feel right now?" by choosing a response option from 1 ("Not at all") to 7 ("Completely").
+
+With these data, @beyens2020 found little relation between active or passive social media use and well-being for the average person.
+However, the authors found that relation to vary strongly between adolescents, and have since confirmed this finding with extensions to the data and other well-being outcomes [@beyens2021b].
+In this manuscript, we use their openly available data [@beyens2021a].
+
+```{r}
+#| label: tbl-data-1
+#| tbl-cap: Example data.
+
+dat |> 
+  slice(1:3, .by = person) |> 
+  head() |> 
+  kable()
+```
+
+We show the first three measurement occasions for two random participants in Table @tbl-data-1).
+The first three columns are, respectively, a classroom identifier, an anonymous person identifier, and the observation index for that person.
+For our example analyses, we collapsed across active and passive social media use; consequently `sm` is the reported hours of total social media use in the past hour.
+Our key outcome variable is `h`, the happiness rating (rescaled to 0-1 interval).
+We also include `hlag`, the time-1 lagged version of `h`.
+Note that it is missing on each participant's first observation.
+
+In our first example analysis, we examine these data for potential autocorrelative trends in happiness.
+To do so, we build a minimal two-level DSEM model that estimates an autocorrelative term of happiness that varies randomly between individuals and a latent mean term that is used to within-person-center the lagged happiness ratings.
+
+## Centering
+
+The issue of isolating between- and within-cluster variance in repeated measures data is widely appreciated [@asparouhovLatentVariableCentering2019; @asparouhov2018; @bolger2013; @enders2007; @mcneish2020], and is a key topic of our write-up.
+Typically, researchers isolate within- and between-person variance by entering two predictors to their regressions instead of one.
+First, using the person-means as predictors allows examining how people differ from one another in their average responses.
+Second, researchers subtract the person-means from each predictor value, and use the resulting within-person-centered variable to examine within-person associations.
+That is, within- and between-person centering is an easy but important data-manipulation step.
+
+However, using these calculated variables can lead to biased parameter estimates especially when participants provide only a small number of observations [@mcneish2020].
+Specifically, using the means calculated from data to center variables leads to Nickell’s (negative bias in autoregressive effects) [@nickell1981] and Lüdtke’s (bias in other time-varying effects) [@lüdtke2008] biases.
+To avoid these biases, analysts can instead *latent*-mean-center their variables.
+In contrast to observed mean centering, latent mean centering is performed as part of the model, such that uncertainty in the estimated means is incorporated to the resulting model estimates.
+We now turn to our first example model that illustrates obtaining latent means and using them to center within-person variables inside a multilevel model.
+
+## Model
+
+This model examines the extent of first-degree autocorrelation among happiness responses, and how it varies across individuals.
+We write this univariate two-level AR(1) model of happiness ($H$) for the $i^{th}$ individual at their $t^{th}$ measurement occasion as
+
+$$
+\begin{align*}
+H_{it} &\sim N(\beta_{0i} + \beta_{1i}(H_{it-1} - \beta_{0i}), \sigma^2), \\
+\beta_{0i} &= \bar{\beta}_0 + u_{0i}, \\
+\beta_{1i} &= \bar{\beta}_1 + u_{1i}, \\
+\begin{bmatrix}
+  u_{0i} \\ u_{1i}
+\end{bmatrix} &\sim MVN\left(
+  \begin{bmatrix}
+    0 \\ 0
+  \end{bmatrix},
+  \begin{pmatrix}
+    \tau_0 \ & \\ 
+    \rho_{01} \ &\tau_1
+  \end{pmatrix}
+\right).
+\tag{1}
+\end{align*}
+$$
+
+In words, we assume the happiness ratings to be normally distributed around person-specific intercepts ($\beta_{0i}$) and time-1 autocorrelative coefficients ($\beta_{1i}$), which are multivariate normal distributed around means $\bar{\beta}_0$ and standard deviations $\tau$, with $\rho_{01}$ indicating the degree to which individuals' intercepts and slopes are correlated.
+A critical term in Equation 1 is $\beta_{1i}(H_{it-1} - \beta_{0i})$.
+It both specifies the autocorrelative coefficient $\beta_{1i}$ and creates its predictor, the latent-mean-centered time-1 lagged happiness rating for person $i$ by subtracting the person's estimated (latent) happiness mean $\beta_{0i}$ from their time-1 lagged happiness response.
+
+The brms syntax for estimating Model 1 is shown below, where we save the model formulas to an object named `m1`.
+We specify the syntax inside the `bf()` (short for brmsformula) function so that it is appropriately interpreted.
+
+```{r}
+#| label: model-1-specify
+#| echo: true
+
+m1 <- bf(
+  happy ~ b0 + b1 * (hlag - b0),
+  c(b0, b1) ~ 1 + (1 |p| person),
+  nl = TRUE
+)
+```
+
+In the first argument to `bf()`, we specified an outcome (`happy`) on the left-hand side (LHS) of the tilde, and on the RHS a (potentially nonlinear) combination of the predictor variables and parameters to be estimated.
+This level-1 regression equation is for the location parameter of the chosen outcome distribution---we used the default gaussian distribution and will do so throughout.
+All names in this equation refer either to variables in the data (`happy`, `hlag`), or parameters to be estimated (`b0`, `b1`).
+To distinguish, the former must exist in the data.
+
+On the following line we specified the level-2 linear equations of the level-1 equation's parameters.
+We used `c(b0, b1)` to specify that the RHS applies to both parameters.
+`1` specifies a global intercept, and `(1 |p| person)` specifies person-specific deviations from that global intercept.
+Because there are multiple parameters with person-specific deviations, we used `|p|` to ensure that they will be estimated within a common multivariate normal distribution whose correlations are also estimated.
+Finally, `nl = TRUE` on the last line is required to be able to name parameters on the first line.
+
+To estimate this model, we pass the above model formula object to `brm()`:
+
+```{r}
+#| label: model-1-sample
+#| echo: true
+
+fit1 <- brm(
+  formula = m1,
+  data = dat,
+  file = "cache/brm-1"
+)
+```
+
+In addition to the model formula in `m1`, above we specified two more arguments to `brm()`: Name of the data in R (we have loaded the data illustrated in Table @tbl-data-1) to an R object called `dat`), and the file in which the estimated model should be saved.
+The latter is useful because bayesian estimation can be time-consuming (this example took less than a minute; we created the `cache/` directory before).
+After the model has been estimated, calling `summary(fit1)` will print a numerical summary of the estimated model in the R console.
+We show that printout in Appendix 1, and here focus on posterior summaries of the key parameter estimates, shown in Table @tbl-fit-1).
+
+```{r}
+#| label: tbl-fit-1
+#| tbl-cap: Summaries of Model 1 key parameters' posterior distributions.
+
+fit1 |> 
+  sm() |> 
+  mutate(
+    variable = str_remove_all(variable, "person__")
+  ) |> 
+  kable()
+```
+
+The first row of Table @tbl-fit-1), `b_b0` describes the happiness intercept for the average person.
+In Equation 1, we refer to these "average betas" with horizontal bars ($\bar{\beta}$) to indicate that they are the means around which the person-specific deviations vary.
+The posterior mean of this parameter, in the "mean" column, indicates the expectation of the average person's happiness mean for an observation with an average previous happiness.
+The following values in Table @tbl-fit-1) are the posterior standard deviation (analogous to the standard error) and 2.5 and 97.5 percentiles.
+The latter two values delimit the 95% Credibility Interval (CI), which describes the most likely values of this model parameter given the prior distributions and data.
+
+The second row of Table @tbl-fit-1) describes the posterior distribution of the happiness autocorrelation parameter for the average person.
+We can determine with 95% confidence that this parameter is positive---happiness predicts itself over time.
+The `b_` prefix to these parameters is brms' convention to indicate that they refer to regression coefficients.
+
+The rows prepended with `sd_` describe the standard deviations of the person-specific deviations $u_{0i}$ and $u_{1i}$ around their respective means, and `cor_` their correlations.
+We glean from the standard deviations that there is considerable between-person heterogeneity in both model intercepts and autocorrelations [@bolgerCausalProcessesPsychology2019].
+Their correlation is estimated with a great degree of uncertainty and we therefore cannot confidently even determine the direction of their association.
+Finally, `sigma` is the model's residual standard deviation, which we did not model on covariates.
+
+We plot this model's estimated response values (`happy`) across values of its lagged version in Figure @fig-fit-1.
+
+```{r}
+#| label: fig-fit-1
+#| fig-height: 2.6
+#| fig-cap: "Model 1 results.
+A.
+Scatterplot of person-specific intercepts ($u_0$) and autocorrelation parameters ($u_1$).
+Dark points are posterior means from Model 1; light circles are point estimates from models fitted separately to individuals' data.
+Blue point and ellipse indicates the posterior mean and 95%CI of $\\bar{\\beta}_0$ and $\\bar{\\beta}_1$.
+B.
+Implied regression line of happiness on time-1 lagged happiness for the average person.
+Points indicate data values and are jittered to reduce overplotting.
+C.
+As B.
+but for nine individuals arranged by increasing autocorrelation coefficient."
+
+# Get slopes in increasing order and select 9
+fit1_coef <- coef(fit1)$person |> 
+  as.data.frame() |>
+  rownames_to_column("person") |> 
+  tibble() |> 
+  rename_with(~str_remove_all(., "_Intercept"))
+
+nine_persons <- fit1_coef |> 
+  arrange(Estimate.b1) |> 
+  filter(((1:63) %% 7)==0) |> 
+  pull(person)
+
+epred_average <- dat |> 
+  data_grid(
+    hlag = seq_range(hlag, n = 9)
+  ) |> 
+  add_epred_draws(fit1, re_formula = NA) |> 
+  mean_qi()
+
+epred_person <- dat |> 
+  filter(person %in% nine_persons) |> 
+  group_by(person) |> 
+  data_grid(
+    hlag = seq_range(hlag, n = 9)
+  ) |> 
+  add_epred_draws(fit1) |> 
+  mean_qi()
+
+p1 <- epred_average |> 
+  ggplot(aes(hlag, .epred)) +
+  geom_ribbon(
+    aes(ymin = .lower, ymax = .upper),
+    fill = "dodgerblue2", alpha = 0.2
+  ) +
+  geom_line(col = "dodgerblue2") +
+  geom_point(
+    data = dat |> 
+      drop_na(happy),
+    aes(y = happy),
+    shape = 1, alpha = 0.5, size = 0.5,
+    position = position_jitter(.01, .01)
+  )
+
+p2 <- epred_person |>   
+  ggplot(aes(hlag, .epred)) +
+  geom_ribbon(
+    aes(ymin = .lower, ymax = .upper),
+    fill = "grey80"
+  ) +
+  geom_line(linewidth = 0.33) +
+  geom_point(
+    data = dat |> 
+      filter(person %in% nine_persons) |> 
+      drop_na(happy),
+    aes(y = happy),
+    shape = 1, alpha = 0.5, size = 0.5,
+    position = position_jitter(.01, .01)
+  ) +
+  facet_wrap("person")
+
+p1p2 <- (p1 | p2) & 
+  scale_x_continuous(
+    "Social media use",
+    breaks = c(0, 0.5, 1),
+    labels = c("0", "0.5", "1")
+  ) &
+  scale_y_continuous(
+    "Happy",
+    breaks = c(0, 0.5, 1),
+    labels = c("0", "0.5", "1")
+  ) &
+  coord_cartesian(
+    xlim = 0:1, ylim = 0:1
+  )
+
+# Intercept-slope scatterplot
+fit1_coef_lm <- dat |> 
+  group_by(person) |> 
+  group_modify(
+    ~summary(lm(happy ~ hlag, data = .x))$coefficients |> 
+      as.data.frame() |> 
+      rownames_to_column("variable") |> 
+      select(variable, estimate = Estimate, se = `Std. Error`) |> 
+      pivot_wider(
+        names_from = variable, 
+        values_from = c(estimate, se)
+      )
+  ) |> 
+  ungroup()
+
+fit1_fixef <- sm(fit1, variables = "b_") |> 
+  pivot_wider(names_from = variable, values_from = -variable)
+fit1_fixef_ellipse <- as_draws_df(fit1, variable = "b_", regex = TRUE) |> 
+  summarize_draws(mean, sd)
+
+fit1_fixef_ellipse <- ellipse(
+  cor(as.data.frame(fit1)[,1:2])[1,2],
+  scale = fit1_fixef_ellipse$sd, centre = fit1_fixef_ellipse$mean
+) |> as_tibble()
+
+p3 <- fit1_coef |> 
+  left_join(fit1_coef_lm) |> 
+  ggplot(aes(x = `estimate_(Intercept)`, y = estimate_hlag)) +
+  scale_x_continuous(
+    "Intercept",
+    expand = expansion(-0.1)
+  ) +
+  scale_y_continuous(
+    "Slope",
+    expand = expansion(-0.2)
+  ) +
+  # geom_segment(
+  #   aes(xend = Estimate.b0, yend = Estimate.b1),
+  #   linewidth = 0.1, alpha = 0.5
+  # ) +
+  geom_point(
+    aes(`estimate_(Intercept)`, estimate_hlag),
+    col = "gray80", size = 0.75
+  ) +
+  geom_point(
+    aes(x = Estimate.b0, y = Estimate.b1), 
+    col = "black", size = 1
+  ) +
+  geom_point(
+    data = fit1_fixef,
+    aes(
+      x=mean_b_b0, 
+      y=mean_b_b1
+    ), 
+    col = "dodgerblue2", size = 2
+  ) +
+  geom_path(
+    data = fit1_fixef_ellipse,
+    aes(x, y),
+    col = "dodgerblue", linewidth = 0.75
+  ) +
+  theme(aspect.ratio = 1)
+
+(p3 | p1p2) + plot_annotation(tag_levels = "A")
+```
+
+Although the model estimated here already included key DSEM features, dynamic relations and latent variables, it only included one modelled variable and therefore is likely to be of limited practical utility.
+Next, we will build a more useful model that includes social media use as a time-varying covariate.
+
+# Example 2: Multivariate two-level AR1 model
+
+For this next example, we expand on the previous univariate autocorrelative model of happiness by including social media use in the past hour as a predictor.
+Our research question with these data is: To what extent does an individual's total social media use in the past hour predict that individual's current happiness when their immediately preceding happiness is adjusted for?
+That is, we are specifically interested in the within-person correspondence between social media use and well-being.
+
+## Model
+
+As above (Equation 1), we model the happiness rating of the $i^{th}$ individual at their $t^{th}$ measurement occasion as normally distributed.
+We model the response distribution's mean on an intercept ($\beta_{0i}$), a regression coefficient of the previous happiness rating ($\beta_{1i}$), and a coefficient for within-person-centered social media use ($\beta_{2i}$).
+Critically, we again center the lagged happiness and social media use predictors within the model by subtracting the estimated person-specific latent means $\beta_{0i}$ and $\beta_{3i}$ from the corresponding observed values.
+Formally, the model is (we omit the sampling statement for the person-varying parameters for brevity)
+
+$$
+\begin{align*}
+H_{it} &\sim N(\eta_{it}, \sigma^2_0), \\
+\eta_{it} &= \beta_{0i} + \beta_{1i}(H_{it-1} - \beta_{0i}) + \beta_{2i}(SM_{it} - \beta_{3i}), \\
+SM_{it} &\sim N(\beta_{3i}, \sigma^2_1), \\
+\beta_{0i} &= \bar{\beta}_0 + u_{0i}, \\
+\beta_{1i} &= \bar{\beta}_1 + u_{1i}, \\
+\beta_{2i} &= \bar{\beta}_2 + u_{2i}, \\
+\beta_{3i} &= \bar{\beta}_3 + u_{3i}.
+\tag{2}
+\end{align*}
+$$
+
+The key difference in Equation 2 above, compared to Equation 1, is the second sampling statement for social media use ($SM$) needed to estimate its person-specific means.
+However, this model is difficult to specify in common multilevel modelling software, because its regression formulas for different outcomes have common parameters ($\beta_{3i}$).
+
+There is a somewhat simple solution to this issue: We can "stack" the multivariate data to a longer univariate data table.
+To do so, we can for example use the `pivot_longer()` function from the tidyr R package [@tidyr2024].
+We show the result of this data stacking in @tbl-data-2).
+Instead of separate columns for happiness and social media use, there are now two indicator columns `Ih` (for happiness) and `Ism` (social media) that are 1 when that corresponding outcome's value is in the `y` column and 0 otherwise.
+
+```{r}
+#| label: tbl-data-2
+#| tbl-cap: Six rows of example data stacked to long format.
+
+dat <- dat |> 
+  pivot_longer(sm:happy, names_to = "outcome", values_to = "y") |> 
+  mutate(
+    Ih = if_else(outcome == "happy", 1, 0) |> as.integer(),
+    Ism = if_else(outcome == "sm", 1, 0) |> as.integer(), 
+    .before = y
+  ) |> 
+  left_join(dat)
+
+dat2 <- dat |> 
+  select(class, person, time, Ih, Ism, sm, happy, hlag, y)
+
+dat2 |> 
+  slice(1:4, .by = person) |> 
+  head() |> 
+  kable()
+```
+
+Consequently, we then reparameterize the model for the stacked data table to treat `y` as the outcome, and allocate parameters appropriately to either speak to happiness or social media use using the two indicator variables `Ih` and `Ism`.
+The level-1 equations of this model is then written as
+
+$$
+\begin{align*}
+Y_{it} \sim\ &N(\eta_{it}, \sigma_{it}^2), \\
+\eta_{it} =\ &I_{Hit}(\beta_{0i} + 
+  \beta_{1i}(H_{it-1} - \beta_{0i}) + 
+  \beta_{2i}(SM_{it} - \beta_{3i})) + \\
+  &I_{SMit}(\beta_{3i}), \\
+\sigma_{it} =\ &\text{exp}(I_{Hit}\kappa_0 + I_{SMit}\kappa_1).
+\tag{3}
+\end{align*}
+$$
+
+We omit the level-2 equations because they are identical to those in Equation 2.
+This model estimates the same parameters as Equation 2, but has been rearranged to a univariate form with the help of the two indicator variables $I_H$ and $I_{SM}$.
+The former is 1 on row $it$ if the outcome on that row is `happy` (see line 2 in @tbl-data-2)), and zero otherwise.
+The latter is 1 on row $it$ if the outcome on that row is `sm` (see line 1 in @tbl-data-2)), and zero otherwise.
+Accordingly, we now model the outcome $Y$ for the $it^{th}$ observation, and those indicator variables assign appropriate coefficients to the regression formula for $Y$'s mean.
+Importantly, the last line of Equation 3 also models separate residual standard deviations for the two outcomes.
+Because standard deviations must be positive, it is common to model these parameters through a log-link function.
+brms specifies this link function automatically.
+
+With the data in Table @tbl-data-2), we can now specify Equation 3 with brms' model syntax as follows:
+
+```{r}
+#| label: model-2-specify
+#| echo: true
+
+bform <- bf(
+  y ~ 
+    Ih * (b0 + b1 * (hlag - b0) + b2 * (sm - b3)) + 
+    Ism * b3,
+  c(b0, b1, b2, b3) ~ 1 + (1 |p| person),
+  nlf(sigma ~ Ih * k0 + Ism * k1),
+  c(k0, k1) ~ 1,
+  nl = TRUE
+)
+```
+
+In the above code snippet, we have specified that `y` is modeled on one of two regression equations, depending on values of the indicator variables `Ih` ($I_H$) and `Ism` ($I_{SM}$).
+We then drew samples from this model's posterior distribution as in Example 1, and summarize the key parameters' posterior distributions in Table @tbl-fit-2).
+The estimated autocorrelation for the average person ($\bar{\beta}_1$) is very similar to Model 1's estimate.
+A key addition in Model 2 is the $\bar{\beta}_2$ parameter, which indicates the change in happiness associated with a one-hour increase in the past hour's social media use for the average person.
+The 95%CI of this parameter ranges from 0.03 to 0.11: Social media use predicts slightly increased happiness within-person.
+
+```{r}
+#| label: model-2-sample
+
+fit2 <- brm(
+  bform,
+  data = dat,
+  control = list(adapt_delta = 0.99),
+  file = "cache/brm-2"
+)
+```
+
+```{r}
+#| label: tbl-fit-2
+#| tbl-cap: Summaries of estimated parameters from Model 2.
+
+sm(fit2) |> 
+  kable()
+```
+
+In addition, we might be interested in the correlation parameters of this model.
+$\rho_{03}$ in Equation 2.1 indicates the extent to which the intercepts, or the (latent) person-means of `happy` and `sm` are correlated.
+We see from @tbl-fit-2) that this correlation is positive on expectation (0.13), but that we have not estimated it very precisely with the 95%CI ranging from -0.24 to 0.5.
+Finally, we visualize the key results of this model in @fig-fit-2).
+
+```{r}
+#| label: fig-fit-2
+#| fig-height: 4
+#| fig-cap: Implied regression lines of happiness on the previous hour's aggregated social media use for the average person (left) and for nine individuals arranged by increasing slope coefficient (right).
+Points indicate data values and are jittered to reduce overplotting.
+
+# Get slopes in increasing order and select 9
+tmp <- coef(fit2)$person[,,"b2_Intercept"] |> 
+  as.data.frame() |> 
+  rownames_to_column("person") |> 
+  arrange(Estimate) |> 
+  filter(((1:63) %% 7)==0) |> 
+  distinct(person)
+
+epred_average <- dat |> 
+  data_grid(
+    sm = seq_range(sm, n = 9),
+    hlag = mean(hlag, na.rm = TRUE),
+    Ih = 1, Ism = 0
+  ) |> 
+  add_epred_draws(fit2, re_formula = NA) |> 
+  mean_qi()
+
+epred_person <- dat |> 
+  filter(person %in% tmp$person) |> 
+  group_by(person) |> 
+  data_grid(
+    sm = seq_range(sm, n = 9),
+    hlag = mean(hlag, na.rm = TRUE),
+    Ih = 1, Ism = 0
+  ) |> 
+  add_epred_draws(fit2) |> 
+  mean_qi()
+
+p1 <- epred_average |> 
+  ggplot(aes(sm, .epred)) +
+  geom_ribbon(
+    aes(ymin = .lower, ymax = .upper),
+    fill = "grey80"
+  ) +
+  geom_line() +
+  geom_point(
+    data = dat |> 
+      filter(Ih == 1) |> 
+      drop_na(happy),
+    aes(y = happy),
+    shape = 1, alpha = 0.5,
+    position = position_jitter(.01, .01)
+  )
+
+p2 <- epred_person |>   
+  ggplot(aes(sm, .epred)) +
+  geom_ribbon(
+    aes(ymin = .lower, ymax = .upper),
+    fill = "grey80"
+  ) +
+  geom_line() +
+  geom_point(
+    data = dat |> 
+      filter(person %in% tmp$person, Ih == 1) |> 
+      drop_na(happy),
+    aes(y = happy),
+    shape = 1, alpha = 0.5,
+    position = position_jitter(.01, .01)
+  ) +
+  facet_wrap("person")
+
+(p1 | p2) & 
+  scale_x_continuous(
+    "Social media use",
+    breaks = c(0, 0.5, 1)
+  ) &
+  scale_y_continuous(
+    "Happy",
+    breaks = c(0, 0.5, 1)
+  ) &
+  coord_cartesian(
+    xlim = 0:1, ylim = 0:1
+  )
+```
+
+Models 1 and 2 have now examined the extent to which happiness responses in this design are predicted by earlier happiness reports and current social media use, and allowed variability in all parameters between individuals.
+However, the models thus far did not allow for another potentially important source of variation (or clustering), the classroom of each student.
+Example 3 turns the focus on that by estimating a three-level model.
+
+# Example 3: Three-level model
+
+The 63 adolescents in this study were all students in one of six classes.
+Our previous two-level model did not acknowledge potential between-class variation, but there are at least two reasons for modeling it.
+One, ignoring the clusters might lead to estimates that are too certain and therefore likely to lead to conclusions that are made with unwarranted confidence.
+Second, differences between clusters might be theoretically interesting by informing analysts about variation between naturally occurring groups in how they respond to the study items.
+
+## Model
+
+We next illustrate how these issues can be easily addressed by extending the the previous two-level model to a three-level model where responses are nested among individual students, and students nested among classrooms.
+We specify an otherwise identical model to Model 2 but add class-deviation terms to each level-2 equation.
+More formally, we subscript the relevant model features with $j$ for classrooms, and add terms $v_{0-3j}$ to represent classroom deviations from the happiness intercept, autocorrelation, coefficient of social media use, and social media use intercept for the average class and student.
+
+$$
+\begin{align*}
+Y_{ijt} \sim\ &N(\eta_{ijt}, \sigma_{ijt}^2), \\
+\eta_{ijt} =\ &I_{Hijt}(\beta_{0ij} + 
+    \beta_{1ij}(H_{ijt-1} - \beta_{0ij}) + 
+    \beta_{2ij}(SM_{ijt} - \beta_{3ij})) + \\
+  &I_{SMijt}(\beta_{3ij}), \\
+\sigma_{ijt} =\ &\text{exp}(I_{Hijt}\kappa_0 + I_{SMijt}\kappa_1), \\
+\beta_{0ji} =\ &\bar{\beta}_0 + u_{0i} + v_{0j}, \\
+\beta_{1ji} =\ &\bar{\beta}_1 + u_{1i} + v_{1j}, \\
+\beta_{2ji} =\ &\bar{\beta}_2 + u_{2i} + v_{2j}, \\
+\beta_{3ji} =\ &\bar{\beta}_3 + u_{3i} + v_{3j}.
+\tag{4}
+\end{align*}
+$$
+
+Notice that the only difference between equations 3 and 4 is that the latter includes class-specific $v$ terms and their corresponding multivariate distribution (which we omit for brevity).
+Similarly, the brms code to specify this model is identical but for the addition of the `(1 |c| class)` term below:
+
+```{r}
+#| label: model-3-specify
+#| echo: true
+
+bform <- bf(
+  y ~ 
+    Ih * (b0 + b1*(hlag - b0) + b2*(sm - b3)) + 
+    Ism * b3,
+  nlf(sigma ~ Ih * k0 + Ism * k1),
+  c(b0, b1, b2, b3) ~ 1 + 
+    (1 |p| person) + (1 |c| class),
+  c(k0, k1) ~ 1,
+  nl = TRUE
+)
+```
+
+Drawing 2000 samples from this model took about five minutes on a modern multicore laptop. We display the key parameter estimates from this model in Table @tbl-fit-3).
+
+```{r}
+#| label: model-3-sample
+
+fit3 <- brm(
+  bform,
+  data = dat,
+  control = list(adapt_delta = 0.99),
+  file = "cache/brm-3"
+)
+```
+
+```{r}
+#| label: tbl-fit-3
+#| tbl-cap: Summaries of estimated parameters from Model 3.
+
+sm(fit3) |> 
+  mutate(
+    variable = str_replace_all(variable, "_class", "_c"),
+    variable = str_replace_all(variable, "_person", "_p")
+  ) |> 
+  kable()
+```
+
+# Discussion
+
+## Advantages
+
+Using brms for latent means models enables the use of a wide variety of methods, including nonlinear model terms, smooth terms (generalized additive mixed models), mixture models, and more.
+
+## Limitations
+
+Here, we outlined the use of the general-purpose regression modelling R package brms for estimating latent means models.
+This approach has two main limitations.
+First, by requiring us to stack the data we cannot model different variables using different distributions.
+In our examples, we assumed that all outcomes were normally distributed.
+This practical limitation can be problematic in cases, for example, when one of the outcomes is binary and the other one is a continuous variable.
+Second, although a wide variety of e.g.
+cross-lagged location-scale within-between latent means models can be fitted with brms, the commercial alternative Mplus is currently more flexible.
+For example, users of Mplus can specify residual DSEMs which cannot be fitted with brms.
+
+Moreover, data stacking increases the practical size of the dataset and thus can be slower to estimate than a true multivariate model, such as ones that can be fitted with Mplus.
+
+Missing data.
+
+# Disclosures
+
+## Data and code availability
+
+The online analysis supplement is readable at {{<meta site-url>}}. Our materials are available at GitHub ({{<meta repo-url>}}) and the OSF ({{<meta osf-url>}}).
+
+## Author contributions
+<!-- https://casrai.org/credit/ -->
+
+Conceptualization: MV\
+Methodology: MV\
+Software: MV, JJ\
+Validation: MV\
+Formal Analysis: MV, MGV, JJ\
+Data Curation: MV\
+Writing---Original Draft: MV\
+Writing---Review & Editing: MV\
+Visualization: MV\
+Project Administration: MV
+
+## Competing interests
+
+The author(s) declare no competing interests.
+
+<!-- Format references better in non-html formats -->
+::: {.content-hidden when-format="html"}
+# References
+
+:::{.refs}
+:::
+:::
+
+<!-- Material after this only appears in html output -->
+::: {.content-visible when-format="html"}
+
+# Appendix 1: Full Model 1 summary
+
+Here, we display the full summary that brms prints for Model 1.
+
+```{r}
+#| label: fit1-full-summary
+
+summary(fit1, prior = TRUE)
+```
+
+From top to bottom, the printout first tells us about the probability distribution we assumed for the outcome (Family), and the link functions from the regression equations to the response distribution's location (mu), scale (sigma), and potentially other parameters.
+Then, the formula is printed, along with a summary of the data and Hamiltonian Monte Carlo draws [@standevelopmentteam2023].
+
+Following this, the printout shows the model parameters' prior distributions (Priors).
+Because we did not specify any, brms automatically applied weakly informative prior distributions (appropriately scaled Student's *t*-distributions) on the standard deviation parameters, and a uniform LKJ prior on the correlation parameter [@standevelopmentteam2023].
+These weakly informative priors serve only to facilitate the estimation algorithm and do not bias the resulting posterior distributions.
+By default, regression coefficients are not assigned any prior distribution, and are therefore not printed.
+
+The following section describes the dispersion and correlation parameters of the group-level (e.g.
+person-level) parameters.
+These are labelled "Multilevel Hyperparameters".
+For each, the posterior mean (Estimate) standard deviation (Est.Error), and lower (l-95%) and upper (u-95%) limits of the quantile-based credibility interval are listed.
+Additionally, Rhat evaluates the HMC algorithm's convergence and should be near 1.00, and the two ESS numbers indicate the effective numbers of independent samples from the posterior distribution and should be near the number of desired samples.
+In this example, these two values are wrapped to separate lines.
+
+The Regression Coefficients section summarizes the posterior distributions of the coefficients specified on the regression equation's RHS with the same numeric summaries as above.
+Further Distributional Parameters include, in the gaussian case, the residual standard deviation's posterior summary.
+
+# Appendix 2: Parameter recovery simulation
+
+Here we simulate from, specify, and validate latent means models with brms.
+
+## Univariate model
+
+The data simulation and Stan code here is based on [Joran's blog post](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/).
+
+The model is for each individual i and timepoint t,
+
+y_it ~ normal(eta_it, sigma)
+eta_it = (alpha_hat + alpha_i) + (beta_hat + beta_i) * (y_i[t-1])
+[alpha_i, beta_i] ~ mvn([0, 0], Sigma)
+
+That is we model y on an intercept and an autoregressive coefficient with varying intercepts and slopes over people.
+
+### Data
+
+We first set up the data simulation.
+
+The function returns a list that we can pass to Stan. It also contains the data generation parameters.
+
+```{r}
+#| label: sim-example
+
+source("R/simulate-univariate.R")
+
+set.seed(31121)
+dat <- simulate_univariate()
+glimpse(simulate_univariate())
+```
+
+### Stan
+
+Let us then compile and sample from the Stan model (`model-ar1.stan`) with cmdstanr
+
+```{r}
+#| label: stan-compile
+#| output: false
+
+path <- "cache/cmdstanr-example.rds"
+if (!file.exists(path)) {
+  model <- cmdstan_model("stan/model-ar1.stan", compile = FALSE)
+  model$format(
+    canonicalize = list("deprecations"),
+    overwrite_file = TRUE,
+    backup = FALSE
+  )
+  model$compile(dir = "cache/")
+  post_cmdstanr <- model$sample(
+    data = dat, 
+    adapt_delta = 0.99
+  ) |> 
+    as_draws_df()
+  saveRDS(post_cmdstanr, path)
+} else {
+  post_cmdstanr <- readRDS(path)
+}
+```
+
+Posterior summary:
+
+```{r}
+#| label: stan-ex
+
+post_cmdstanr <- post_cmdstanr |> 
+  select(
+    alpha_hat_raw, alpha_scale_raw, 
+    beta_hat, beta_scale, 
+    sigma_raw
+  ) |> 
+  rename_with(~str_remove(., "_raw"))
+
+summarize_draws(post_cmdstanr) |> 
+  kable()
+```
+
+### brms
+
+brms takes a data frame, so we convert the list here
+
+```{r}
+#| label: sim-data-show
+
+dat <- as_tibble(dat)
+dat <- dat |> 
+  mutate(yt1 = lag(y), .by = person)
+head(dat) |> 
+  kable()
+```
+
+#### Model specification
+
+We can specify this model using brms' nonlinear formulas. Note that we created the `yt1` variable (i.e. y[t-1]) above.
+
+```{r}
+#| label: sim-brms-example
+#| output: false
+
+bform <- bf(
+  y ~ 0 + alpha + beta * (yt1 - alpha),
+  alpha ~ 1 + (1 | person),
+  beta ~ 1 + (1 | person),
+  nl = TRUE
+)
+fit_brms <- brm(
+  bform,
+  data = dat,
+  file = "cache/brm-example.rds"
+)
+```
+
+We see that the estimated parameters are very close to coefficients from the Stan model
+
+```{r}
+#| label: sim-brms-table
+
+post_brms <- as_draws_df(fit_brms) |> 
+  select(
+    alpha_hat = b_alpha_Intercept,
+    alpha_scale = sd_person__alpha_Intercept,
+    beta_hat = b_beta_Intercept,
+    beta_scale = sd_person__beta_Intercept,
+    sigma
+  )
+
+bind_rows(
+  "Stan" = summarize_draws(post_cmdstanr),
+  "brms" = summarize_draws(post_brms),
+  .id = "Model"
+) |> 
+  arrange(variable, Model) |> 
+  kable()
+```
+
+### Simulation
+
+Above, we estimated the Stan and brms models once. Here, we evaluate whether they recover the specified parameters with a small simulation.
+
+```{r}
+#| label: sim-run
+
+path <- "cache/sim-univariate.rds"
+if (!file.exists(path)) {
+  # Ensure that models are compiled
+  model_stan <- cmdstan_model("stan/model-ar1.stan", compile = TRUE, dir = "cache")
+  fit_brms <- brm(
+    bform,
+    data = dat,
+    chains = 0,
+    threads = 1
+  )
+
+  # Define simulation outputs
+  K <- 1
+  out <- vector("list", length = K)
+
+  # Parallelization scheme:
+  # Each of two workers will launch a session that runs 4 chains in parallel
+  # resulting in total 8 cores used
+  library(future)
+  plan(multisession, workers = 2)
+
+  for (k in 1:K) {
+    this_sample <- simulate_univariate()
+
+    this_stan <- future({
+      model_stan$sample(
+        data = this_sample, 
+        adapt_delta = 0.99, 
+        parallel_chains = 4,
+        iter_sampling = 2000,
+        refresh = 0, 
+        show_messages = FALSE
+      )
+    }, seed = TRUE)
+
+    this_brms <- future({
+      update(
+        fit_brms, 
+        newdata = as_tibble(this_sample) |> 
+          mutate(yt1 = lag(y), .by = person),
+        control = list(adapt_delta = 0.99),
+        cores = 4,
+        iter = 2000,
+        recompile = FALSE,
+        refresh = 0,
+        silent = 2
+      )
+    }, seed = TRUE)
+
+    this_stan <- value(this_stan) 
+    this_stan <- this_stan |> 
+      as_draws_df() |> 
+      select(
+        alpha_hat_raw, alpha_scale_raw, 
+        beta_hat, beta_scale, 
+        sigma_raw
+      ) |> 
+      rename_with(~str_remove(., "_raw")) |> 
+      summarize_draws()
+
+    this_brms <- value(this_brms) 
+    this_brms <- this_brms |> 
+      as_draws_df() |> 
+      select(
+        alpha_hat = b_alpha_Intercept,
+        alpha_scale = sd_person__alpha_Intercept,
+        beta_hat = b_beta_Intercept,
+        beta_scale = sd_person__beta_Intercept,
+        sigma
+      ) |> 
+      summarize_draws()
+
+    out[[k]] <- bind_rows(
+      "Stan" = this_stan,
+      "brms" = this_brms,
+      .id = "model"
+    )
+  }
+
+  out <- bind_rows(out, .id = "k")
+  saveRDS(out, path)
+} else {out <- readRDS(path)}
+```
+
+#### Results
+
+```{r}
+#| label: tbl-simulation
+
+real_values <- tribble(
+  ~variable, ~true,
+  "sigma", 1, 
+  "alpha_hat", 4, 
+  "alpha_scale", 1, 
+  "beta_hat", 0.4, 
+  "beta_scale", 0.1
+)
+out |> 
+  summarise(
+    mean = str_glue("{number(mean(mean), .01)} ({number(sd(mean), .01)})"),
+    .by = c(model, variable)
+  ) |> 
+  pivot_wider(
+    names_from = model, values_from = mean
+  ) |> 
+  left_join(
+    real_values
+  ) |> 
+  kable()
+```
+
+## Conclusion
+
+This small simulation shows that brms perfectly recovers parameters of a simple univariate autoregressive latent variable model.
+The mean estimates over 30 simulation iterations were identical for brms and Stan to the second significant digit, and within MCSE of the true values.
+
+# Appendix 3: Further examples
+
+Here, we replicate models 4-8 from "A Primer on Two-Level Dynamic Structural Equation Models for Intensive Longitudinal Data in Mplus" [@mcneish2020].
+This appendix serves both to illustrate the syntax in more detail and compare the resulting estimates from those obtained from Mplus.
+
+```{r}
+#| label: data-mcneish-hamaker
+
+# "A primer on two-level dynamic structural equation models for intensive longitudinal data in Mplus" (McNeish & Hamaker, 2020). Those data are available on the OSF (<https://osf.io/wuprx/>).
+
+path <- "data/mh.zip"
+url <- "https://files.osf.io/v1/resources/wuprx/providers/osfstorage/5bfc839601593f0016774697/?zip="
+if (!file.exists(path)) {
+  dir.create(dirname(path), showWarnings = FALSE)
+  download.file(
+    url,
+    destfile = path
+  )
+  unzip(path, exdir = "data/mh/")
+}
+
+path <- "data/mcneish-hamaker.rds"
+if (!file.exists(path)) {
+d <- read_csv(
+  "data/mh/Data/Two-Level Data.csv", 
+  col_names = c("urge", "dep", "js", "hs", "person", "time")
+) %>% 
+  select(
+    person, time, js, hs, urge, dep
+  ) |> 
+  mutate(
+    person = factor(person),
+    time = as.integer(time-1)
+  )
+write_rds(d, path)
+}
+```
+
+```{r}
+#| label: mplus-parameters
+# Read and wrangle MH's Mplus model parameters for comparison
+
+path <- "data/mplus-estimates.rds"
+if (!file.exists(path)) {
+
+  library(MplusAutomation)
+
+  mplus <- readModels("data/mh/Output Files")
+  mplus <- map(mplus, ~pluck(., "parameters") |> pluck("unstandardized"))
+  names(mplus) <- str_extract(names(mplus), "equation\\.[0-9]")
+
+  mplus$equation.4 <- mplus$equation.4[c(3, 5, 6, 4, 2, 1, 7, 9, 10, 8), ]
+  mplus$equation.5 <- mplus$equation.5[c(11, 7, 8, 12, 3, 4, 13, 5, 6, 10, 9, 2, 1, 15, 16, 17, 14), ]
+  mplus$equation.6 # This is missing parameters
+  mplus$equation.7
+
+  mplus <- map(mplus, ~rename(., q50 = est, q2.5 = lower_2.5ci, q97.5 = upper_2.5ci))
+
+  mplus <- map(
+    mplus, 
+    ~mutate(
+      .,
+      across(where(is.numeric), ~number(., .01)),
+      mplus = str_glue("{q50} [{q2.5}, {q97.5}]")
+    )
+  )
+  write_rds(mplus, path)
+}
+
+mplus <- read_rds(path)
+```
+
+## Data
+
+These example data are 100 individuals' reports of urge to smoke (`urge`) and depression (`dep`) at 50 time points [@mcneish2020].
+We also have their (non-time-varying) job stress (`js`) and home stress (`hs`).
+We also create lagged versions of urge to smoke (`u_lag`) and depression (`u_dep`) for use in the analyses.
+We show these data in @tbl-a1-1).
+
+```{r}
+#| label: tbl-a1-1
+#| tbl-cap: First six rows of the example dataset on urge to smoke and depression.
+
+dat <- read_rds("data/mcneish-hamaker.rds") |> 
+  mutate(
+    u_lag = lag(urge), 
+    d_lag = lag(dep), 
+    .by = person
+  )
+dat |> 
+  head() |> 
+  kable()
+```
+
+## Univariate latent means model
+
+We begin with a univariate model of the urge to smoke.
+This model examines the degree of autocorrelation in the urge to smoke and how it varies between people.
+For individual *i* in 1...I=100 and time point *t* in 1...T=50, we model `urge` (U) as normally distributed.
+We model the mean on person-specific intercepts $\alpha_i$ and slopes of that person's within-person centered `urge` at a previous time point $\phi_i$.
+Person-specific parameters are modelled as multivariate normal on the population-level parameters (with hats).
+We do not model correlations between the intercepts and slopes.
+
+First, let us pay some attention to the issue of within-person centering.
+Instead of decomposing urge to smoke into its within- and between-person components before fitting the model, we use "latent mean centering".
+What this means is that we estimate the person means along with other model parameters, and subtract those means from the observed values in one step.
+We refer to the latent person mean variables with an upperscript c (e.g.
+$U^c_{it}$).
+
+$$
+\begin{align*}
+U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1}, \sigma^2), \\
+\begin{bmatrix}
+  \alpha_i \\ \phi_i
+\end{bmatrix} &\sim MVN\left(
+  \begin{bmatrix}
+    \bar{\alpha} \\ \bar{\phi}
+  \end{bmatrix},
+  \begin{pmatrix}
+  \tau_\alpha \ &0 \\ 0 \ &\tau_\phi
+  \end{pmatrix}
+\right).
+\tag{a1}
+\end{align*}
+$$
+
+This model is sometimes written with separate "level 1" and "level 2" equations:
+
+$$
+\begin{align*}
+U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1}, \sigma^2), \\
+\alpha_i &= \bar{\beta}_{00} + u_{0i}, \\
+\phi_i &= \bar{\beta}_{10} + u_{1i}, \\
+\beta_i &= \bar{\beta}_{20} + u_{2i}, \\
+\mathbf{u} &\sim MVN(\dots)
+\tag{a2}
+\end{align*}
+$$
+
+We will use the R package brms [@bürkner2017] to estimate this model.
+The following code chunk shows how to specify this model inside brms' `bf()` ("brmsformula") function.
+In the first line, we specify a regression equation for `urge`.
+Everything on the right-hand side of this formula (to the right of the tilde) is treated as a regression coefficient to be estimated from data *unless* it is the exact name of a variable in the data.
+Thus we will be estimating an `alpha` (intercept) and a `phi` (the autoregressive coefficient).
+
+The only "special" part in this syntax is `(u_lag - alpha)` that does the "latent mean centering" of the lagged urge variable.
+That is, it just subtracts `alpha` from each lagged value, and then uses the resulting demeaned lagged variable to predict the current urge.
+The first line can be considered the "level 1" equation or rather the *nonlinear* part of the model.
+
+```{r}
+model <- bf(
+  urge ~ alpha + phi * (u_lag - alpha),
+  alpha ~ 1 + (1 | person),
+  phi ~ 1 + (1 | person),
+  nl = TRUE
+)
+```
+
+The second and third lines then specify the "level 2" equation, or rather linear equations to predict the parameters in the above (potentially) nonlinear model.
+Both regression parameters are modelled on a population level intercept (the gamma in Equation A1) and person-specific deviations from it.
+
+The fourth line specifying `nl = TRUE` is critical, because it allows us to specifically name parameters inside `bf()`, and thereby to e.g.
+construct the latent mean centered variable on the first row.
+
+Note that the linear models of `alpha` and `phi` are identical. We can therefore use the following shorthand to assign them the same linear model:
+
+```{r}
+#| label: ex-1-bf
+#| echo: true
+
+model <- bf(
+  urge ~ alpha + phi * (u_lag - alpha),
+  alpha + phi ~ 1 + (1 | person),
+  nl = TRUE
+)
+```
+
+We will use that shorthand going forward.
+We could also indicate the distribution that we assume for the data.
+But in this work we model everything as gaussian, which is the software default and thus doesn't need to be separately indicated.
+
+We then sample from the model. Everything from here on is standard operating procedure.
+
+```{r}
+#| label: ex-1-fit
+
+fit <- brm(
+  model,
+  data = dat,
+  file = "cache/brm-example-univariate"
+)
+```
+
+The object `fit` now contains the estimated model (the data, posterior samples, and lots of brms-specific information). We can call `summary(fit)` to see a default summary of the model.
+
+```{r}
+#| label: ex-1-summary
+
+summary(fit)
+```
+
+The first few rows above print information about the model (the formulas, data, and number of posterior samples).
+Then, "Group-Level Effects" are standard deviations (and correlations, if estimated) of the parameters that we allowed to vary across individuals (as indicated by `~person`).
+For each of those parameters, one row indicates its posterior summary statistics; "Estimate" is the posterior mean, "Est.Error" is the posterior standard deviation, "l-" and "u-95% CI" are the lower and upper bounds of the 95% credibility interval (so the 2.5 and 97.5 percentiles of the posterior samples).
+Then, Rhat is the convergence metric which should be smaller than 1.05 (optimally 1.00) to indicate that the estimation algorithm has converged.
+"Bulk_" and "Tail_ESS" indicate the effective sample sizes of the posterior draws, and should be pretty large.
+
+The "Population-Level Effects" indicate the same information but for the means of the person-specific parameters' distributions; or the "fixed effects".
+For the average person, there is a positive autocorrelation in these data.
+Finally, the "Family Specific Parameters" indicate parameters that are specific to the outcome distribution.
+We used the default gaussian distribution, and thus get an estimated residual standard deviation.
+
+We show the results in @tbl-a1-2).
+
+```{r}
+#| label: tbl-a1-2
+#| tbl-cap: Summaries of main parameters from the example univariate model.
+
+fit |> 
+  sm() |> 
+  kable()
+```
+
+## Multilevel AR(1) Model
+
+We then turn to models with more than one variable of interest. In what follows we replicate "Two-Level Dynamic Structural Equation Models" (equations 4 to 7) from @mcneish2020.
+
+We start with the two-level AR(1) model in @mcneish2020 (equations 4a-c).
+This model is similar to our example model from above, but we also now include a regression coefficient for within-person centered depression, where this centering is on the latent mean of depression.
+The model is
+
+$$
+\begin{align*}
+U_{it} &\sim N(\alpha_i + \phi_i U^c_{it-1} + \beta_i D^c_{it}, \sigma^2), \\
+U^{c}_{it} &= U^{\text{raw}}_{it} - \alpha^U_i, \\
+D^{c}_{it} &= D^{\text{raw}}_{it} - \alpha^D_i, \\
+\begin{bmatrix}
+  \alpha^U_i \\ \alpha^D_i \\ \phi_i \\ \beta_i
+\end{bmatrix} &\sim MVN\left(
+  \begin{bmatrix}
+    \bar{\alpha^U} \\ \bar{\alpha^D} \\ \bar{\phi} \\ \bar{\beta_i}
+  \end{bmatrix},
+  \begin{pmatrix}
+    \tau_\alpha^U \ &0 \ &0 &0 \\ 
+    0 \ &\tau_\alpha^D \ &0 \ &0 \\ 
+    0 \ &0 \ &\tau_\beta \ &0 \\
+    0 \ &0 \ &0 \ &\tau_\phi
+  \end{pmatrix}
+\right)
+\end{align*}
+$$
+
+We then see from @eq-2 that we need to refer to different outcomes' parameters across model formulas.
+That is, when predicting the urge to smoke, we need a way to refer to the (latent) mean of depression so that we can appropriately center the depression predictor.
+Currently brms does not support sharing parameters across formulas for different outcomes, but we can overcome this limitation with a small data wrangling trick
+
+That is, we "stack" our data into the long format with respect to the two different outcomes, urge to smoke and depression.
+Then, on each row we have all variables from that measurement occasion, in addition to new ones that indicate the value of the outcome, and which outcome it refers to @tbl-a1-2).
+
+```{r}
+#| label: tbl-a1-3
+#| tbl-cap: Rearranged data for multivariate models.
+
+dat <- dat |> 
+  select(-js, -hs) |> 
+  pivot_longer(c(urge, dep), names_to = "outcome") |> 
+  mutate(
+    i_urge = if_else(outcome == "urge", 1, 0),
+    i_dep = if_else(outcome == "dep", 1, 0)
+  ) |> 
+  # Include predictors from each row
+  left_join(dat)
+
+dat |> 
+  head() |> 
+  kable()
+```
+
+Now we can model `value` and use `i_urge` and `i_dep` to control whether a parameter should apply to the outcome on a particular row.
+So here we want alpha, phi, and beta to apply to urge, but depb to depression.
+Then we can use depb to create the predictor for urge on the fly.
+
+```{r}
+#| label: ex-2-bf
+#| echo: true
+
+bform <- bf(
+  value ~ 
+    i_urge * (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2)) + 
+    i_dep * alpha2,
+  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
+  alpha1 + phi + beta + alpha2 ~ 1 + (1 | person),
+  sigma1 + sigma2 ~ 1,
+  nl = TRUE
+)
+```
+
+Sample from model
+
+```{r}
+#| label: ex-2-fit
+
+fit <- brm(
+  bform,
+  data = dat,
+  control = list(adapt_delta = 0.95),
+  file = "cache/brm-example-4"
+)
+```
+
+Compare model summary to @mcneish2020.
+We can see the estimates match to within differences in priors and MCSE.
+Note in the code I transform standard deviations by first exponentiating draws of residual standard deviations, and then square to put them on the variance scale as in @mcneish2020.
+
+```{r}
+#| label: mplus-parameters2
+#| tbl-cap: Mplus parameter estimates.
+
+source("R/mplus-parameters.R")
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2),
+    across(starts_with("b_sigma"), ~exp(.)^2)
+  ) |> 
+  summarise_draws(
+    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
+      number(.01) |> 
+      str_glue_data("{q50} [{q2.5}, {q97.5}]")
+  ) |> 
+  mutate(
+    variable = str_replace(variable, "sd_person__", "var_"),
+    mplus = mplus$equation.4$mplus
+  ) |> 
+  kable()
+```
+
+## Time-Invariant Covariates
+
+This is as above but includes time-invariant covariates js and hs, and the between-person effect of depression on urge.
+I label the latter as "gamma03" as in Equation 5 in @mcneish2020.
+Note also that I must include it in the regression equation for the outcome, and not in the "level 2" equation because the latter must be "linear formulas" in brms.
+There's probably other ways of doing it too.
+
+```{r}
+#| label: ex-3-bf
+#| echo: true
+
+bform <- bf(
+  value ~ 
+    i_urge * 
+    (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2) + gamma03 * alpha2) + 
+    i_dep * alpha2,
+  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
+  alpha1 + phi + beta ~ 1 + js + hs + (1 | person),
+  alpha2 ~ 1 + (1 | person),
+  gamma03 + sigma1 + sigma2 ~ 1,
+  nl = TRUE
+)
+```
+
+With flat priors the gamma03 parameter is wonky so I set a normal(0, 3) on it.
+
+```{r}
+#| label: ex-3-prior
+
+p <- get_prior(bform, dat)
+p <- p |> 
+  mutate(
+    prior = case_when(
+      class == "b" & nlpar != "gamma03" ~ "normal(0, 1)",
+      class == "b" & nlpar == "gamma03" ~ "normal(0, 3)",
+      class == "sd" ~ "student_t(7, 0, 1)",
+      TRUE ~ prior
+    )
+  )
+```
+
+Sample from model
+
+```{r}
+#| label: ex-3-fit
+
+fit <- brm(
+  bform,
+  data = dat,
+  prior = p,
+  control = list(adapt_delta = 0.99),
+  file = "cache/brm-example-5",
+)
+```
+
+Model summary.
+
+```{r}
+#| label: ex-3-summary
+
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2),
+    across(starts_with("b_sigma"), ~exp(.)^2)
+  ) |> 
+  summarise_draws(
+    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
+      number(.01) |> 
+      str_glue_data("{q50} [{q2.5}, {q97.5}]")
+  ) |> 
+  mutate(
+    variable = str_replace(variable, "sd_person__", "var_"),
+    mplus = mplus$equation.5$mplus
+  ) |> 
+  kable()
+```
+
+## Multilevel Location-Scale model
+
+Equations 5 and 6 in @mcneish2020.
+
+```{r}
+#| label: ex-5-bf
+#| echo: true
+
+bform <- bf(
+  value ~ 
+    i_urge * 
+    (alpha1 + phi * (u_lag - alpha1) + beta * (dep - alpha2) + gamma03 * alpha2) + 
+    i_dep * alpha2,
+  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
+  alpha1 + phi + beta ~ 1 + js + hs + (1 | person),
+  alpha2 + sigma1 + sigma2 ~ 1 + (1 | person),
+  gamma03 ~ 1,
+  nl = TRUE
+)
+```
+
+Sample from model
+
+```{r}
+#| label: ex-5-fit
+
+fit <- brm(
+  bform,
+  data = dat,
+  prior = p,
+  control = list(adapt_delta = 0.99),
+  file = "cache/brm-example-6"
+)
+```
+
+Model summary. These match.
+
+```{r}
+#| label: tbl-ex5
+
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2),
+    across(starts_with("b_sigma"), ~exp(.)^2)
+  ) |> 
+  summarise_draws(
+    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
+      number(.01) |> 
+      str_glue_data("{q50} [{q2.5}, {q97.5}]")
+  ) |> 
+  mutate(
+    variable = str_replace(variable, "sd_person__", "var_")
+  ) |> 
+  kable()
+```
+
+## Multilevel VAR(1) Model
+
+Equations 7 and 8 in @mcneish2020.
+
+```{r}
+#| label: ex-6-bf
+#| echo: true
+
+bform <- bf(
+  value ~ 
+    i_urge * (alpha1 + phi1 * (u_lag - alpha1) + phi4 * (d_lag - alpha2)) + 
+    i_dep * (alpha2 + phi2 * (d_lag - alpha2) + phi3 * (u_lag - alpha1)),
+  nlf(sigma ~ i_urge * sigma1 + i_dep * sigma2),
+  alpha1 + alpha2 + phi1 + phi2 + phi3 + phi4 + sigma1 + sigma2 ~ 1 + (1 | person),
+  nl = TRUE
+)
+```
+
+Sample
+
+```{r}
+#| label: ex-6-fit
+
+fit <- brm(
+  bform,
+  data = dat,
+  control = list(adapt_delta = 0.99),
+  file = "cache/brm-example-7"
+)
+```
+
+Summarise. I wasn't sure how to transform the variance components here to match the variances on the link function scale so left them as is.
+
+```{r}
+#| label: tbl-ex6
+
+as_draws_df(fit, variable = c("b_", "sd_"), regex = TRUE) %>% 
+  mutate(
+    across(starts_with("sd_"), ~.^2)
+  ) |> 
+  summarise_draws(
+    brms = ~quantile2(., probs = c(.5, .025, .975)) |> 
+      number(.01) |> 
+      str_glue_data("{q50} [{q2.5}, {q97.5}]")
+  ) |> 
+  mutate(
+    variable = str_replace(variable, "sd_person__", "var_"),
+    mplus = mplus$equation.7$mplus
+  ) |> 
+  kable()
+```
+
+## Conclusion
+
+We estimated models 4-7 in @mcneish2020 with brms and got very similar if not identical results.
+
+:::

### 2025-01-28 index.qmd
Update & lint

@@
     shape = 1, alpha = 0.5, size = 0.5,
-    position = position_jitter(.01, .01)
+    position = position_jitter(.05, .05)
   )
@@
 fit1_fixef_ellipse <- ellipse(
-  cor(as.data.frame(fit1)[,1:2])[1,2],
-  scale = fit1_fixef_ellipse$sd, centre = fit1_fixef_ellipse$mean
+  x = cor(as.data.frame(fit1)[,1:2])[1,2],
+  scale = fit1_fixef_ellipse$sd,
+  centre = fit1_fixef_ellipse$mean
 ) |> as_tibble()
@@
   ) +
-  # geom_segment(
-  #   aes(xend = Estimate.b0, yend = Estimate.b1),
-  #   linewidth = 0.1, alpha = 0.5
-  # ) +
   geom_point(
@@
 
-dat <- dat |> 
+dat2 <- dat |>
   pivot_longer(sm:happy, names_to = "outcome", values_to = "y") |>
@@
 
-dat2 <- dat |> 
+dat2 <- dat2 |>
   select(class, person, time, Ih, Ism, sm, happy, hlag, y)
@@
 
-epred_average <- dat |> 
+epred_average <- dat2 |>
   data_grid(
@@
 
-epred_person <- dat |> 
+epred_person <- dat2 |>
   filter(person %in% tmp$person) |>
@@
   geom_point(
-    data = dat |> 
+    data = dat2 |>
       filter(Ih == 1) |>
@@
   geom_point(
-    data = dat |> 
+    data = dat2 |>
       filter(person %in% tmp$person, Ih == 1) |>
@@
   bform,
-  data = dat,
+  data = dat2,
   control = list(adapt_delta = 0.99),
